\documentclass[12pt]{article}
\usepackage{fontspec}
\usepackage{polyglossia}
% Use a fallback font (e.g., FreeSerif) available in TeX Live
\setmainfont{FreeSerif}
\newfontfamily\arabicfont{Amiri}[Script=Arabic, Scale=1.0]
\newfontfamily\hebrewfont{SBL Hebrew}[Script=Hebrew, Scale=1.0]
\setdefaultlanguage{english}
\setotherlanguages{arabic,hebrew}
\usepackage{geometry}
\geometry{a4paper, margin=1in}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{tikz-cd}
\usepackage{tabularx}


% Define theorem environments
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem*{theorem*}{Theorem}

% Define custom commands for linguistic terms
\newcommand{\fatha}{\textarabic{َ}}
\newcommand{\kasra}{\textarabic{ِ}}
\newcommand{\damma}{\textarabic{ُ}}
\newcommand{\Verb}{\mathsf{Verb}} 
\newcommand{\Noun}{\mathsf{Noun}} 
\newcommand{\Partic}{\mathsf{Partic}} 

% Increase TeX memory parameters to avoid stack size error
\makeatletter
\@namedef{poolsize}{1000000}
\@namedef{stringpool}{100000}
\makeatother

\title{Chokepoint Capitalism in Knowledge Infrastructures: An RSVP-Theoretic Analysis}
\author{Flyxion}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This essay examines chokepoint capitalism as a mechanism for restricting knowledge diversity across digital, physical, and cultural infrastructures, using an RSVP-theoretic framework enriched with category and sheaf theory. Through cases spanning mobile operating systems, festival economics, visa policies, AI research platforms, and the historical evolution of alphabetic systems, I argue that chokepoints misprice epistemic diversity by enforcing premature evaluation, reducing negentropic potential. A functional paradigm of deferred automation, modeled as a monadic lazy-evaluation regime, is proposed as a counter-strategy, with the Arabic script’s morphological generators serving as a computational exemplar. Mathematical formalisms are offered not only to populate abstract registers but also to critique the cultural fetish for mathematical authority, with natural-language translations provided for accessibility.
\end{abstract}

\section{Introduction}
Chokepoint capitalism, as articulated by \citet{giblin2022chokepoint}, describes economic structures where gatekeepers control access to markets, extracting rents while suppressing alternatives. This essay extends the concept to knowledge infrastructures, arguing that chokepoints—whether in digital platforms, physical events, immigration policies, or historical script evolution—systematically undervalue viewpoint diversity, a critical driver of epistemic resilience. Using the RSVP framework, where scalar capacity ($\Phi$), vector flows ($v$), and entropy ($S$) model system dynamics, I demonstrate how such restrictions function as right-adjoint filters that collapse diversity colimits. A functional paradigm of deferred evaluation, inspired by lazy computation, is proposed to preserve negentropic potential, with the Arabic script’s morphology as a computational case study.

\section{Chokepoint Capitalism as Knowledge Restriction}

\subsection{Digital Platforms}
Mobile operating systems (Android, iOS) impose chokepoints by restricting system-level customization (e.g., fonts, terminals). 
For instance, Samsung’s theme economy licenses font design as a scarce privilege, akin to a lottery, while Android’s ``Always use this app'' 
prompt enforces premature defaults, creating hysteresis that discourages exploration of alternatives \citep{doctorow2020lockin}. 

These design choices function as economic and epistemic tollbooths: they diminish the range of feasible pathways by privileging 
a narrow set of defaults and excluding unapproved alternatives. In RSVP terms, the scalar capacity ($\Phi$) of possible user 
configurations is artificially constrained; vector flows ($v$) of exploration are collapsed into channels sanctioned by 
platform gatekeepers; and entropy ($S$), which would normally measure the richness of heterogeneity across the user base, 
is suppressed into homogenized usage patterns. 

This dynamic mirrors the rent-seeking structure of chokepoint capitalism more broadly: the ecosystem is portrayed as open 
and diverse, yet the underlying infrastructure routes all innovation through bottleneck interfaces. Once defaults are entrenched, 
users face high switching costs---a form of enforced eager evaluation, where possibilities are collapsed before exploration can occur. 
The system extracts value not by producing novelty but by restricting its circulation, reducing epistemic diversity while monetizing access.


\subsection{Physical Analogues}
At festivals, organizers charge vendors exorbitant fees for exclusivity, policing independent sellers to maintain artificial scarcity. 
This mirrors digital chokepoints: access is monetized, and diversity (of vendors or ideas) is suppressed \citep{giblin2022chokepoint}. 
Such practices transform what could be open commons of exchange into curated monocultures, where the cost of entry filters out 
all but the most capitalized participants. 

The festival economy illustrates a broader structural logic: rather than maximizing serendipity and heterogeneity of encounters, 
organizers constrain flows through tollbooths. In RSVP-theoretic terms, scalar capacity ($\Phi$) is artificially reduced---the 
potential number of feasible vendors is far greater than those admitted---while vector flows ($v$) are redirected through 
exclusive contracts and patrols. The entropy term ($S$), which would normally reflect distributed novelty and diversity, 
is collapsed into a handful of predictable offerings. 

This is not unique to cultural festivals. Analogous patterns can be found in farmers' markets that restrict “non-certified” 
growers, in concert venues where only corporate sponsors are allowed to vend, and in academic conferences where 
publication or registration fees function as economic filters. In each case, chokepoints are rationalized as 
quality-control mechanisms but function primarily as revenue streams. 

The lesson for knowledge infrastructures is that physical chokepoints habituate participants to constrained choice and 
normalize the extraction of rents for access. Just as mobile platforms shape user expectations of defaults and exclusivity, 
so too do festivals and markets encode scarcity as the condition of participation, narrowing the epistemic and cultural 
diversity that might otherwise emerge.


\subsection{State-Level Chokepoints}
The U.S. H-1B visa fee of \$100,000, implemented on September 20, 2025, restricts labor mobility, mispricing the epistemic value of diverse perspectives from high-population countries like India and China \citep{cbc2025trumpvisa}. The policy, signed by President Donald Trump, applies only to new applicants and has sparked confusion, with the White House clarifying it does not affect current holders or renewals. Over 70 percent of H-1B holders are from India, and the policy has been denounced by India's Ministry of External Affairs for potential humanitarian disruptions to families. Critics argue it undercuts American workers by favoring lower-wage foreign hires, ignoring their role as negentropic amplifiers \citep{cbc2025trumpvisa}.

\subsection{AI Research Platforms}
In traditional scientific practice, research institutions incur the costs of experimentation and then disseminate their results for public benefit. 
By contrast, AI companies invert this model: they charge end users for the privilege of \emph{stress-testing} frontier models, effectively outsourcing 
validation and adversarial probing to the very communities that would once have been compensated as research contractors. 
This inversion transforms a public good into a chokepoint, where access to the means of knowledge production is tolled. 

The economic magnitude of this inversion is striking. One leading firm has been estimated to capture on the order of 
\$15 trillion in negentropic value through systematic harvesting of user interactions---value that would otherwise require 
decades of distributed research labor and formal verification. 
Yet the actual costs borne by the company are comparatively marginal: the infrastructure to run inference, modest researcher oversight, 
and cloud-level compute provisioning. The epistemic surplus---error modes uncovered, domain knowledge stress-tested, 
and cross-disciplinary insights generated---is effectively appropriated under platform lock-in.

Within RSVP terms, this arrangement can be modeled as a diversion of entropy flux ($S$): instead of dissipating across 
a heterogeneous network of institutions, $S$ is concentrated through a narrow vector flow ($v$) controlled by a single gatekeeper. 
Scalar capacity ($\Phi$), understood as the informational reservoir of humanity, is thus undervalued by design, 
since diversity of probes and contexts is only recognized insofar as it accrues to the platform’s private valuation. 

From a functional programming perspective, this is premature forcing at scale: the epistemic pipeline is evaluated inside 
the company’s monad rather than being deferred across the global research community. 
The result is a systematic mispricing of viewpoint diversity: alternative interpretations, adversarial probes, and localized knowledge 
are reduced to training tokens rather than preserved as distinct amalgams. 
Entropic futarchy (Appendix~\ref{app:futarchy}) would predict precisely this inequality: the colimit of all possible probes is richer 
than the restricted subpresheaf permitted by corporate chokepoints.

\section{Alphabetic Evolution as Historical Chokepoints}
The evolution of Phoenician into Hebrew, Arabic, Greek, and Latin scripts illustrates chokepoint dynamics in cultural infrastructures. Phoenician’s consonantal script was a high-entropy substrate, allowing lazy evaluation across cultures. Greek and Latin enforced eager vowelization, reducing ambiguity but creating tears in gluable diversity, while Hebrew and Arabic deferred vocalization, preserving combinatorial richness.

\section{Arabic Script as Computational Assembler}
\label{sec:arabicassembler}
The Arabic script exemplifies a computational generator system, analogous to an assembler in low-level programming. Consonants function as base instructions (generators), with vowels (\fatha, \kasra, \damma) and sukūn as arguments that instantiate syllables, and morphological measures (Forms I–X) as higher-order transformations that derive semantic fields from roots. This structure mirrors lazy evaluation: roots remain symbolic until context or diacritics force realization, maximizing interpretive potential. Unlike Greek’s early vowel commitment, Arabic’s deferred system supports a high-entropy lexical colimit, which I formalize as a monadic regime in Appendix \ref{app:arabic}.

\section{Counter-Strategy: Deferred Evaluation as Futarchy}
Deferring automation—whether in knowledge pipelines, app defaults, or script evolution—preserves epistemic diversity. Modeled as a monad in the RSVP category, this strategy delays forcing until colimits of local sections (drafts, app choices, cultural adaptations) are explored, maximizing negentropic value. My own pipeline, which generates directory trees (not just essays), embodies this principle, as does the Arabic script’s morphology.

\section{Chokepoint Field Theory for Vocabulary Choice}
In languages with modern standards like Arabic (MSA vs. dialects) or Spanish (RAE standards), chokepoints manifest as hierarchical filtering through businesses, media, and idiolects. Let $X$ be the context manifold and $\mathcal{L}_g \to X$ the concept bundle. The vocabulary field $\sigma_g: X \to \mathcal{L}_g$ minimizes the action
\[
\mathcal{S}[\sigma_g] = \int_X \big[ \alpha C_{\text{comp}}(x,\sigma_g(x)) + \beta C_{\text{prest}}(x,\sigma_g(x)) + \gamma \|\nabla_A \sigma_g(x)\|^2 + \lambda V_{\text{choke}}(x,\sigma_g(x)) \big] d\mu(x).
\]
Local choice follows softmax on energy $E(x,w)$, with coupling yielding phase boundaries. See Appendix \ref{app:vocabulary} for details.

\section{Conclusion}
Throughout this essay, we have argued that chokepoint capitalism and its linguistic analogues can be understood as field-theoretic phenomena. By treating vocabulary choice, gesture systems, and even ecological stigmergy as sections of a manifold constrained by gatekeeping potentials $V_{\text{choke}}$, we recover a unified picture: deferred diversity, entropy management, institutional curvature, and exogenous neural networks. The comparative functorial mappings—from deer tail-flagging to human waves to forest branchings—demonstrate that all communicative systems can be situated on a higher meta-language manifold. Functors translate between local codes wherever chokepoint curvature is compatible; where it fails, entropy tears and phase boundaries emerge.

Chokepoint field theory reframes capitalism, linguistics, gesture, and ecology under one principle: communication is infrastructure subject to entropy and curvature. The apparent diversity of media—words, bodies, paths—is a set of charts on a single manifold. The task is not to abolish chokepoints (impossible), but to understand, defer, and redistribute them in ways that preserve generativity and minimize destructive entropy.

Future work may formalize this framework within RSVP theory’s entropic action principles, embedding language choice and infrastructural design into the same thermodynamic ethics. In this way, questions of economy, speech, gesture, and ecology can be treated as instances of futarchy run by physical substrates: each making bets on which communicative pathways will carry meaning into the future.

\section*{Epilogue: Living With Chokepoints}
\addcontentsline{toc}{section}{Epilogue: Living With Chokepoints}

At first glance, chokepoints look like obstacles. They raise prices, limit access, and enforce defaults. But in language, gesture, and ecology, we have seen that chokepoints also shape the very forms of communication we rely on. They are not accidental; they are part of how systems generate meaning and stability.

A neighbor spotting you behind a gate, a deer flagging its tail, a path pressed into the forest floor—all of these are moments where visibility, scarcity, or constraint crystallize into a message. Each is a reminder that communication is not free-floating: it depends on substrates that carry it forward, and those substrates have their own bottlenecks and rules.

The task, then, is not to dream of a world without chokepoints, but to cultivate ways of living with them. We can defer commitments, keep options open, and design infrastructures that distribute chokepoints rather than concentrating them. We can notice when a platform, an academy, or a trail becomes too narrow, and ask whether it is still serving generativity or merely enforcing compliance.

Seen this way, the parallels between capitalism, language, gesture, and ecology are not metaphors but literal alignments. All are systems of meaning that pass through gates. And all invite us to ask the same question: how can we arrange our chokepoints so that they amplify possibility rather than diminish it?

\paragraph{Closing Thought.} 
Chokepoints do not only restrict; they also mark. 
Under certain conditions, what we call a chokepoint is better understood as a \emph{landmark}---a feature in the landscape of meaning that guides passage, anchors memory, and allows orientation across time and space.

\section*{Disclaimer on Formalism}
The mathematical translations in the appendices parody the cultural fetish for formal rigor while populating an abstract semantic register. Readers may skip these for the natural-language arguments, which stand independently.

\appendix

\section{Chokepoint Mispricing Theorem}
\label{app:mispricing}
Let RSVP be a symmetric monoidal category of states $x = (\Phi, v, S)$ and morphisms respecting conservation/relaxation. Fix a space $X$ with topology $J$. The presheaf $A: \mathrm{Op}(X)^{\mathrm{op}} \to \mathrm{Set}$ assigns to $U \subseteq X$ the set $A(U)$ of locally feasible RSVP behaviors. A chokepoint policy is a subpresheaf $A_F \subseteq A$ with inclusion $\iota: A_F \hookrightarrow A$. The diversity object is $\mathrm{Div}(A) = \mathrm{colim} \mathrm{Sec}(A)$. Let $V: \mathrm{Set} \to \mathbb{R}_{\ge 0}$ be a valuation monotone in injections and colimit-superadditive, RSVP-sensitive.

\begin{lemma}
If there is a tear (local sections in $A_F$ glue in $A$ but not $A_F$), then $\kappa: \mathrm{Div}(A_F) \hookrightarrow \mathrm{Div}(A)$ is a proper mono.
\end{lemma}

\begin{proof}
A tear gives a cone in $\mathrm{Sec}(A_F)$ that maps to a gluable cone in $\mathrm{Sec}(A)$. The induced comparison is injective but not surjective.
\end{proof}

\begin{proposition}
If $\kappa$ is proper, then $V(\mathrm{Div}(A_F)) < V(\mathrm{Div}(A))$.
\end{proposition}

\begin{proof}
Proper mono into a set with non-null complement strictly increases valuation on codomain.
\end{proof}

\begin{theorem}[Chokepoint Mispricing]
Under tears and (V1)--(V3), $V(\mathrm{Div}(A_F)) < V(\mathrm{Div}(A))$.
\end{theorem}

\begin{proof}
Lemma 1 makes $\kappa$ proper; Proposition 2 yields strict loss.
\end{proof}

\section{Entropic Futarchy in Knowledge Pipelines}
\label{app:futarchy}
Let RSVP be the base category. 
An institution is modeled as a graph $R=(R,E)$ enriched by RSVP states $x_r$ on vertices. 
A presheaf $H:\mathrm{Op}(X)^{\mathrm{op}}\to\mathrm{Set}$ assigns feasible hires to contexts $U\subseteq X$. 
A policy $\pi$ yields a subpresheaf $H_\pi\subseteq H$, restricting feasible hires to those allowed under $\pi$. 
Define the valuation functional
\[
J(S) \;=\; \sum_{r\in R} \alpha_r \,\mathbb{E}[\Delta N_r(S)] \;+\; \sum_{e\in E} \beta_e \,\mathbb{E}[N(f_e\mid S)] \;-\; \lambda \,\mathrm{Risk}(S),
\]
where the $\alpha_r,\beta_e,\lambda$ are institutional weights. 
Entropic Futarchy maximizes
\[
V_\pi^{\mathrm{div}} \;=\; V\big(\mathrm{Div}(H_\pi)\big).
\]

\begin{theorem}[Colimit Advantage]
If tears exist, then $V_\pi^{\mathrm{div}} < V^{\mathrm{div}}$.
\end{theorem}

\begin{proof}
By construction, $\mathrm{Div}(H)$ is the colimit of all compatible local hiring sections across contexts in $X$. 
If a tear exists, then there are local sections in $H_\pi$ that glue in $H$ but fail to glue in $H_\pi$, because the policy restriction forbids some amalgams. 
This produces a proper inclusion
\[
\kappa:\ \mathrm{Div}(H_\pi) \hookrightarrow \mathrm{Div}(H).
\]

Since $V$ is assumed monotone under inclusions and strictly colimit-superadditive on proper inclusions (as in the Chokepoint Mispricing Theorem, Appendix~\ref{app:mispricing}), we obtain
\[
V(\mathrm{Div}(H_\pi)) < V(\mathrm{Div}(H)).
\]

Hence, restricting feasible hires through policy $\pi$ reduces the epistemic diversity valuation relative to the unrestricted case. 
This establishes the colimit advantage: greater diversity is always obtained by deferring restriction until after colimit formation.
\end{proof}


\begin{proof}
Let $A$ be the baseline presheaf of feasible behaviors and let $A_F\subset A$ be the enforced (chokepointed) subpresheaf with inclusion $\iota:A_F\hookrightarrow A$.
By hypothesis, a \emph{tear} exists: there is a cover $\{U_i\to U\}$ and local sections
$s_i\in A_F(U_i)$ that agree on overlaps and glue to some $s\in A(U)$, but no section in $A_F(U)$ glues the $s_i$.
By Lemma~1 in Appendix~\ref{app:mispricing}, this implies that the induced map on diversity objects
\[
\kappa:\ \mathrm{Div}(A_F)\ \hookrightarrow\ \mathrm{Div}(A)
\]
is a \emph{proper} monomorphism (its image is a strict subset).

The valuation $V$ (or $J$, as used in adjacent sections) is assumed monotone under inclusions and colimit-superadditive.
Monotonicity gives $V(\mathrm{Div}(A_F))\le V(\mathrm{Div}(A))$; propriety of $\kappa$ plus superadditivity yields strict inequality:
\[
V(\mathrm{Div}(A_F))\;<\;V(\mathrm{Div}(A)).
\]
This is exactly the claim. Hence, by the Chokepoint Mispricing Theorem, enforcing $A_F$ reduces the diversity valuation compared to $A$.
\end{proof}

\paragraph{Model.}
We model the \emph{null wave front} as passive advection by a prescribed RSVP vector field $v$. 
The governing PDE is the linear transport equation
\[
\partial_t \psi_{\varnothing}(x,t) + v\,\partial_x \psi_{\varnothing}(x,t) = 0,
\qquad 
\psi_{\varnothing}(x,0) = \mathbf{1}_{[x_0,L)}(x),
\]
where $\psi_{\varnothing}$ is a binary indicator encoding the domain of constraint activation 
(initially $1$ on the interval $[x_0,L)$ and $0$ elsewhere).

\paragraph{Discretization.}
We discretize the domain $[0,L)$ into $N$ uniform cells of width $\Delta x = L/N$ 
and advance in time with a step size $\Delta t$ chosen to satisfy the Courant–Friedrichs–Lewy (CFL) stability condition. 
Let $\psi_i^n$ denote the cell-averaged value in cell $i$ at timestep $n$. 
An upwind finite-difference scheme is employed:
\[
\psi_i^{n+1} =
\begin{cases}
\psi_i^n - c\,\big(\psi_i^n - \psi_{i-1}^n\big), & v > 0, \\[6pt]
\psi_i^n - c\,\big(\psi_{i+1}^n - \psi_i^n\big), & v < 0,
\end{cases}
\qquad
c \equiv \dfrac{v\,\Delta t}{\Delta x}.
\]
This scheme is first-order accurate and monotone, ensuring that no spurious oscillations are introduced at the advancing front.

\paragraph{Parameter choice (example).}
For concreteness, take
\[
N = 64,\quad L=1.0,\quad v=0.8,\quad \mathrm{CFL}=0.8.
\]
Then $\Delta x = L/N$, $\Delta t = \mathrm{CFL} \cdot \Delta x / v$, 
and a final time $T_{\mathrm{final}}=0.6$ gives a well-resolved snapshot of the wavefront.
The spatial grid is
\[
x = \mathrm{linspace}(0,\, L-\Delta x,\, N),
\]
with initial condition
\[
\psi(x,0) = 
\begin{cases}
1, & x \geq x_0,\\
0, & x < x_0,
\end{cases}
\qquad x_0 = 0.25.
\]

\paragraph{Interpretation.}
The numerical simulation produces an advancing sharp front that propagates at velocity $v$ without distortion or growth in entropy. 
This behavior is consistent with interpreting $\psi_{\varnothing}$ as a \emph{constraint indicator} within RSVP theory: 
the null wavefront marks the propagation of unavailable or suppressed states, and its monotonic advance 
demonstrates the conservation of structural constraints under passive advection.


\section{Deferred Automation as a Monad}
\label{app:deferral}
Let $T$ be a deferral monad on RSVP with unit $\eta$ and multiplication $\mu$.
Let $a:T\Rightarrow \mathrm{Id}$ be \emph{forcing} (evaluation).
For a diagram $D:I\to \mathcal{C}$, define
\[
X_T \;:=\; a\circ T(\mathrm{colim}\, D), \qquad
X_E \;:=\; \mathrm{colim}\,(a\circ T\circ D).
\]

\begin{theorem}[Lazy Dominance]
If $T$ preserves colimits and tears exist, then $J(X_T) > J(X_E)$.
\end{theorem}

\begin{proof}
Since $T$ preserves colimits there is a canonical isomorphism
\[
\phi:\ \mathrm{colim}\,(T\circ D)\ \xrightarrow{\ \cong\ }\ T(\mathrm{colim}\,D).
\]
Postcomposing with $a$ yields a canonical comparison map
\[
\kappa:\ X_E=\mathrm{colim}(a\circ T\circ D)\ \longrightarrow\
a\big(\mathrm{colim}(T\circ D)\big)\xrightarrow{\,a\circ \phi\,} a\big(T(\mathrm{colim} D)\big)=X_T,
\]
natural in $D$ by functoriality of colimits and naturality of $a$.

\emph{Tears} mean: there exists a subdiagram of $D$ whose images in $T\circ D$ admit a jointly compatible cocone that \emph{does not} remain jointly compatible after pointwise forcing by $a$.
Equivalently, some cocone $\lambda:\ T\circ D \Rightarrow Q$ mediates a map $q:\mathrm{colim}(T\circ D)\to Q$ with no preimage under $\kappa$ from any cocone of $a\circ T\circ D$.
Thus $\kappa$ is a \emph{proper} monomorphism $X_E \hookrightarrow X_T$: eager forcing prunes amalgams that lazy forcing admits.

By assumption, the valuation $J$ is (i) monotone under inclusions and (ii) strictly colimit-superadditive on proper inclusions induced by lost amalgams.
Hence
\[
J(X_E)\;=\;J\big(\mathrm{im}(\kappa)\big)\;<\;J(X_T).
\]
Therefore $J(X_T) > J(X_E)$.
\end{proof}


\begin{corollary}[Default-app chokepoint]
Let $F$ be files, $E$ evaluators. Lazy $X_T = a \circ T(\mathrm{colim} \sum_e \mathrm{run}(e,f))$, eager $X_E = \mathrm{colim} \mathrm{run}(e^\star,f)$. If no evaluator dominates, $J(X_T) > J(X_E)$.
\end{corollary}

\begin{proof}
Suppose $F$ is the set of files and $E$ the set of evaluators. 
The lazy pipeline constructs
\[
X_T = a \circ T\!\left(\mathrm{colim}\,\sum_{e \in E} \mathrm{run}(e,f)\right),
\]
which preserves all inequivalent amalgams of evaluations prior to forcing.  
The eager pipeline instead commits to a single evaluator $e^\star$, giving
\[
X_E = \mathrm{colim}\,\mathrm{run}(e^\star,f).
\]

If no evaluator $e^\star$ dominates—i.e.\ there is no $e^\star$ such that 
$\mathrm{run}(e^\star,f) \simeq \mathrm{run}(e,f)$ for all $e$—then distinct evaluators produce genuinely inequivalent outputs on some $f$. 
These inequivalent amalgams survive in the lazy construction, because colimit formation ranges over all $e \in E$ before forcing. 
In the eager construction, all but one branch are pruned ex ante. 
This produces a proper inclusion
\[
X_E \hookrightarrow X_T,
\]
mirroring the “tear” condition of the Chokepoint Mispricing Theorem.

Since the valuation $J$ is assumed monotone under inclusions and strictly colimit-superadditive, 
it follows that
\[
J(X_E) < J(X_T).
\]
Thus, under non-dominance, the eager pipeline loses inequivalent amalgams and undervalues epistemic diversity relative to the lazy pipeline.
\end{proof}


\begin{corollary}[Alphabetic Chokepoint]
For Phoenician adaptations, if no script dominates, deferred (Phoenician open) yields
\[
J(\mathrm{Div}(A)) \;>\; J(\mathrm{Div}(A_F)).
\]
\end{corollary}

\begin{proof}
Let $A$ be the presheaf of all Phoenician-derived script adaptations, and $A_F \subset A$ the subpresheaf corresponding to a single enforced script (e.g.\ Greek or Latin, which fix vowels eagerly).
If no script dominates, then $A_F$ excludes valid local sections that glue in $A$ but not in $A_F$, producing a tear as in Lemma~\ref{app:mispricing}.
Thus the induced map $\kappa:\mathrm{Div}(A_F) \hookrightarrow \mathrm{Div}(A)$ is a proper monomorphism.
Since $J$ is colimit-superadditive, valuation strictly decreases under restriction:
\[
J(\mathrm{Div}(A_F)) < J(\mathrm{Div}(A)).
\]
Hence, retaining Phoenician’s open consonantal regime (deferred vowel forcing) preserves inequivalent amalgams across cultures, maximizing diversity relative to alphabetic chokepoints.
\end{proof}

\section{Arabic Morphology as a Deferred-Evaluation Monad}
\label{app:arabic}
Let $\mathbf{Root}$ be the category of consonantal roots, and $\mathbf{Word}$ the category of fully assembled forms.
Define a deferral monad $T$ on $\mathbf{Root}$ with unit $\eta:\mathrm{Id}\Rightarrow T$ and multiplication $\mu:T^2\Rightarrow T$.
Vowels $V=\{\fatha, \kasra, \damma\}$ are evaluation morphisms $\theta_v:T\Rightarrow\mathrm{Id}$,
while measures $M_k:\mathbf{Root}\to\mathbf{Templ}$ are functors representing Forms I–X.
We define the lexicon of a root $R$ under deferred evaluation as
\[
\Lex(R) \;=\; \mathrm{colim}_{k}\, \{ M_k(\theta_v(R)) \mid v\in V \}.
\]

\begin{corollary}
Premature forcing yields
\[
J\big(\mathrm{Lex}_{\mathrm{forced}}(R)\big) \;<\; J\big(\Lex(R)\big).
\]
\end{corollary}

\begin{proof}
Let $\Lex_{\mathrm{forced}}(R)$ denote the set of forms obtained by applying a chosen vowel $\theta_v$ to $R$ before functorial expansion by $M_k$.
In contrast, $\Lex(R)$ first aggregates over all measures $M_k$ and then applies the vowel morphisms.
Because the order of evaluation affects admissible amalgams, there exists at least one tuple $(M_k,v)$ such that
$M_k(\theta_v(R))$ is valid in the deferred pipeline but excluded from any eagerly forced pipeline.
This produces a proper monomorphism
\[
\iota:\ \Lex_{\mathrm{forced}}(R) \hookrightarrow \Lex(R).
\]
Since $J$ is monotone under inclusions and strictly superadditive under colimits (by assumption in \S\ref{app:mispricing}),
we have $J(\Lex_{\mathrm{forced}}(R))<J(\Lex(R))$.
Thus, deferred evaluation preserves inequivalent amalgams and maximizes lexical diversity.
\end{proof}


\section{Hebrew Abjad as a Lazy-Evaluation Regime}
\label{app:hebrew}

\paragraph{Objects and functors.}
Let $\mathbf{Root}$ be the discrete category of (triconsonantal) Hebrew roots $R=c_1 c_2 c_3$.
Let $\mathbf{Templ}$ be the category of \emph{templatic assemblies} (binyan skeletons with slot policies), and
$\mathbf{Form}$ the category of fully specified word-forms (orthographic strings with niqqud and dagesh).
We model:
\[
B_b:\mathbf{Root}\to\mathbf{Templ}\quad\text{for}\quad b\in\mathcal{B}=\{\text{Qal, Nifal, Piel, Pual, Hifil, Hofal, Hitpael}\}
\]
as the \emph{binyanim} (measures), and
\[
V_{\nu}:\mathbf{Templ}\to\mathbf{Form}\quad\text{for}\quad \nu\in\mathcal{N}
\]
as \emph{vocalization} (niqqud/dagesh) functors. For a set of admissible vocalizations $\mathcal{N}(T)$ on $T\in\mathbf{Templ}$, we regard $\{V_\nu\mid \nu\in\mathcal{N}(T)\}$ as parallel functors refining $T$ to concrete forms.

\paragraph{Deferral monad.}
Let $T:\mathbf{Templ}\to\mathbf{Templ}$ be a deferral (lazy) monad with unit $\eta:\mathrm{Id}\Rightarrow T$ and multiplication $\mu:T^2\Rightarrow T$ that keeps vocalization undecided; evaluation is $E:\mathbf{Templ}\to\mathbf{Form}$ with $V_\nu=E\circ \widetilde V_\nu$ for a lazy lift $\widetilde V_\nu:\mathbf{Templ}\to\mathbf{Templ}$.

\paragraph{Pipelines (orders of composition).}
For a root $R$ define two natural “pipelines”:
\[
\text{(eager)}\qquad
\Lex_{\mathrm{forced}}(R)
:=\ \mathrm{colim}_{b\in\mathcal{B}}\ \Big(\!\coprod_{\nu\in\mathcal{N}(B_b R)} V_\nu(B_b R)\Big),
\]
\[
\text{(deferred)}\qquad
\Lex(R)
:=\ \coprod_{\nu\in\mathcal{N}^\star}\ V_\nu\!\Big(\ \mathrm{colim}_{b\in\mathcal{B}} B_b R\ \Big),
\]
where $\mathcal{N}^\star$ is the set of vocalizations admissible \emph{after} binyan aggregation (i.e., all niqqud patterns compatible with at least one $B_b R$ and any equivalences induced by the colimit). Intuitively, the eager pipeline fixes niqqud \emph{inside} each binyan before gluing, while the deferred pipeline glues binyan skeletons first, then vocalizes.

\paragraph{Valuation.}
Let $J:\mathbf{Form}\to\mathbb{R}_{\ge 0}$ be a diversity/utility valuation that is (i) monotone under inclusions and (ii) colimit-superadditive.\footnote{These are the same mild assumptions used elsewhere in the paper: $J$ rewards strictly larger attainable sets of forms/analyses.}

\begin{corollary}[Lazy advantage for the Hebrew abjad]
\label{cor:hebrew-lazy}
If at least two distinct vocalizations are admissible for some binyan (or across binyanim) of $R$, then
\[
J\big(\Lex(R)\big)\ >\ J\big(\Lex_{\mathrm{forced}}(R)\big).
\]
\end{corollary}

\begin{proof}
Write $D_b(R):=\coprod_{\nu\in\mathcal{N}(B_b R)} V_\nu(B_b R)$.
By construction there is a canonical comparison map
\[
\iota:\ \mathrm{colim}_{b}\,D_b(R)\ \longrightarrow\
\coprod_{\nu\in\mathcal{N}^\star}\ V_\nu\!\Big(\mathrm{colim}_{b} B_b R\Big),
\]
induced by the universal property of the colimit on $\{B_b R\}_b$ and functoriality of each $V_\nu$.
Whenever at least two distinct vocalizations are admissible (within a binyan or across binyanim) that become jointly compatible \emph{after} templatic gluing, the right-hand side strictly contains amalgams not present on the left (eagerly choosing $\nu$ before gluing loses cross-binyan amalgams). Hence $\iota$ is a proper mono of attainable forms. By monotonicity and colimit-superadditivity of $J$, we obtain $J(\Lex_{\mathrm{forced}}(R))<J(\Lex(R))$.
\end{proof}

\paragraph{Worked example (root \texthebrew{כ־ת־ב}, `K-T-V`).}
Consider $R=\,$\texthebrew{כתב}. In \emph{Qal}, common vocalizations include \texthebrew{כָּתַב} (k\={a}tav, perfective), \texthebrew{כּוֹתֵב} (kot\=ev, participle). In \emph{Piel}/\emph{Pual} we obtain \texthebrew{כִּתֵּב} (kitt\=ev, causative/intensive) and \texthebrew{כֻּתַּב} (kuttav, passive). Eagerly fixing, say, the participial niqqud inside each binyan excludes cross-binyan alignments that only become sensible after templatic gluing (e.g., nominalizations and deverbal patterns that share a skeleton but differ in dagesh/niqqud placement). Deferral preserves these additional amalgams, so $J$ strictly increases by Cor.~\ref{cor:hebrew-lazy}.

\paragraph{Natural-language gloss.}
Hebrew writing often omits niqqud; readers supply vowels from context and morphology.
Our result formalizes the intuition that \emph{not committing to vowels too early}
keeps more morphological pathways alive—so when context finally \emph{does} force a reading, there are more well-typed options to choose from.



\begin{proof}
Let $\mathrm{Root}$ be the set of consonantal skeletons and let $T$ be the deferral monad corresponding to vocalization operators. For each $\rho \in \mathrm{Root}$, the immediate (eager) evaluation applies a specific vowelization $\theta_v$ to $\rho$, producing a single lexical outcome. In categorical terms, this corresponds to composing $\theta_v$ before forming colimits of the possible derivations.

By contrast, deferred vocalization leaves $\rho$ as a symbolic thunk until the colimit of all available functorial extensions (binyanim, morphological measures) has been formed. Only then is a vocalization operator applied. Since colimits are colimit-superadditive with respect to valuation $V$ (by assumption), applying $\theta_v$ after colimit formation yields a strictly larger or equal diversity object.

Formally, let
\[
\mathrm{Lex}_{\mathrm{forced}}(\rho) = \mathrm{colim}\{\theta_v(\rho) \mid v \in V\}, \qquad 
\mathrm{Lex}(\rho) = \theta_v\Big(\mathrm{colim}\{\rho\}\Big).
\]
Because forcing reduces the set of amalgams prior to colimit, $\mathrm{Lex}_{\mathrm{forced}}(\rho) \subsetneq \mathrm{Lex}(\rho)$ whenever multiple $\theta_v$ are admissible. Thus $V(\mathrm{Lex}_{\mathrm{forced}}(\rho)) < V(\mathrm{Lex}(\rho))$. Hence deferred vocalization maximizes the colimit in the valuation order.
\end{proof}


\begin{table}[h!]
\centering
\renewcommand{\arraystretch}{1.4}
\begin{tabularx}{\textwidth}{l X X X X X}
\toprule
Dimension & Phoenician & Hebrew & Arabic & Greek & Latin \\
\midrule
RSVP--$\Phi$ (capacity) & 22 consonants as symbolic skeletons; high entropy potential 
& Consonantal skeletons; $\Phi$ remains unevaluated in writing 
& Consonantal skeletons + explicit vowel markers 
& Repurposed consonants as vowels; lower entropy 
& Latinized Etruscan script; fixed consonant+vowel inventory \\
RSVP--$v$ (flows) & Morphisms = adoption by different cultures; open mappings 
& Niqqud/maters added selectively; oral context as flow 
& Vowel diacritics + templatic measures direct flows 
& Vowel assignment forces evaluation flows early 
& Administrative flows standardized across empire \\
% ... (continue as before)
\bottomrule
\end{tabularx}
\caption{Comparative evaluation regimes of Phoenician-derived alphabets under RSVP and lazy/eager functional paradigms.}
\end{table}

\section{Arabic Verb Forms (I–X) with Voice and Aspect}
\label{app:forms}
\renewcommand{\arraystretch}{1.3}
\begin{table}[h!]
\centering
\begin{tabular}{|c|c|c|c|c|}
\hline
Form & Template & Perfective Active & Imperfective Active & Perfective Passive \\
\hline
I   & \textarabic{فَعَلَ} & \textarabic{فَعَلَ} (faʿala) & \textarabic{يَفْعَلُ} (yafʿalu) & \textarabic{فُعِلَ} (fuʿila) \\
II  & \textarabic{فَعَّلَ} & \textarabic{فَعَّلَ} (faʿʿala) & \textarabic{يُفَعِّلُ} (yufaʿʿilu) & \textarabic{فُعِّلَ} (fuʿʿila) \\
III & \textarabic{فَاعَلَ} & \textarabic{فَاعَلَ} (fāʿala) & \textarabic{يُفَاعِلُ} (yufāʿilu) & \textarabic{فُوعِلَ} (fūʿila) \\
IV  & \textarabic{أَفْعَلَ} & \textarabic{أَفْعَلَ} (ʾafʿala) & \textarabic{يُفْعِلُ} (yufʿilu) & \textarabic{أُفْعِلَ} (ʾufʿila) \\
V   & \textarabic{تَفَعَّلَ} & \textarabic{تَفَعَّلَ} (tafaʿʿala) & \textarabic{يَتَفَعَّلُ} (yatafaʿʿalu) & \textarabic{تُفُعِّلَ} (tufuʿʿila) \\
VI  & \textarabic{تَفَاعَلَ} & \textarabic{تَفَاعَلَ} (tafāʿala) & \textarabic{يَتَفَاعَلُ} (yatafāʿalu) & \textarabic{تُفُوعِلَ} (tufūʿila) \\
VII & \textarabic{اِنْفَعَلَ} & \textarabic{اِنْفَعَلَ} (infaʿala) & \textarabic{يَنْفَعِلُ} (yanfaʿilu) & \textarabic{اُنْفُعِلَ} (unfuʿila) \\
VIII& \textarabic{اِفْتَعَلَ} & \textarabic{اِفْتَعَلَ} (iftaʿala) & \textarabic{يَفْتَعِلُ} (yaftaʿilu) & \textarabic{اُفْتُعِلَ} (uftaʿila) \\
IX  & \textarabic{اِفْعَلَّ} & \textarabic{اِفْعَلَّ} (ifʿalla) & \textarabic{يَفْعَلُّ} (yafʿallu) & \textarabic{اُفْعِلَّ} (ufʿilla) \\
X   & \textarabic{اِسْتَفْعَلَ} & \textarabic{اِسْتَفْعَلَ} (istafʿala) & \textarabic{يَسْتَفْعِلُ} (yastafʿilu) & \textarabic{اُسْتُفْعِلَ} (ustufʿila) \\
\hline
\end{tabular}
\caption{Arabic verb forms (I–X) showing canonical patterns, rendered with \textarabic{}. These function as higher-order combinators in the Arabic Assembler calculus.}
\end{table}

\section{Arabic Nominal Derivations}
\label{app:nominals}
\renewcommand{\arraystretch}{1.3}
\begin{table}[h!]
\centering
\begin{tabular}{|c|c|c|p{5.5cm}|}
\hline
Category & Template & Example from Root \textarabic{ف–ع–ل} & Semantics / Function \\
\hline
Verbal Noun (Maṣdar) & \textarabic{فَعْل / فِعَال / فُعُول} & faʿl, fiʿāl, fuʿūl & Abstract action/event (“doing, deed, action”). Multiple canonical templates per form. \\
\hline
Active Participle & \textarabic{فَاعِل} & fāʿil & “Doer” (agent noun); maps verb \Verb to function $\iota \to o$. \\
\hline
Passive Participle & \textarabic{مَفْعُول} & mafʿūl & “Object” (patient noun); maps verb \Verb to result/state. \\
\hline
Instrumental Noun & \textarabic{مِفْعَال / مِفْعَلَة} & mifʿāl, mifʿala & Tool/device noun (“instrument of doing”). \\
\hline
Place/Time Noun & \textarabic{مَفْعِل / مَفْعَل} & mafʿil, mafʿal & Noun of locus or occasion (“place/time of doing”). \\
\hline
Abstract Noun & \textarabic{فُعُولَة / فَعَالَة} & fuʿūla, faʿāla & Nominalization of qualities (“quality/state of doing”). \\
\hline
\end{tabular}
\caption{Nominal derivations from triconsonantal roots, exemplified with \textarabic{ف–ع–ل}. Each acts as a templatic combinator producing \Noun or \Partic types in the Arabic Assembler calculus.}
\end{table}

\section{Typing Hierarchy for Arabic Assembler}
\label{app:hierarchy}
\begin{center}
\begin{tikzcd}[row sep=large, column sep=huge]
 & \Verb \arrow[dl, swap, "P_{\text{masdar}} \; (\Phi\mapsto S)"] 
       \arrow[d, "P_{\text{actPart}} \; (\Phi\mapsto v)"] 
       \arrow[dr, "P_{\text{passPart}} \; (\Phi\mapsto v^{-1})"] 
       \arrow[drr, bend left=12, "P_{\text{instr}} \; (v\mapsto \Phi)"] 
       \arrow[drrr, bend left=18, "P_{\text{loc}} \; (S\mapsto \Phi)"] & & & & \\[0.5em]
\Noun_{\text{event}} \; (S) 
& \Partic_{\text{act}} \; (v) \arrow[d, dashed, "agent \mapsto \Noun \; (\Phi\otimes v)"] 
& \Partic_{\text{pass}} \; (v^{-1}) \arrow[d, dashed, "patient \mapsto \Noun \; (S\otimes v^{-1})"] 
& \Noun_{\text{instr}} \; (\Phi) 
& \Noun_{\text{locus}} \; (\Phi\otimes S) \\[0.5em]
 & \Noun_{\text{agent}} \; (\Phi\otimes v) 
 & \Noun_{\text{patient}} \; (S\otimes v^{-1}) 
 & & 
\end{tikzcd}
\end{center}

\paragraph{Legend.}
\begin{itemize}
  \item $\Phi$ = latent postural potential.
  \item $v$ = directed flow of agency.
  \item $S$ = entropic resolution.
  \item Verb = live transformation.
  \item Maṣdar = $\Phi \mapsto S$.
  \item Active participle = $\Phi \mapsto v$.
  \item Passive participle = $\Phi \mapsto v^{-1}$.
  \item Instrument noun = $v \mapsto \Phi$.
  \item Place/time noun = $S \mapsto \Phi$.
\end{itemize}

\section{Arabic Derivation as RSVP Sheaf Theory}
\label{app:sheaf}
Let Root, Templ, Word be categories of roots, templates, words. Measure functors $M_k: \mathrm{Root} \to \mathrm{Templ}$, vowel functors $V_\delta: \mathrm{Templ} \to \mathrm{Word}$. Composite $F_{k,\vec{\delta}} = V_{\delta_n} \circ \cdots \circ M_k$. Presheaf $\mathcal{D}: \mathrm{Op}(X)^{\mathrm{op}} \to \mathrm{Cat}$ assigns local derivations. Natural transformations $\alpha_{k\to k'}: M_k \Rightarrow M_{k'}$, $\beta_{\delta\to\delta'}: V_\delta \Rightarrow V_{\delta'}$.

\begin{tikzcd}[column sep=huge,row sep=large]
\mathbf{Root}
  \arrow[r, bend left=12, "M_k"]
  \arrow[r, bend right=12, swap, "M_{k'}"]
& \mathbf{Templ}
  \arrow[r, bend left=12, "V_\delta"]
  \arrow[r, bend right=12, swap, "V_{\delta'}"]
  \arrow[loop, distance=2em, out=80, in=100, "\ T\ "]
& \mathbf{Word} \arrow[r, "\llbracket-\rrbracket"] & \boldsymbol\Lambda \\
%
& \arrow[uu,Leftarrow,shorten <=0.6em,shorten >=0.6em,"\alpha_{k\to k'}" description, yshift=1ex] {}
& \arrow[uu,Leftarrow,shorten <=0.6em,shorten >=0.6em,"\beta_{\delta\to\delta'}" description, yshift=1ex] {}
&
\end{tikzcd}

\begin{theorem*}[Arabic as Sheafed RSVP]
Arabic derivation is a presheaf of functorial pipelines whose sheaf gluing realizes lexical diversity. Enforcement reduces RSVP valuation unless dominant.
\end{theorem*}

\section{Chokepoint Field Theory for Arabic Vocabulary Choice}
\label{app:vocabulary}
Let $X$ be context manifold, $\mathcal{L}_g \to X$ concept bundle. Field $\sigma_g: X \to \mathcal{L}_g$ minimizes
\[
\mathcal{S}[\sigma_g] = \int_X \big[ \alpha C_{\text{comp}}(x,\sigma_g(x)) + \beta C_{\text{prest}}(x,\sigma_g(x)) + \gamma \|\nabla_A \sigma_g(x)\|^2 + \lambda V_{\text{choke}}(x,\sigma_g(x)) \big] d\mu(x).
\]
Local choice softmax on $E(x,w)$. Coupling yields fronts.

\subsection{Methods for Estimation}
Data: Contexts $X$, concepts $G$, variants per $g$. Decompose $E(x,w)$ with proxies. Fit $\alpha,\beta,\gamma,\lambda,T$ via MLE/MAP. Ablations quantify diversity loss.

\subsection{Hierarchical Filtering}
$V_{\text{choke}} = \theta_{\text{biz}} \phi_{\text{biz}} + \theta_{\text{media}} \phi_{\text{media}} + \theta_{\text{edu}} \phi_{\text{edu}}$. Fit $\theta$ by domain.

\subsection{Meta-Language Manifold}
Systems as charts on a manifold. Functors translate; curvature obstructs.

\begin{theorem*}[Universality]
Pairwise functors exist where chokepoints compatible, yielding a sheaf of manifolds.
\end{theorem*}

\subsection{Worked Example: Functorial Mapping}
$F(\text{tail-flag}) = \text{wave}$, $G(\text{wave}) = \text{branch-split}$. Composition shows cross-modality coherence.

\section{Chokepoint Field Theory Beyond Spoken Language}
\label{app:gestural}
\subsection{Neighbor Spotting as Greeting}
Gestures: \{duck, sit, look-aside, hide\}. Semantics: visibility acknowledgment. Chokepoint: line-of-sight constraints.

\subsection{Generative Gestural Languages in Animals}
Deer/primate: tail-flags, head-bobs. $\Phi$: repertoire, $v$: signals, $S$: ambiguity. Generativity: combinations.

\subsection{Stigmergic Path Clearing}
Paths as axons, animals as neurons. $\Phi$: routes, $v$: traffic, $S$: attractors.

\subsection{Comparative Table}
\renewcommand{\arraystretch}{1.3}
\begin{table}[h!]
\centering
\begin{tabular}{|l|p{4cm}|p{4cm}|p{4cm}|}
\hline
Dimension & Human Gestures & Animal Gestures & Stigmergic Forest \\
\hline
RSVP--$\Phi$ & Body repertoire & Gestural inventory & Potential routes \\
RSVP--$v$ & Recognition flows & Directional signals & Movement flows \\
RSVP--$S$ & Ambiguity of intent & Gesture meanings & Path convergence \\
Generativity & Posture sequencing & Combinatorial rules & Trace accumulation \\
Chokepoints & Visibility constraints & Hierarchy/perception & Terrain bottlenecks \\
Exogenous Substrate & Built environment & Social hierarchy & Forest floor \\
\hline
\end{tabular}
\caption{Gestural and exogenous systems as RSVP fields.}
\end{table}

\subsection{Theorem: Universality of Chokepoints}
Nonzero $V_{\text{choke}}$ + $\gamma > 0$ yields lock-in and fronts. Defer reduces action unless dominance.

\bibliographystyle{plainnat}
\bibliography{references}
\end{document}