\section{Mathematical Formalism}
\subsection{Categories of Context and Generation}
Let $\C$ denote the category of contexts:
\begin{itemize}
    \item Objects: corpora $\U$ (dialogue histories, artistic traditions, datasets).
    \item Morphisms $f: \U \to \V$: inclusions or transformations (e.g., genre shifts).
\end{itemize}
Let $\M$ denote the category of generative models:
\begin{itemize}
    \item Objects: models trained on corpora.
    \item Morphisms: fine-tunings or transfer functors.
\end{itemize}
A conversation is a functor $\F: \C_{\text{conv}} \to \M_{\text{utterances}}$, with utterances as samples from $\F(\U)$. An artwork is an endomorphism $a: \M \to \M$, reconfiguring model weights. An oeuvre is a subcategory $\Ocal \subseteq \M$ closed under such endomorphisms.

\subsection{Sheaf-Theoretic Encoding}
Let $(X, \tau)$ be a site of contexts. A presheaf $\F: X^{\text{op}} \to \mathrm{Set}$ assigns generative acts to each open set $\U$. Stages:
\begin{itemize}
    \item \textbf{Gossip}: Presheaf with non-unique gluing (contradictory overlaps).
    \item \textbf{Religion}: Sheafification enforces unique extensions (dogma).
    \item \textbf{Platforms}: Algorithmic sheaves hide provenance (e.g., recommendation engines).
    \item \textbf{Semantic infrastructure}: Derived stacks with homotopy colimits, preserving obstructions as cohomology classes.
\end{itemize}
Global sections are:
\[
\F(U) \simeq \hocolim \left( \prod_i \F(U_i) \rightrightarrows \prod_{i < j} \F(U_{ij}) \underset{\to}{\to} \prod_{i < j < k} \F(U_{ijk}) \cdots \right).
\]

\subsection{RSVP Field Mapping}
Generative acts are fields:
\begin{itemize}
    \item \textbf{Scalar density ($\Phi$)}: Semantic weight, e.g., norm density.
    \item \textbf{Vector flow ($\vvec$)}: Transport of tropes, e.g., conversational flows.
    \item \textbf{Entropy ($\Scal$)}: Contradictions, measured as cohomology classes.
\end{itemize}
Semantic infrastructures preserve $\Scal$; chokepoints expel it ($\Scal \to 0$).

\subsection{Theorem and Corollary}
\textbf{Theorem.} Every conversation and artwork is a functorial generative model trained on its context. \\
\textbf{Proof Sketch.} Construct $\C$ and $\M$. Conversations are functors $\F: \C \to \M$, artworks are endomorphisms in $\M$. Generativity follows functoriality. \hfill $\Box$

\textbf{Corollary.} Suppressing obstructions degenerates conversations/artworks into chokepoint bots. \\
\textbf{Justification.} Rigid sheaves erase cohomology, compressing $\Phi$, bottling $\vvec$, expelling $\Scal$.

\subsection{Energy Functional for Generative Cinema}
For a trope-selected shot with camera trajectory $\p(t)$ and look-at $\lvec(t)$, the cost functional is:
\[
\E[\p, \lvec] = w_c C(\p, \lvec) + w_f F(\p, \lvec) + w_m M(\p) - w_r R(\p, \lvec),
\]
where $C$ is continuity cost, $F$ is framing error, $M$ is motion jerk, and $R$ is trope relevance. Minimizing $\E$ yields cinematic coherence.

\subsection{General Constraint Cycle}
Infrastructures are functors $G: \C \to \M$. Dynamics depend on:
\begin{itemize}
    \item \textbf{Metabolizing}: $H^*(F)$ preserved.
    \item \textbf{Chokepoint}: $H^*(F) \to 0$.
\end{itemize}

\section{Notation}
\begin{itemize}
    \item $\Phi$: Scalar density (semantic load).
    \item $\vvec$: Vector flow (reference transport).
    \item $\Scal$: Entropy (contradictions).
    \item $\C$: Contexts.
    \item $\M$: Generative models.
    \item $\F$: Conversation functor.
    \item $a: \M \to \M$: Artwork endomorphism.
    \item $\hocolim$: Homotopy colimit.
    \item $H^*(F)$: Obstruction cohomology.
    \item $\p(t), \lvec(t)$: Camera functions.
    \item $\E[\p, \lvec]$: Cinema energy functional.
\end{itemize}

\section{Example: From Gossip to Derived Sheaves}
For a site $(X, \tau)$, a presheaf $\F: X^{\text{op}} \to \mathrm{Set}$ assigns claims to contexts $U$. \\
\textbf{Gossip}: $\F(U)$ is narratives; overlaps $U_{ij}$ disagree up to $\sim$. \\
\textbf{Religion}: Sheafification forces agreement, losing ambiguity. \\
\textbf{Platforms}: Algorithmic sheaves select one section, hiding provenance. \\
\textbf{Semantic Infrastructure}: Derived stacks with:
\[
\F(U) \simeq \hocolim \left( \prod_i \F(U_i) \rightrightarrows \prod_{i < j} \F(U_{ij}) \underset{\to}{\to} \prod_{i < j < k} \F(U_{ijk}) \cdots \right).
\]
Obstructions in $H^k(\check{C}(\{U_i\}, \pi_* \F))$ are appeal handles.

\section{Generative Cinema: Shakespeare vs. Netflix}
A generative movie bot uses:
\[
A_k(t) = \alpha U(t) + \beta C_k(t) + \gamma \sum_s w_{s,t} A_{k-1}(s) - \delta R(t) - \eta I_k(t).
\]
\textbf{Shakespeare}: Preserves contradictions (variant quartos) for reinterpretation \citep{shakespeare1623folio}. \\
\textbf{Netflix}: Collapses tropes to metrics, expelling $\Scal$.