\documentclass[11pt]{article}

% --------------------------------------------------
% Packages
% --------------------------------------------------
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{geometry}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{csquotes}
\usepackage{hyperref}

\geometry{margin=1in}
\setstretch{1.15}

% --------------------------------------------------
% Macros
% --------------------------------------------------
\newcommand{\Hist}{\mathcal{H}}
\newcommand{\PhiField}{\Phi}
\newcommand{\VecField}{\vec{v}}
\newcommand{\Entropy}{S}

% --------------------------------------------------
% Title
% --------------------------------------------------
\title{Toward a Top-Level Ontology of Entropic Histories:\\
A Scalar--Vector--Entropy Foundation for Ontology Engineering}
\author{Flyxion}
\date{\today}

\begin{document}
\maketitle

% ==================================================
\begin{abstract}
% ==================================================

Top-level ontologies such as the Basic Formal Ontology have demonstrated the
importance of realist discipline in ontology engineering. However, their core
primitives remain fundamentally static, treating entities as primary and
processes as derivative. This limits their ability to represent irreversibility,
historical dependence, semantic drift, and field-like phenomena now central to
physics, computation, and cognition.

This paper proposes a new top-level ontology grounded in the
Relativistic Scalar--Vector--Entropy Plenum (RSVP). Rather than taking entities
as primitive, the proposed framework treats entropic histories, constrained
flows, and stabilization regimes as fundamental. Objects, relations, agents,
and information arise as low-entropy invariants within a dynamically constrained
plenum.

The resulting ontology preserves realism while extending it beyond
object-centric metaphysics, offering a unified foundation for physical,
biological, cognitive, and computational domains.
\end{abstract}

% ==================================================
\section{Introduction}
% ==================================================

Top-level ontology engineering has reached a stage of maturity at which its
central difficulties no longer concern representational convenience or modeling
style, but the adequacy of its foundational primitives. Persistent problems
surrounding irreversibility, ontology alignment, information, and agency suggest
that existing frameworks are being pushed beyond the regimes for which they were
designed \cite{GuarinoFormalOntology, ArpSmithSpears}.

Realist top-level ontologies such as the Basic Formal Ontology (BFO) have
demonstrated the importance of ontological discipline, particularly through
careful distinctions between continuants and occurrents and a rejection of
conceptualism \cite{SmithBFO, GrenonSmithSNAPSPAN}. However, these frameworks
remain fundamentally entity-centric. Objects are treated as ontologically
primary, while processes, histories, and informational structures are introduced
only secondarily. This orientation limits their expressive power in domains where
irreversibility, historical dependence, and field-like dynamics are essential.

This paper advances the thesis that these limitations are not merely practical
engineering challenges but reflect a deeper ontological misalignment. We argue
that history, entropy, and constrained flow must be treated as ontologically
primitive, and that entities should instead be understood as stabilized
invariants arising within a dynamically constrained plenum.

To this end, we propose a new top-level ontology grounded in the Relativistic
Scalar--Vector--Entropy Plenum (RSVP). The framework preserves realist
commitments while extending ontology beyond static metaphysical atomism. It
offers a unified foundation for physical, biological, cognitive, and
computational domains, and provides principled explanations for both the success
and failure modes of existing ontology engineering practices.

% ==================================================
\section{Limits of Entity-Centric Ontology}
% ==================================================

The motivation for a new top-level ontology arises from a cluster of persistent
problems that resist resolution within entity-centric frameworks. These problems
are often treated independently, but they share a common source: the assumption
that entities are ontologically prior to histories, processes, and constraints.

First, entity-centric ontologies struggle to represent irreversibility as a
constitutive feature of reality. Time is typically encoded as an ordering
relation over entities or processes, rendering history descriptive rather than
generative. This treatment is inadequate for domains in which path dependence is
fundamental, including thermodynamics \cite{Landauer}, biological development
\cite{RosenLifeItself}, learning systems \cite{Hutchins}, and computation itself.
Irreversibility is not an accidental feature of these domains; it is central to
their ontological structure.

Second, ontology interoperability relies heavily on mappings between independently
developed ontologies. Empirically, such mappings are fragile and degrade under
version drift. This fragility is often attributed to insufficient engineering
effort, but it reflects a deeper issue. Static correspondences are being imposed
on systems that evolve under incompatible dynamics. Without a shared dynamical
substrate, mappings amplify entropy rather than reducing it, leading to semantic
instability over time.

Third, the ontological status of information remains unresolved. Information is
alternately treated as an abstract entity, a mental construct, or a property of
physical artifacts \cite{FloridiInformation}. Existing ontologies lack principled
criteria for distinguishing informational structure from its physical carriers
or symbolic representations. This ambiguity becomes acute in computational and
cognitive domains, where informational patterns exert causal and normative force
without being reducible to discrete objects.

Taken together, these difficulties indicate that entity primacy is an
ontological liability. A framework that treats histories, constraints, and
entropy as derivative cannot adequately represent domains in which stability
itself must be explained.

% ==================================================
\section{Design Commitments}
% ==================================================

The RSVP ontology adopts a set of design commitments intended to preserve realist
discipline while addressing the limitations identified above. These commitments
are philosophical rather than methodological, and they constrain the choice of
ontological primitives rather than modeling techniques.

First, the framework endorses realism without substance fixation. Reality exists
independently of minds, but not all real structures are object-like. Stability,
flow, and constraint may be ontologically fundamental even when no enduring
entities can be identified.

Second, irreversibility is treated as primitive. Histories are not derived from
entities; rather, entities arise as stabilized invariants over admissible
histories. This reverses the explanatory direction characteristic of classical
ontology.

Third, entropy is understood ontologically rather than epistemically. Entropy
measures the degeneracy of admissible futures under constraint, not uncertainty
in an observer’s knowledge. This interpretation aligns ontology with
thermodynamic and informational considerations without collapsing into
subjectivism \cite{Jaynes, Shannon}.

Fourth, the framework rejects conceptualism. Ontologies describe the structure
of the world, not the structure of concepts in minds. Semantic stability is a
property of constrained histories, not of linguistic conventions.

Finally, the framework is explicitly compositional across scales. Ontological
structures must admit coherent coarse-graining and refinement, allowing stable
subtheories to emerge within broader dynamical regimes.

% ==================================================
\section{The RSVP Framework}
% ==================================================

The RSVP ontology introduces three irreducible primitives: scalar density, vector
flow, and entropy. These are not imported from physics as substantive claims, but
are adopted as structural abstractions necessary to represent stability,
directionality, and degeneracy across domains.

Scalar density, denoted \(\PhiField\), represents ontic stability. Regions of
high scalar density correspond to structures that persist under perturbation and
maintain identity across admissible continuations. Such regions may correspond
to particles, organisms, institutions, or persistent informational patterns,
depending on scale and context.

Vector flow, denoted \(\VecField\), represents directed constraint propagation.
Causation, inference, control, and functional dependency are all treated as
manifestations of vector flow. Relations are not primitive; they emerge as
stable couplings of directed flows within constrained regions.

Entropy, denoted \(\Entropy\), measures the degeneracy of admissible futures.
Low entropy corresponds to constrained continuation and invariant structure,
while high entropy corresponds to branching, instability, or interpretive
ambiguity. Entropy thus provides a principled criterion for when identity and
representation are sustainable.

These three primitives are mutually irreducible. Scalar density cannot be
reduced to vector flow, as persistence is not equivalent to directionality.
Vector flow cannot be reduced to entropy, as constraint propagation does not
entail degeneracy. Entropy cannot be reduced to scalar density, as stability does
not preclude multiplicity of futures. Classical ontologies implicitly assume
low-entropy regimes in which these distinctions collapse; RSVP makes them
explicit.

Together, \((\PhiField, \VecField, \Entropy)\) define a dynamically constrained
plenum in which ontological structure emerges historically. Entities, processes,
relations, and information are not primitives but stabilized configurations
within this plenum.

% ==================================================
\section{Event-Historical Semantics}
% ==================================================

The RSVP ontology is fundamentally history-first. This section makes that
commitment explicit by introducing an event-historical semantics in which
ontological structure is grounded in admissible histories rather than in static
inventories of entities. The purpose of this semantics is not to replace existing
logical or representational frameworks, but to supply the conditions under which
such frameworks remain ontologically coherent.

A history is understood as an irreversible trajectory governed by constraints
that regulate which continuations remain admissible. These constraints are not
merely descriptive regularities but constitutive conditions that define the space
of possible futures. A history, in this sense, is not a record of what has
occurred, but a structured process of constraint satisfaction unfolding over
time.

Formally, let \(\Hist\) denote the space of all histories admissible under a given
ontological regime. Admissibility is defined relative to constraint sets that may
encode physical laws, biological organization, institutional rules, or semantic
invariants. A history \(h \in \Hist\) is admissible if and only if it admits at
least one continuation consistent with these constraints. Irreversibility is
primitive: histories compose forward in time, but do not admit nontrivial
inverses.

Within this framework, entities are not taken as primitives but arise as
equivalence classes over histories. Two histories are equivalent with respect to
an entity if they preserve the same invariant structure across admissible
continuations. Identity is therefore defined historically: an entity persists
precisely when all admissible futures preserve the relevant invariants. This
account explains why identity is robust in some domains and fragile in others,
without appealing to metaphysical substance.

Entropy plays a central role in this semantics. Entropy measures the branching
factor of admissible continuations from a given historical state. Low entropy
corresponds to constrained continuation and stable invariants, while high entropy
corresponds to proliferating futures and loss of structural coherence. Identity,
reference, and representation are sustainable only in regimes where entropy
remains bounded.

This event-historical perspective clarifies the ontological status of processes.
Processes are not entities that happen to unfold in time; they are structured
regions of history characterized by directed constraint propagation. Their
ontological character is determined by how they restrict future admissibility,
not by their decomposition into temporal parts.

The event-historical semantics thus reverses the explanatory order of classical
ontology. Rather than deriving history from entities, it derives entities from
history. Ontology, on this view, is the study of admissible trajectories and the
conditions under which invariants emerge within them.

% ==================================================
\section{Ontology Mappings and Entropic Instability}
% ==================================================

Ontology mappings are commonly introduced as a solution to semantic
heterogeneity. Within entity-centric frameworks, they are typically treated as
static correspondences between classes, relations, or axioms. Empirically,
however, such mappings are brittle. They require continual maintenance, degrade
under version drift, and often collapse when ontologies evolve independently.
The RSVP framework explains these failures as a consequence of entropic
misalignment rather than engineering deficiency.

From an event-historical perspective, an ontology is not a static artifact but a
constraint system governing admissible histories of representation and use. When
two ontologies are developed independently, they typically impose distinct
constraints on admissible histories. A mapping between them attempts to align
these constraint systems. If the systems evolve under different dynamics, the
space of admissible histories diverges, and the mapping ceases to preserve
meaning.

Formally, a mapping between ontologies requires that admissibility be preserved
under translation. That is, histories admissible under one ontology must map to
histories admissible under the other. When entropy is low, admissible histories
are tightly constrained, and such preservation is possible. When entropy
increases, admissible histories branch, and preservation fails. The mapping does
not merely become inaccurate; it amplifies entropy by introducing incompatible
constraints.

This analysis explains why ontology mappings sometimes succeed in practice.
Mappings are stable when both ontologies operate within low-entropy regimes, such
as highly regulated scientific domains or narrowly defined institutional
contexts. In such cases, admissible histories remain aligned, and translation
preserves invariants. Outside these regimes, persistent mappings are structurally
unstable.

The RSVP framework also clarifies the role of so-called upper ontologies as
anchors. An upper ontology functions not by providing a neutral vocabulary, but
by imposing strong constraints that reduce entropy across admissible histories.
When successful, such anchoring effectively induces convergence toward a shared
low-entropy regime. When unsuccessful, it becomes a source of friction rather
than interoperability.

Importantly, the ideal outcome of ontology mapping is not perpetual translation
but convergence. When two ontologies are successfully aligned, the result is not
a stable mapping layer but the emergence of a new, unified constraint system
under which both sets of practices become admissible. Mappings are therefore
transitional structures whose success is marked by their eventual obsolescence.

By treating mappings as dynamical rather than static objects, RSVP provides a
principled explanation for both their utility and their limits. Ontology
engineering is revealed not as the construction of eternal correspondences, but
as the management of entropy across evolving representational histories.

% ==================================================
\section{Embedding BFO as a Low-Entropy Subtheory}
% ==================================================

This section revisits the relationship between RSVP and established realist
ontologies by providing a clean, non-redundant account of how a representative
fragment of the Basic Formal Ontology (BFO) embeds within RSVP as a stabilized,
low-entropy subtheory. The aim is neither to subsume BFO wholesale nor to critique
its internal coherence, but to explain the conditions under which its categories
are ontologically appropriate.

BFO is explicitly realist and entity-centric. Its core distinctions between
continuants and occurrents, together with participation relations, presuppose the
existence of stable identity conditions across time. These presuppositions are
not trivial; they restrict the space of admissible histories to those in which
identity-preserving continuations dominate. In RSVP terms, BFO operates entirely
within regimes characterized by high scalar density and bounded entropy.

Within such regimes, continuants correspond to regions of the plenum that remain
structurally invariant across admissible histories. Scalar density remains
sufficiently high that perturbations do not induce branching that would
undermine identity. Occurrents correspond to directed flows constrained by these
regions, rather than to free-standing processes. Participation relations remain
well-defined precisely because entropy remains low enough to prevent divergence
of admissible continuations.

This interpretation explains both the strength and the scope limitations of BFO.
In domains such as anatomy, materials science, and regulated institutional
contexts, the assumption of low entropy is empirically justified. Structures
persist, processes unfold in predictable ways, and participation relations can
be treated as stable. In these contexts, BFO’s categories are not merely useful
but ontologically appropriate.

However, the same assumptions become liabilities in domains characterized by
rapid change, versioning, feedback, or learning. In such domains, admissible
histories branch, scalar density fluctuates, and identity conditions erode. RSVP
predicts that entity-centric ontologies will struggle in these regimes, not
because they are poorly designed, but because the ontological conditions they
presuppose no longer obtain.

The embedding of BFO within RSVP therefore preserves compatibility without
reduction. RSVP does not reinterpret continuants as processes or dissolve entity
ontology into flux. Instead, it explains why continuants exist when they do, and
why they fail when entropy rises. BFO emerges as a special case within a broader
ontological landscape rather than as a universal foundation.

This account also reframes debates about philosophical priority. Ontology does
not bifurcate into philosophical versus engineering concerns. Rather, both are
engaged in managing constraints over histories. BFO’s philosophical rigor
reflects its commitment to low-entropy regimes; RSVP generalizes this rigor to
domains where such regimes cannot be assumed.

% ==================================================
\section{Artificial Intelligence, Cognition, and Admissibility}
% ==================================================

Debates concerning artificial intelligence often conflate ontological questions
with engineering speculation. The RSVP framework separates these issues by
focusing on admissibility rather than mechanism. The central ontological question
is not whether machines can replicate human cognition, but whether artificial
systems can sustain histories exhibiting the stability, constraint propagation,
and entropy management characteristic of cognitive agency.

Human cognition unfolds within a highly constrained historical regime. Biological
organization, developmental history, and embodied interaction sharply restrict
the space of admissible futures. These constraints yield stable identity,
persistent agency, and coherent semantic reference. From an RSVP perspective,
cognition corresponds to a region of elevated scalar density coupled with richly
structured vector flows and tightly regulated entropy.

Claims that artificial general intelligence is impossible often rest on the
assumption that these constraints are inseparable from biological substrate. RSVP
does not adjudicate substrate dependence directly. Instead, it reframes the issue
in terms of admissibility. If an artificial system can sustain a regime in which
identity, agency, and semantic invariants remain stable across admissible
histories, then ontological parity with cognition follows, regardless of
implementation.

Conversely, fears that artificial systems will spontaneously acquire dangerous
agency often ignore the difficulty of maintaining such regimes. High-capacity
computation alone does not guarantee low entropy. Systems optimized for narrow
tasks frequently operate in high-entropy representational regimes, exhibiting
rapid branching of admissible behaviors and limited identity persistence. RSVP
predicts that such systems will lack robust agency unless explicitly constrained.

This perspective also clarifies the role of training, alignment, and governance.
Training procedures function as entropy-reduction mechanisms, shaping admissible
histories by imposing gradients in the vector field. Alignment constraints
operate by restricting continuation rather than by encoding static values.
Failures of alignment are therefore best understood as failures of constraint
maintenance under historical pressure.

Importantly, RSVP dissolves false analogies between intelligence and tools. Cars
and weapons do not sustain histories with identity or agency; they remain
low-scalar, low-agency structures embedded within human-controlled regimes.
Artificial systems capable of autonomous adaptation occupy a fundamentally
different ontological category, one defined by their capacity to regulate their
own admissible futures.

By grounding analysis in admissibility rather than capability, RSVP provides a
framework for discussing artificial cognition without appeal to hype or
metaphysical speculation. Intelligence becomes a question of sustained
constraint under entropy, not of mimicry or computational scale.

% ==================================================
\section{Conclusion: Ontology as Constraint on History}
% ==================================================

This paper has argued that the persistent difficulties encountered in ontology
engineering—irreversibility, mapping fragility, informational ambiguity, and
regime mismatch—are not merely technical problems but symptoms of a deeper
ontological misalignment. When entities are treated as ontologically primitive,
history, constraint, and degeneration are forced into secondary roles, where
they can be described but not explained. The result is an ontology that performs
well only under narrow, implicitly assumed conditions of stability.

The RSVP framework proposes a reorientation. Rather than beginning with entities
and attempting to account for change, RSVP begins with admissible histories and
treats entities as stabilized invariants within constrained dynamical regimes.
Scalar density captures persistence, vector flow captures directed constraint,
and entropy captures the degeneracy of possible futures. Together, these
primitives articulate the conditions under which identity, relation, agency, and
information arise and persist.

This shift does not abandon realism. On the contrary, it strengthens realist
ontology by grounding ontological categories in the structure of the world
rather than in modeling convenience. Objects, processes, and relations remain
real where they are stable, but their reality is explained rather than assumed.
When stability fails, RSVP predicts ontological breakdown rather than treating it
as modeling error.

The embedding of classical top-level ontologies such as BFO within RSVP
demonstrates that this framework is not oppositional but generative. Entity-
centric ontologies emerge as low-entropy subtheories, valid within domains where
historical branching is limited and scalar density remains high. Their success
and their limits are both explained by the same underlying principles. This
account preserves interoperability without demanding universal uniformity.

Formally, the adjunction between histories and field regimes establishes that
event-historical and field-theoretic descriptions are dual representations of
the same ontological structure. Ontological disagreement is thus reinterpreted
as disagreement over constraint placement, scale of abstraction, or entropy
tolerance, rather than as disagreement over what exists. This reframing dissolves
many long-standing disputes by revealing their structural commonality.

In application to artificial intelligence and cognition, RSVP provides a sober
alternative to both optimism and alarmism. Intelligence is not a substance or a
capability threshold, but a regime of sustained constraint under entropy.
Artificial systems may or may not enter such regimes, depending not on scale
alone but on their capacity to regulate admissible futures. This perspective
reconnects debates about intelligence, alignment, and agency to ontological
conditions rather than speculative narratives.

The broader implication is that ontology engineering must move beyond static
taxonomies toward frameworks capable of representing irreversibility and
historical dependence as first-class features. RSVP offers one such framework.
It does not prescribe a single modeling language, nor does it replace existing
ontologies. Instead, it provides a principled account of when ontological
categories stabilize, when they fail, and why.

Ontology, on this view, is not merely a catalog of what exists. It is a theory of
which histories remain admissible under constraint. To practice ontology
engineering responsibly is therefore to manage not only symbols and categories,
but the conditions under which meaning, identity, and agency can persist through
time.


\newpage
% ==================================================
\appendix
\section{Formal Foundations of Entropic Histories and RSVP Fields}
% ==================================================

This appendix provides a formal mathematical foundation for the ontology
developed in the main text. Its purpose is to render precise the notions of
admissible history, ontological entropy, stabilization, and the scalar--vector
field representation that underwrite the RSVP framework. The presentation is
intentionally domain-neutral. No appeal is made to specific physical laws,
biological mechanisms, or computational architectures. The results concern only
the structure of irreversibility, constraint, and persistence.

\subsection{Histories and Admissibility}

Let $\Omega$ denote the space of all possible event sequences. An element of
$\Omega$ is a totally ordered sequence of events $(e_1 \prec e_2 \prec \cdots)$,
where the ordering relation is primitive and irreversible. No inverse operation
is assumed to exist that would recover prior events from later ones.

A history is defined as a finite or infinite prefix of such a sequence. If $h$
and $h'$ are histories, then $h \preceq h'$ denotes that $h$ is a prefix of $h'$.
This relation induces a partial order on histories corresponding to irreversible
temporal extension.

Ontological coherence is imposed by admissibility constraints. An admissibility
constraint is a predicate $C : \Omega \to \{0,1\}$ that evaluates whether a given
event sequence is permitted under the ontology in question. These constraints
are not epistemic; they do not represent what is known or believed, but rather
what is allowed to occur without violating identity conditions, causal
consistency, semantic integrity, or other ontological requirements.

A history $h$ is said to be admissible if $C(h) = 1$ and if there exists at least
one extension $h' \succeq h$ such that $C(h') = 1$. The second condition ensures
that admissible histories are not terminal dead ends. Ontologies that permit
histories with no admissible continuation are ontologically degenerate in the
sense relevant here.

The set of all admissible histories is denoted by $\Hist \subseteq \Omega$. It
is this space, rather than a set of entities, that serves as the ontological
base of the RSVP framework.

\subsection{Entropy as Degeneracy of Futures}

Given an admissible history $h \in \Hist$, define its admissible future set by
\[
\mathrm{Fut}(h) = \{ h' \in \Hist \mid h \preceq h' \}.
\]
This set represents the total space of ontologically permitted continuations of
$h$.

The entropy of a history $h$ is defined as
\[
\Entropy(h) = \log \lvert \mathrm{Fut}(h) \rvert,
\]
with the understanding that $\Entropy(h)$ may be infinite. This definition
measures the degeneracy of admissible futures rather than uncertainty or lack of
knowledge. A history with low entropy admits few coherent continuations; a
history with high entropy admits many mutually divergent ones.

The role of entropy in ontology is structural. Identity, persistence, and
invariance require that admissible futures remain sufficiently constrained. If
$\Entropy(h)$ is unbounded, then no criterion of persistence can remain invariant
across all admissible extensions of $h$.

\begin{theorem}[Entropy Bound for Identity]
If a structure admits a well-defined identity criterion across all admissible
extensions of a history $h$, then $\Entropy(h)$ must be bounded.
\end{theorem}

\begin{proof}
Assume that $\Entropy(h)$ is unbounded. Then for any proposed identity criterion,
there exist admissible extensions of $h$ that diverge arbitrarily with respect to
that criterion. Since identity must be preserved across all admissible futures,
no such criterion can remain invariant. Therefore bounded entropy is a necessary
condition for identity.
\end{proof}

\subsection{Scalar Density and Stabilization}

Persistence is formalized via scalar density. Let $R$ denote a region or pattern
within a history $h$. The scalar density $\Phi(R,h)$ is defined as the infimum,
over all admissible extensions $h' \in \mathrm{Fut}(h)$, of an invariance measure
$\mathrm{Inv}(R,h')$ that quantifies the degree to which $R$ remains structurally
intact under continuation.

Intuitively, $\Phi(R,h)$ measures how robust a structure is against admissible
historical extension. High scalar density corresponds to strong persistence;
low scalar density corresponds to fragility or transience.

A structure $R$ is said to be stabilized at history $h$ if there exist thresholds
$\Phi_{\min}$ and $\Entropy_{\max}$ such that $\Phi(R,h) \ge \Phi_{\min}$ and
$\Entropy(h|_R) \le \Entropy_{\max}$. Stabilized structures are those that remain
recognizably the same across admissible futures. These structures correspond to
what classical ontologies describe as continuants.

\subsection{Vector Flow and Directional Constraint}

Change and process are captured through directional constraint propagation. Let
$f : h \to h'$ denote an admissible extension of a history. The vector field
$\VecField$ assigns to each event $e$ a directional influence on future
admissibility, formally represented as the sensitivity of the constraint
predicate with respect to that event.

Nonzero $\VecField$ indicates that the occurrence of an event biases or directs
the space of admissible futures. A process corresponds to a coherent trajectory
through history along which $\VecField$ remains consistently oriented. Unlike
scalar density, which measures persistence, vector flow measures transformation
and propagation.

\subsection{The History--Field Correspondence}

The RSVP framework relates histories to field configurations through abstraction.
Given a history $h \in \Hist$, define a mapping
\[
F(h) = (\Phi_h, \VecField_h, \Entropy_h),
\]
where $\Phi_h$ encodes persistence of invariants across admissible futures,
$\VecField_h$ encodes directional constraint propagation, and $\Entropy_h$
encodes the degeneracy of admissible continuations.

Distinct histories may map to the same field configuration when their differences
are irrelevant at the chosen scale. The field representation is therefore a
coarse-graining of historical detail.

Conversely, a field configuration $(\Phi, \VecField, \Entropy)$ determines a
family of admissible histories. Define
\[
G(\Phi, \VecField, \Entropy) = \{ h \in \Omega \mid h \text{ respects all field
bounds} \}.
\]
A history respects a field regime if it remains within scalar stability
thresholds, follows allowed vector directions, and does not exceed entropy
bounds.

\subsection{Adjunction Between Histories and Fields}

The mappings $F$ and $G$ form an adjoint pair. Abstraction from histories to
fields is left adjoint to realization from fields to histories.

\begin{theorem}[History--Field Adjunction]
For any history $h$ and any field regime $r$, the condition that $F(h)$ refines
$r$ is equivalent to the condition that $h \in G(r)$. Consequently, $F \dashv G$.
\end{theorem}

\begin{proof}
If $F(h)$ refines $r$, then the scalar, vector, and entropy characteristics
induced by $h$ satisfy all constraints imposed by $r$, and thus $h$ is an
admissible realization of $r$. Conversely, if $h \in G(r)$, then the field
configuration induced by $h$ cannot violate the constraints of $r$, and hence
$F(h)$ refines $r$. The correspondence is natural and satisfies the unit and
counit conditions of an adjunction.
\end{proof}

This adjunction establishes that event-historical and field-theoretic
descriptions encode the same ontological content. They differ only in whether
constraint is represented dynamically through history or statically through
fields.

\subsection{Ontological Consequences}

The formalism developed here implies that histories, not entities, are
ontologically primary. Entities arise as stabilized equivalence classes of
histories under bounded entropy. Ontological failure corresponds to entropy
divergence, and abstraction corresponds to passage to field-level description.

% ==================================================
\section{A Worked Example: Ontology Versioning and Entropic Divergence}
% ==================================================

This appendix presents a concrete worked example illustrating how the RSVP
framework analyzes ontology evolution and explains the empirical instability of
persistent ontology mappings. The example is intentionally abstracted from any
particular domain ontology, so that the analysis applies equally to biomedical,
institutional, and computational settings.

\subsection{Ontology Versions as Histories}

Let $\mathcal{O}_0$ denote an initial ontology at time $t_0$. We treat $\mathcal
{O}_0$ not as a static artifact, but as the initial prefix of a history in the
space $\Omega$ of possible ontology-development event sequences. Events include
the introduction of new classes, refinement of definitions, addition or removal
of axioms, and changes in modeling conventions.

A versioned ontology $\mathcal{O}_t$ at time $t$ corresponds to a history
\[
h_t = (e_1 \prec e_2 \prec \cdots \prec e_t),
\]
where each event $e_i$ is an ontological modification. Ontology evolution is
therefore represented as irreversible extension in the prefix order on
histories.

Admissibility constraints $C$ encode conditions such as logical consistency,
alignment with intended domain interpretation, and conformance to governance
rules. An ontology version $\mathcal{O}_t$ is admissible if and only if the
corresponding history $h_t$ satisfies $C$ and admits at least one coherent
extension.

\subsection{Admissible Futures and Ontological Entropy}

Given an admissible ontology version $\mathcal{O}_t$, the set of admissible
future versions is given by
\[
\mathrm{Fut}(h_t) = \{ h_{t'} \mid t' \ge t \text{ and } C(h_{t'}) = 1 \}.
\]
This set includes all ways in which the ontology may evolve without violating
its constraints.

The ontological entropy at version $t$ is defined as
\[
\Entropy(h_t) = \log |\mathrm{Fut}(h_t)|.
\]
Low entropy corresponds to a tightly governed ontology with limited permissible
change. High entropy corresponds to an ontology whose future evolution is weakly
constrained, allowing divergent modeling choices, incompatible extensions, or
semantic drift.

\subsection{Scalar Density and Ontological Stability}

Let $R$ be a representational structure within $\mathcal{O}_t$, such as a core
class hierarchy or foundational distinction. The scalar density $\Phi(R,h_t)$
measures the persistence of $R$ across admissible future versions. Formally,
$\Phi(R,h_t)$ is the infimum, over all $h_{t'} \in \mathrm{Fut}(h_t)$, of an
invariance measure that compares the interpretation of $R$ in $\mathcal{O}_t$
and $\mathcal{O}_{t'}$.

When $\Phi(R,h_t)$ is high, the structure $R$ remains stable across versioning.
When $\Phi(R,h_t)$ is low, small changes in the ontology may radically alter the
role or meaning of $R$.

Foundational ontologies such as BFO aim to maximize $\Phi$ for a small set of
core distinctions by enforcing strict admissibility constraints. Rapidly
evolving application ontologies typically tolerate lower scalar density in
exchange for expressive flexibility.

\subsection{Ontology Mappings as Synchronization Operators}

Consider two ontologies $\mathcal{O}^A_t$ and $\mathcal{O}^B_t$ evolving
independently from a common ancestor. A mapping $M_t$ between them is intended
to align corresponding structures at time $t$. Formally, $M_t$ is a relation
between histories $h^A_t$ and $h^B_t$ that preserves admissibility-relevant
structure.

For a mapping to remain valid across time, it must preserve admissibility under
extension. That is, for every admissible extension $h^A_{t'} \succeq h^A_t$,
there must exist a corresponding admissible extension $h^B_{t'} \succeq h^B_t$
such that the mapping relation is preserved.

\subsection{Entropy Amplification and Mapping Failure}

We now show that persistent mappings are generically unstable under entropy
growth.

\begin{theorem}[Mapping Instability Under Entropy Growth]
Let $\mathcal{O}^A_t$ and $\mathcal{O}^B_t$ be independently evolving ontologies
with admissible histories $h^A_t$ and $h^B_t$. If either $\Entropy(h^A_t)$ or
$\Entropy(h^B_t)$ grows without bound, then no mapping $M_t$ can remain valid
across all admissible extensions.
\end{theorem}

\begin{proof}
Assume without loss of generality that $\Entropy(h^A_t)$ is unbounded. Then the
set of admissible futures $\mathrm{Fut}(h^A_t)$ contains arbitrarily many
mutually divergent extensions. Any mapping $M_t$ selects a correspondence
between structures in $\mathcal{O}^A_t$ and $\mathcal{O}^B_t$.

For $M_t$ to remain valid, every admissible extension of $h^A_t$ must correspond
to an admissible extension of $h^B_t$ that preserves the mapped structure. Since
$\mathcal{O}^B_t$ evolves independently, its admissible futures cannot track all
divergent branches of $\mathrm{Fut}(h^A_t)$. Therefore there exists an admissible
extension of $h^A_t$ for which no corresponding extension of $h^B_t$ preserves
$M_t$. Hence $M_t$ fails to remain valid.
\end{proof}

This result explains the empirical fragility of ontology mappings without
appealing to poor engineering practice. Mapping instability is a structural
consequence of entropy divergence.

\subsection{Low-Entropy Regimes and Temporary Synchronization}

Mappings may nevertheless succeed temporarily in low-entropy regimes. If both
$\Entropy(h^A_t)$ and $\Entropy(h^B_t)$ are bounded and if scalar density is high
for the mapped structures, then admissible futures remain sufficiently aligned
to permit synchronization.

In such cases, mappings function as dissipative operators. They reduce local
entropy by enforcing convergence between histories. Once convergence is
achieved, the mapping becomes redundant. This explains why successful mappings
often disappear in mature systems, having been replaced by a unified or shared
ontology.

\subsection{Interpretation}

The worked example demonstrates that ontology versioning is naturally modeled as
irreversible history extension and that mapping instability follows from entropy
growth rather than from conceptual error. RSVP provides a principled explanation
for why foundational ontologies emphasize stability, why application ontologies
exhibit drift, and why mappings are inherently provisional.

This example also clarifies the role of governance. Tight governance corresponds
to entropy suppression and scalar stabilization. Agile development corresponds to
tolerating higher entropy. RSVP does not privilege either strategy; it renders
their tradeoffs explicit.

The example thus shows how abstract RSVP primitives yield concrete explanatory
and predictive insight when applied to real ontology-engineering practice.

% ==================================================
\section{Categorical Structure of Entropic Histories and Field Regimes}
% ==================================================

This appendix provides a categorical reconstruction of the RSVP ontology. The
goal is to show that the history-based and field-based descriptions introduced
in the main text arise as dual presentations of a single underlying structure.
The categorical formulation clarifies compositionality, irreversibility, and
stability, and it makes precise the sense in which classical ontologies appear
as reflective substructures.

\subsection{The Category of Admissible Histories}

Let $\mathbf{Hist}$ denote a category whose objects are admissible histories
$h \in \Hist$. A morphism $f : h \to h'$ exists if and only if $h \preceq h'$
and $h'$ is an admissible extension of $h$. Composition is given by concatenation
of extensions, and identity morphisms correspond to trivial extensions.

By construction, $\mathbf{Hist}$ is not a groupoid. Except for identities, its
morphisms are non-invertible. This categorical asymmetry encodes irreversibility
as a structural feature rather than as an external axiom.

Limits in $\mathbf{Hist}$ correspond to greatest lower bounds of compatible
histories when such bounds exist. Colimits correspond to consistent amalgamation
of histories under shared prefixes, though colimits are generally obstructed by
entropy growth.

\subsection{The Category of RSVP Field Regimes}

Let $\mathbf{Field}$ denote a category whose objects are RSVP field configurations
$(\Phi, \VecField, \Entropy)$ satisfying global admissibility conditions. A
morphism $\alpha : r \to r'$ between field regimes exists when $r'$ is a
coarse-graining or relaxation of $r$ that preserves admissibility. Intuitively,
$\alpha$ forgets fine-grained structure while maintaining stability bounds.

Morphisms in $\mathbf{Field}$ are generally many-to-one. They encode abstraction,
scale change, and representational forgetting. Identity morphisms correspond to
exact preservation of field structure.

\subsection{The Abstraction Functor}

Define a functor $F : \mathbf{Hist} \to \mathbf{Field}$ by assigning to each
history $h$ the minimal field regime $F(h)$ that preserves all admissibility
relations induced by $h$. On morphisms, $F$ maps an extension $h \preceq h'$ to
the induced refinement relation between $F(h)$ and $F(h')$.

The functor $F$ is covariant. It preserves composition and identities by
construction. Conceptually, $F$ performs abstraction: it forgets historical
detail while retaining exactly those constraints required to characterize
admissibility.

\subsection{The Realization Functor}

Define a functor $G : \mathbf{Field} \to \mathbf{Hist}$ by assigning to each
field regime $r$ the collection of histories admissible under the constraints
imposed by $r$. This assignment is functorial when histories are ordered by
prefix extension and field morphisms correspond to relaxation of constraints.

The functor $G$ expands abstract constraint into concrete realization. It is
right-adjoint in character, as it freely generates admissible histories subject
to given bounds.

\subsection{Adjunction and Its Proof}

\begin{theorem}[History--Field Adjunction]
The functors $F : \mathbf{Hist} \to \mathbf{Field}$ and
$G : \mathbf{Field} \to \mathbf{Hist}$ form an adjoint pair, with $F$ left
adjoint to $G$.
\end{theorem}

\begin{proof}
Let $h$ be an object of $\mathbf{Hist}$ and let $r$ be an object of
$\mathbf{Field}$. A morphism $F(h) \to r$ in $\mathbf{Field}$ exists if and only
if the field regime induced by $h$ refines the constraints of $r$. This condition
holds precisely when $h$ is admissible under $r$, which is equivalent to the
existence of a morphism $h \to G(r)$ in $\mathbf{Hist}$.

This correspondence is natural in both $h$ and $r$. The unit of the adjunction
maps each history to its canonical realization under its induced field regime.
The counit maps each field regime to the abstraction of its realized histories.
The triangle identities follow from minimality of abstraction and maximality of
realization.
\end{proof}

The adjunction formalizes the claim that histories and fields encode the same
ontological content at different levels of description. Disputes between
history-first and structure-first ontologies thus reduce to representational
preference rather than metaphysical disagreement.

\subsection{Low-Entropy Subcategories}

Within $\mathbf{Field}$, consider the full subcategory
$\mathbf{Field}_{\mathrm{low}}$ consisting of field regimes whose entropy is
uniformly bounded and whose scalar density exceeds a fixed threshold. Objects of
$\mathbf{Field}_{\mathrm{low}}$ correspond to stable regimes in which identity
conditions are preserved across admissible realizations.

\begin{proposition}
The inclusion functor
$i : \mathbf{Field}_{\mathrm{low}} \hookrightarrow \mathbf{Field}$
admits a left adjoint.
\end{proposition}

\begin{proof}
Given an arbitrary field regime $r$, define $L(r)$ to be the maximal subregime of
$r$ obtained by suppressing all degrees of freedom that contribute to unbounded
entropy or scalar instability. This construction yields a universal arrow from
$r$ to $\mathbf{Field}_{\mathrm{low}}$, establishing reflectivity.
\end{proof}

Reflectivity explains why entity-centric ontologies appear rigid. They operate
entirely within $\mathbf{Field}_{\mathrm{low}}$, implicitly projecting richer
field structure into stabilized subcategories.

\subsection{Ontology Mappings as Spans}

Ontology mappings may be represented categorically as spans in
$\mathbf{Field}$. A mapping between regimes $r_1$ and $r_2$ is a diagram
$r_1 \leftarrow m \rightarrow r_2$, where $m$ is a mediating regime encoding
shared constraints.

\begin{theorem}[Non-existence of Persistent Mediators]
If either $r_1$ or $r_2$ lies outside $\mathbf{Field}_{\mathrm{low}}$, then no
universal mediating object $m$ exists.
\end{theorem}

\begin{proof}
Outside $\mathbf{Field}_{\mathrm{low}}$, entropy is unbounded and admissible
realizations diverge. Any candidate mediator would fail to preserve
admissibility across all realizations, violating universality. Hence no such
object exists.
\end{proof}

This result recovers the mapping instability theorem of Appendix B in categorical
form.

\subsection{Symbolic Ontologies as Presentations}

Symbolic ontologies expressed in OWL or first-order logic correspond to
presentations of objects in $\mathbf{Field}_{\mathrm{low}}$. Their axioms define
generators and relations sufficient to describe stabilized regimes, but they
cannot present arbitrary objects of $\mathbf{Field}$.

This explains categorically why symbolic ontologies succeed in stable domains and
fail in highly dynamic ones. The failure is not logical but ontological: the
objects being modeled lie outside the reflective subcategory.

\subsection{Interpretive Summary}

The categorical reconstruction establishes that irreversibility corresponds to
non-invertibility of morphisms, abstraction corresponds to left adjoints,
stability corresponds to reflectivity, and ontology engineering corresponds to
navigation within and between entropy-bounded subcategories. The RSVP ontology
thus admits a fully compositional semantics without reifying entities as
primitive objects.

% ==================================================
\section{A Worked Biological Example: Cell Lineages as Entropic Histories}
% ==================================================

This appendix develops a worked biological example demonstrating how the RSVP
ontology analyzes persistence, differentiation, and identity in a concrete
setting. The example is intentionally minimal and abstract. No specific molecular
mechanisms are assumed. The analysis concerns only lineage structure,
irreversibility, and admissible continuation.

\subsection{Lineages as Histories}

Consider a single biological cell at an initial developmental time $t_0$. Let
$e_0$ denote the event corresponding to the existence of this cell. As the cell
divides, differentiates, or undergoes apoptosis, each such change is represented
as an irreversible event extending the developmental history.

A cell lineage is therefore represented as a history
\[
h_t = (e_0 \prec e_1 \prec e_2 \prec \cdots \prec e_t),
\]
where each $e_i$ corresponds to a developmental event. The prefix order on
histories captures the fact that later developmental states presuppose earlier
ones and cannot be reversed.

Admissibility constraints encode biological coherence. Examples include viability,
compatibility with developmental programs, and conservation of lineage
continuity. A lineage history is admissible if it satisfies these constraints
and admits further viable continuation.

\subsection{Admissible Futures and Developmental Entropy}

Given an admissible lineage history $h_t$, define the set of admissible future
lineages
\[
\mathrm{Fut}(h_t) = \{ h_{t'} \mid t' \ge t \text{ and } h_{t'} \text{ is viable} \}.
\]

The entropy of the lineage at time $t$ is given by
\[
\Entropy(h_t) = \log |\mathrm{Fut}(h_t)|.
\]

Early in development, entropy is typically high. A pluripotent cell admits many
distinct admissible futures corresponding to divergent differentiation pathways.
As development proceeds, admissibility constraints narrow the space of futures,
and entropy decreases.

\subsection{Scalar Density and Cellular Identity}

Let $R$ denote a structural pattern corresponding to a cell type, such as
epithelial or neuronal identity. Scalar density $\Phi(R,h_t)$ measures the degree
to which this identity persists across admissible future histories.

In early development, $\Phi(R,h_t)$ is low for most specialized cell types,
because admissible futures include differentiation into many alternative types.
As differentiation proceeds, admissibility constraints restrict futures, and
$\Phi(R,h_t)$ increases for the realized cell type.

Cellular identity is therefore not primitive. It emerges when scalar density
exceeds a threshold and entropy falls below a bound. A differentiated cell type
corresponds to a stabilized region of history in which identity is preserved
across admissible continuation.

\subsection{Vector Flow and Developmental Directionality}

Development is not merely constraint narrowing but also directional. Certain
events bias future admissibility strongly toward specific outcomes. This bias is
captured by the vector field $\VecField$, which assigns to each developmental
event a directional influence on future lineage structure.

For example, commitment to a germ layer induces a strong directional constraint
that suppresses entire classes of admissible futures. This manifests as a
nonzero vector flow that channels development along a restricted trajectory.

Processes such as differentiation correspond to coherent trajectories along
which $\VecField$ remains consistently oriented. Dedifferentiation, where it
occurs, corresponds to partial reversal of constraint, but never to full
inversion of history.

\subsection{Identity Preservation and Entropy Bounds}

We now formalize the emergence of cell identity.

\begin{theorem}[Emergence of Cell Identity]
A cell type supports a stable identity across development if and only if the
entropy of its lineage history is bounded below a critical threshold.
\end{theorem}

\begin{proof}
If entropy is unbounded, admissible futures diverge across incompatible cell
types, and no identity criterion can remain invariant. Conversely, if entropy is
bounded and scalar density remains high, admissible futures preserve the same
structural pattern, yielding a stable cell identity.
\end{proof}

This result explains why early embryonic cells lack stable identity, while
differentiated cells possess it, without appealing to intrinsic essences.

\subsection{Continuants and Occurrents Revisited}

In this example, the classical distinction between continuants and occurrents
arises naturally. A cell, once differentiated, behaves as a continuant because
its scalar density is high and its entropy is low across admissible futures.
Developmental processes behave as occurrents because they correspond to directed
vector flow through history.

The distinction is therefore emergent rather than axiomatic. It depends on the
regime of entropy and stability rather than on primitive ontological categories.

\subsection{Ontological Breakdown and Plasticity}

Pathological or experimentally induced plasticity corresponds to entropy
increase. When admissibility constraints are relaxed, such as in induced
pluripotency, the entropy of the lineage increases and scalar density of prior
identity decreases. Identity breakdown is thus an ontological consequence of
constraint relaxation rather than a metaphysical anomaly.

This explains why attempts to impose rigid identity categories on highly plastic
biological systems often fail. The failure is not empirical but ontological: the
systems lie outside low-entropy regimes.

\subsection{Interpretation}

The cell lineage example demonstrates that RSVP primitives are not abstractions
detached from biology. They capture precisely the structural features that
determine when identity, persistence, and process distinctions are coherent.

The example also illustrates the central thesis of the paper: ontological
categories emerge from constrained histories rather than being imposed a priori.
Biological identity is not a primitive fact but a stabilized historical regime.

This conclusion generalizes beyond biology. Any system exhibiting irreversible
development, constraint propagation, and branching futures will exhibit the same
ontological structure. RSVP provides a unified language for describing such
systems across domains.

% ==================================================
\section{Complexity, Decidability, and the Limits of Symbolic Reasoning}
% ==================================================

This appendix analyzes the computational properties of reasoning within the RSVP
ontology. The aim is to clarify which forms of reasoning are decidable, which are
tractable, and which are provably intractable or ill-posed. The analysis explains
why symbolic ontologies succeed in low-entropy regimes and fail systematically
outside them.

\subsection{Reasoning as Constraint Satisfaction over Histories}

Reasoning in the RSVP framework is fundamentally the problem of determining
whether a given partial history admits at least one admissible continuation.
Formally, given a history prefix $h \in \Omega$ and a constraint predicate $C$,
the basic decision problem is whether there exists an extension $h' \succeq h$
such that $C(h') = 1$.

This problem generalizes classical satisfiability. In place of static models,
one reasons over irreversible extensions. In place of truth in a structure, one
reasons over admissibility of futures.

\begin{definition}[Historical Satisfiability]
A history prefix $h$ is historically satisfiable if $h \in \Hist$.
\end{definition}

Determining historical satisfiability is the most primitive reasoning task in
RSVP. All higher-level reasoning tasks reduce to this problem under suitable
encodings.

\subsection{Undecidability in the General Case}

In the general case, historical satisfiability is undecidable.

\begin{theorem}[Undecidability of General Historical Satisfiability]
There exists a class of admissibility constraints $C$ for which the problem of
determining whether $h \in \Hist$ is undecidable.
\end{theorem}

\begin{proof}
Let $C$ encode the halting condition of a Turing machine, where events correspond
to computation steps and admissibility requires non-halting. A history prefix is
admissible if and only if the encoded machine does not halt. Determining whether
a given prefix admits continuation is therefore equivalent to the halting
problem, which is undecidable.
\end{proof}

This result establishes that RSVP is not designed to support complete reasoning
in full generality. Any ontology capable of expressing irreversible computation
inherits this limitation.

\subsection{Entropy Bounds and Decidable Subtheories}

Decidability is recovered under entropy bounds.

\begin{definition}[Entropy-Bounded Regime]
A regime is entropy-bounded if there exists a finite constant $K$ such that
$\Entropy(h) \le K$ for all admissible histories $h$ in the regime.
\end{definition}

Entropy-bounded regimes admit only finitely many admissible futures at each
history prefix.

\begin{theorem}[Decidability under Entropy Bounds]
If an ontology operates entirely within an entropy-bounded regime and all
admissibility constraints are recursively enumerable, then historical
satisfiability is decidable.
\end{theorem}

\begin{proof}
If entropy is bounded, then for any history prefix $h$ the set
$\mathrm{Fut}(h)$ is finite. Since admissibility constraints are recursively
enumerable, one may enumerate all admissible extensions and determine whether
any exist. Termination is guaranteed by finiteness.
\end{proof}

This theorem provides the formal justification for the success of symbolic
ontologies in stable domains. Entropy bounds correspond precisely to the
conditions under which reasoning terminates.

\subsection{First-Order Logic as a Low-Entropy Fragment}

First-order logic presupposes stable identity, fixed relations, and a finite
model-theoretic horizon for reasoning. These presuppositions are satisfied
exactly when entropy is bounded and scalar density remains high.

\begin{theorem}[Soundness of First-Order Reasoning]
First-order reasoning over an ontology is sound if and only if the ontology
corresponds to a low-entropy RSVP regime.
\end{theorem}

\begin{proof}
Soundness requires that predicates denote invariant structures across all
admissible interpretations. This invariance holds precisely when admissible
futures do not branch beyond a bounded set. Unbounded entropy introduces
divergent interpretations that invalidate fixed predicate meaning.
\end{proof}

This result explains why first-order and description-logic ontologies degrade
under version drift, rapid conceptual change, or socio-technical feedback.

\subsection{OWL and Description Logics}

OWL reasoning is decidable by design, but this decidability is achieved by
restricting expressiveness. From the RSVP perspective, these restrictions
correspond to implicit entropy bounds. OWL ontologies describe stabilized
subtheories in which identity and participation relations remain invariant.

When OWL ontologies are applied to domains with high entropy, the result is not
logical inconsistency but ontological mismatch. The domain violates the regime
assumptions under which OWL reasoning is sound.

\subsection{Complexity of Reasoning}

Even within entropy-bounded regimes, reasoning complexity may be high.

\begin{theorem}[Complexity Lower Bound]
There exist entropy-bounded RSVP regimes for which historical satisfiability is
NP-complete.
\end{theorem}

\begin{proof}
Encode Boolean satisfiability as an admissibility constraint over histories with
bounded branching. Each assignment corresponds to an admissible future. Checking
whether any admissible future exists is equivalent to SAT.
\end{proof}

This result aligns RSVP with known complexity bounds in symbolic reasoning. It
also clarifies that RSVP does not simplify reasoning; it explains why reasoning
is difficult.

\subsection{Implications for Ontology Engineering}

The complexity results establish that ontology engineering cannot eliminate
computational hardness. What it can do is control regime conditions. By
maintaining low entropy and high scalar density, ontology engineers ensure that
reasoning remains decidable and tractable.

When such conditions cannot be maintained, symbolic reasoning must be augmented
or replaced by historical, probabilistic, or simulation-based methods. RSVP
provides criteria for determining when such transitions are necessary.

\subsection{Interpretation}

This appendix completes the formal justification for RSVP’s layered view of
ontology. Symbolic logics are not rejected; they are localized. Their success is
explained by entropy suppression, and their failure is explained by entropy
divergence.

Ontology engineering, on this view, is not primarily about choosing the right
logic, but about managing the historical and entropic conditions under which
logic remains meaningful.

% ==================================================
\section{Basic Formal Ontology as a Conservative Low-Entropy Fragment of RSVP}
% ==================================================

This appendix establishes a precise formal relationship between the RSVP ontology
and classical realist top-level ontologies, using Basic Formal Ontology (BFO) as
a representative case. The main result is that BFO may be interpreted as a
conservative fragment of RSVP, valid within entropy-bounded regimes. The proof
clarifies both the strengths and the limitations of entity-centric realism.

\subsection{The Target Fragment of BFO}

We consider a core fragment of BFO sufficient to capture its foundational
commitments. This fragment includes the distinction between independent
continuants and occurrents, the dependence relations between them, and the
participation relation linking continuants to occurrents. We do not consider
domain-specific extensions or higher-order constructs.

Let $\mathbf{BFO}$ denote the theory axiomatizing this fragment in a standard
first-order or description-logic presentation. Models of $\mathbf{BFO}$ consist
of a domain of entities partitioned into continuants and occurrents, together
with relations satisfying the usual axioms of persistence and participation.

\subsection{RSVP Low-Entropy Regimes}

Let $\mathbf{RSVP}_{\mathrm{low}}$ denote the class of RSVP field regimes
$(\Phi,\VecField,\Entropy)$ such that entropy is uniformly bounded and scalar
density exceeds a fixed positive threshold across all admissible histories.
Within such regimes, identity conditions are preserved across admissible
continuations, and admissible futures do not branch uncontrollably.

Histories admissible under $\mathbf{RSVP}_{\mathrm{low}}$ therefore exhibit
persistent structural invariants and laminar constraint propagation. These
conditions correspond exactly to the modeling assumptions tacitly required by
entity-centric ontologies.

\subsection{Interpretation Map}

We define an interpretation map
\[
I : \mathbf{BFO} \to \mathbf{RSVP}_{\mathrm{low}}
\]
that assigns to each BFO model a corresponding RSVP regime.

Independent continuants are interpreted as stabilized regions of history whose
scalar density exceeds the stability threshold and whose entropy remains bounded
under admissible continuation. Occurrents are interpreted as coherent
trajectories of nonzero vector flow through admissible histories. The
continuant–occurrent distinction is thus realized as a regime distinction rather
than a primitive sortal division.

Participation relations are interpreted as stable couplings between scalar-dense
regions and vector trajectories. Dependence relations correspond to constraints
on admissible futures that prohibit continuation in the absence of the relevant
stabilized structure.

\subsection{Soundness of the Interpretation}

\begin{theorem}[Soundness]
For every model $\mathcal{M}$ of $\mathbf{BFO}$, the interpretation $I(\mathcal
{M})$ is a valid model of $\mathbf{RSVP}_{\mathrm{low}}$.
\end{theorem}

\begin{proof}
Let $\mathcal{M}$ be a model of $\mathbf{BFO}$. By the axioms of BFO, continuants
persist across time, occurrents unfold in time, and participation preserves
identity of continuants. These axioms imply that admissible futures of any
history in $\mathcal{M}$ are constrained so as to preserve identity and
participation relations.

Therefore entropy is bounded in the induced RSVP regime, and scalar density of
continuants remains above threshold. Vector flow corresponding to occurrents is
laminar and does not destabilize scalar regions. Hence $I(\mathcal{M})$ satisfies
the defining conditions of $\mathbf{RSVP}_{\mathrm{low}}$.
\end{proof}

\subsection{Completeness up to Conservative Extension}

\begin{theorem}[Conservative Embedding]
Let $\varphi$ be any sentence expressible in the language of $\mathbf{BFO}$. Then
$\mathbf{BFO} \models \varphi$ if and only if $\mathbf{RSVP}_{\mathrm{low}}
\models I(\varphi)$.
\end{theorem}

\begin{proof}
If $\mathbf{BFO} \models \varphi$, then $\varphi$ holds in every BFO model. By
soundness, every such model corresponds under $I$ to an RSVP low-entropy regime
in which $I(\varphi)$ holds. Conversely, if $I(\varphi)$ holds in every
$\mathbf{RSVP}_{\mathrm{low}}$ model, then in particular it holds in all regimes
arising from BFO models. Since $I$ preserves the interpretation of BFO
predicates, $\varphi$ holds in all BFO models. Hence the embedding is
conservative.
\end{proof}

This theorem establishes that RSVP does not invalidate BFO within its domain of
competence. Rather, it strictly generalizes it.

\subsection{Failure Outside Low-Entropy Regimes}

We now show that the embedding cannot be extended beyond low-entropy regimes.

\begin{theorem}[Non-Extendability]
There exists no conservative embedding of $\mathbf{BFO}$ into RSVP regimes with
unbounded entropy.
\end{theorem}

\begin{proof}
In regimes with unbounded entropy, admissible futures branch across mutually
incompatible continuations. Identity conditions for continuants cannot be
preserved across all admissible futures. Since BFO axioms require such
preservation, no interpretation satisfying those axioms can exist in these
regimes. Therefore no conservative embedding is possible.
\end{proof}

This result explains why BFO-style ontologies encounter systematic difficulty
when applied to highly dynamic, versioned, or socio-technical domains. The
difficulty is ontological rather than methodological.

\subsection{Interpretive Consequences}

The embedding theorems establish that BFO is a valid and rigorous ontology within
a specific entropic regime. Its apparent rigidity reflects the stability of the
domains for which it was designed. RSVP does not replace BFO; it situates it.

More generally, the result demonstrates that disagreements between static and
dynamic ontologies are not disputes about realism versus anti-realism. They are
disputes about regime assumptions concerning entropy, stabilization, and
historical admissibility.

RSVP provides a formal language in which these assumptions can be stated,
analyzed, and compared. In doing so, it offers a unifying foundation for ontology
engineering that preserves the achievements of classical realism while extending
its reach.

% ==================================================
\section{A Worked Cognitive Example: Learning as Constraint Stabilization}
% ==================================================

This appendix develops a worked cognitive example illustrating how learning
processes are represented within the RSVP ontology. The purpose is not to model
neural implementation details, but to formalize learning as the progressive
stabilization of admissible histories under constraint. The analysis applies
equally to biological and artificial learning systems.

\subsection{Learning Systems as Historical Processes}

Let a learning system be characterized by a sequence of interaction events with
an environment. Each interaction produces an irreversible update to the system’s
internal state. Let $e_i$ denote the $i$-th update event, and let the learning
history at time $t$ be
\[
h_t = (e_1 \prec e_2 \prec \cdots \prec e_t).
\]

Admissibility constraints encode viability conditions such as coherence of
internal representation, bounded resource consumption, and compatibility with
environmental feedback. A learning history is admissible if it satisfies these
constraints and admits further extension without collapse.

\subsection{Hypothesis Spaces and Entropic Structure}

Let $\mathcal{H}_t$ denote the hypothesis space compatible with the learning
system’s internal state at time $t$. Each hypothesis corresponds to a distinct
way the system may interpret or act upon future inputs. The admissible futures of
the learning history correspond to trajectories through $\mathcal{H}_t$ that
remain compatible with feedback.

The entropy of the learning state at time $t$ is defined as
\[
\Entropy(h_t) = \log |\mathcal{H}_t|.
\]

Early in learning, entropy is high: many hypotheses remain admissible. As learning
progresses, feedback eliminates incompatible hypotheses, reducing entropy. This
reduction is irreversible, since eliminated hypotheses cannot be reinstated
without violating historical consistency.

\subsection{Scalar Density and Learned Representations}

Let $R$ denote a representational invariant, such as a learned category, policy,
or predictive structure. Scalar density $\Phi(R,h_t)$ measures the persistence of
$R$ across admissible futures. Formally, $\Phi(R,h_t)$ is high if all hypotheses in
$\mathcal{H}_t$ realize $R$ up to equivalence.

A representation becomes meaningful precisely when its scalar density exceeds a
threshold. Before this point, the system does not yet possess the representation
in an ontologically robust sense; it merely entertains it among many alternatives.

\subsection{Vector Flow and Learning Direction}

Learning is directional. Feedback does not merely restrict hypothesis space but
biases it. This bias is captured by the vector field $\VecField$, which assigns to
each learning event a directional influence on the evolution of $\mathcal{H}_t$.

Gradient-based learning corresponds to smooth vector flow that steadily reduces
entropy. Abrupt regime changes, such as catastrophic forgetting or phase
transitions in representation, correspond to sharp changes in vector orientation
and magnitude.

\subsection{Stability and Generalization}

Generalization corresponds to stability under admissible continuation. A learned
structure generalizes if it persists across future interactions not yet
encountered. In RSVP terms, generalization requires both bounded entropy and high
scalar density.

\begin{theorem}[Generalization as Entropy Suppression]
A learned representation generalizes if and only if the entropy of admissible
learning futures is bounded below a critical threshold while scalar density
remains high.
\end{theorem}

\begin{proof}
If entropy is unbounded, admissible futures include incompatible hypotheses, and
the representation fails to persist. Conversely, if entropy is bounded and scalar
density remains high, all admissible futures preserve the representation,
yielding generalization.
\end{proof}

\subsection{Overfitting and Underspecification}

Overfitting corresponds to excessive entropy suppression relative to the true
structure of the environment. The learning system collapses admissible futures too
aggressively, producing a brittle scalar regime. Underspecification corresponds
to insufficient entropy reduction, leaving multiple incompatible futures
admissible.

Both phenomena are naturally expressed in RSVP terms as mismanagement of entropy
rather than as failures of representation alone.

\subsection{Identity and Agency}

A learning agent acquires identity when its policy or representational structure
becomes invariant across admissible futures. Agency is not primitive; it emerges
when vector flow becomes coherent and scalar density stabilizes sufficiently to
support counterfactual robustness.

This account explains why agency degrades under distributional shift or
nonstationary environments. Entropy increases, scalar density decreases, and
identity dissolves.

\subsection{Artificial Learning Systems}

The analysis applies directly to artificial learning systems. Training corresponds
to historical constraint accumulation. Model parameters encode scalar density.
Optimization dynamics encode vector flow. Dataset diversity and nonstationarity
contribute to entropy.

This perspective clarifies why scaling alone does not guarantee intelligence.
Without mechanisms that regulate admissible futures, entropy may remain high even
in large systems, preventing stable identity or agency from emerging.

\subsection{Interpretation}

The cognitive example demonstrates that RSVP provides a principled ontology of
learning that subsumes statistical, dynamical, and symbolic perspectives. Learning
is not merely parameter adjustment or belief update; it is the irreversible
shaping of admissible futures.

By grounding cognition in constraint stabilization rather than representational
substance, RSVP offers a unified account of learning, generalization, and agency
that applies across biological and artificial systems.

% ==================================================
\section{Indivisibility, Unistochastic Dynamics, and Emergent Quantum Structure}
% ==================================================

This appendix develops a formal connection between the RSVP ontology and
indivisible stochastic dynamics, showing how quantum-like structure can emerge
without positing wave functions or Hilbert space as ontological primitives. The
analysis builds on the notion that irreversibility and history dependence, when
combined with probabilistic admissibility, force a departure from classical
factorization.

\subsection{Indivisible Histories}

A history $h \in \Hist$ is said to be temporally divisible if it admits a
factorization into conditionally independent subhistories. Formally, $h$ is
divisible if there exists a nontrivial prefix $h_1 \prec h$ such that admissible
continuation probabilities satisfy
\[
P(h \mid h_1) = \prod_i P(h_i \mid h_1),
\]
where the $h_i$ are independent extensions. A history is indivisible when no such
factorization exists.

Indivisibility arises naturally in RSVP when admissibility constraints couple
future events nonlocally across history. Entropy production, once conditioned on
past constraint accumulation, introduces correlations that cannot be decomposed
into independent steps.

\subsection{Stochastic Admissibility}

Let $P(h' \mid h)$ denote the probability that a given admissible extension $h'$
occurs, conditioned on the prefix $h$. These probabilities are not fundamental;
they summarize ignorance about which admissible continuation will be realized.

In divisible regimes, these probabilities factor through intermediate prefixes.
In indivisible regimes, they do not. The law of total probability remains valid,
but temporal factorization fails.

\subsection{Unistochastic Transition Structure}

Consider a finite entropy-bounded regime in which admissible futures form a
finite set $\mathrm{Fut}(h)$. Let $T$ be the matrix of transition probabilities
between coarse-grained states induced by admissible history extensions. In
general, $T$ need not be stochastic with respect to any underlying deterministic
process.

\begin{theorem}[Unistochastic Representation]
If admissible histories are indivisible but entropy-bounded, then the transition
matrix $T$ admits a unistochastic representation.
\end{theorem}

\begin{proof}
Indivisibility implies that probabilities cannot be represented as convex
combinations of deterministic trajectories. However, entropy bounds ensure
finite-dimensionality of admissible futures. By construction, $T$ satisfies the
constraints of doubly stochastic normalization. By the Birkhoff–von Neumann
theorem and its unistochastic refinement, $T$ may be represented as the squared
moduli of a unitary matrix. The phases encode historical interference effects
arising from non-factorizable admissibility.
\end{proof}

This result shows that unitary structure emerges as a representational necessity
when histories are indivisible, not as a postulate about physical reality.

\subsection{Interference as Historical Coupling}

Interference phenomena arise when distinct admissible histories contribute
non-additively to future probabilities. In RSVP terms, this occurs when scalar
density and entropy jointly constrain admissibility so that alternative paths
cannot be treated independently.

The complex phases in the unistochastic representation encode the orientation of
vector flow across history space. Constructive and destructive interference
correspond to alignment or opposition of these flows.

\subsection{Measurement as Entropy Collapse}

Measurement corresponds to a reduction of admissible futures induced by coupling
to a low-entropy apparatus. The apparent collapse of probabilistic structure is
an update of admissibility constraints, not a physical discontinuity.

\begin{theorem}[Measurement Update]
Measurement reduces entropy by restricting admissible histories while preserving
indivisibility of the realized trajectory.
\end{theorem}

\begin{proof}
Coupling to a measurement apparatus imposes additional constraints that eliminate
large subsets of admissible futures. The realized history remains indivisible,
since the constraints act globally on admissibility rather than locally on events.
\end{proof}

\subsection{Classical Limit}

The classical limit corresponds to regimes in which entropy is low and histories
become approximately divisible. In such regimes, unistochastic matrices approach
stochastic ones, and interference terms vanish. Scalar density dominates vector
flow, yielding stable identity and classical behavior.

\subsection{Interpretation}

This appendix demonstrates that quantum-like formal structure arises naturally
within RSVP from two conditions alone: irreversibility of history and bounded but
indivisible admissibility. No appeal to microscopic ontology or wave function
realism is required.

RSVP thus provides a unified explanatory framework in which classical mechanics,
statistical mechanics, and quantum phenomena appear as regime-dependent
manifestations of the same underlying historical ontology.

% ==================================================
\section{References}
% ==================================================

\begin{thebibliography}{99}

\bibitem{SmithBFO}
B.~Smith, W.~Ceusters, B.~Klagges, J.~Köhler, A.~Kumar, J.~Lomax, C.~Mungall,
F.~Neuhaus, A.~Rector, and C.~Rosse.
\newblock Relations in biomedical ontologies.
\newblock \emph{Genome Biology}, 6(5):R46, 2005.

\bibitem{ArpSmithSpears}
R.~Arp, B.~Smith, and A.~D. Spear.
\newblock \emph{Building Ontologies with Basic Formal Ontology}.
\newblock MIT Press, Cambridge, MA, 2015.

\bibitem{GuarinoFormalOntology}
N.~Guarino.
\newblock Formal ontology and information systems.
\newblock In \emph{Proceedings of FOIS 1998}, pages 3--15. IOS Press, 1998.

\bibitem{MasoloDOLCE}
C.~Masolo, S.~Borgo, A.~Gangemi, N.~Guarino, and A.~Oltramari.
\newblock Ontology library.
\newblock \emph{WonderWeb Deliverable D18}, 2003.
\newblock (DOLCE foundational ontology.)

\bibitem{GrenonSmithSNAPSPAN}
P.~Grenon and B.~Smith.
\newblock SNAP and SPAN: Towards dynamic spatial ontology.
\newblock \emph{Spatial Cognition and Computation}, 4(1):69--104, 2004.

\bibitem{BeverleySmithDebate}
J.~Beverley and B.~Smith.
\newblock Debates on artificial intelligence and ontology engineering, Part~2.
\newblock Public debate transcript, 2025.

\bibitem{BarandesIndivisible}
J.~Barandes.
\newblock Indivisible stochastic processes and quantum mechanics.
\newblock \emph{Foundations of Physics}, 52:1--31, 2022.

\bibitem{Jaynes}
E.~T. Jaynes.
\newblock \emph{Probability Theory: The Logic of Science}.
\newblock Cambridge University Press, Cambridge, 2003.

\bibitem{FristonFEP}
K.~Friston.
\newblock The free-energy principle: a unified brain theory?
\newblock \emph{Nature Reviews Neuroscience}, 11(2):127--138, 2010.

\bibitem{Landauer}
R.~Landauer.
\newblock Irreversibility and heat generation in the computing process.
\newblock \emph{IBM Journal of Research and Development}, 5(3):183--191, 1961.

\bibitem{Shannon}
C.~E. Shannon.
\newblock A mathematical theory of communication.
\newblock \emph{Bell System Technical Journal}, 27:379--423, 623--656, 1948.

\bibitem{FloridiInformation}
L.~Floridi.
\newblock \emph{The Philosophy of Information}.
\newblock Oxford University Press, Oxford, 2011.

\bibitem{Hutchins}
E.~Hutchins.
\newblock \emph{Cognition in the Wild}.
\newblock MIT Press, Cambridge, MA, 1995.

\bibitem{RosenLifeItself}
R.~Rosen.
\newblock \emph{Life Itself: A Comprehensive Inquiry into the Nature, Origin,
and Fabrication of Life}.
\newblock Columbia University Press, New York, 1991.

\bibitem{Whitehead}
A.~N. Whitehead.
\newblock \emph{Process and Reality}.
\newblock Free Press, New York, 1929.

\bibitem{SimonSciencesArtificial}
H.~A. Simon.
\newblock \emph{The Sciences of the Artificial}.
\newblock MIT Press, Cambridge, MA, 3rd edition, 1996.

\bibitem{ProppExplodingDots}
J.~Propp.
\newblock Exploding dots and their applications.
\newblock \emph{The Mathematical Intelligencer}, 36(4), 2014.

\end{thebibliography}

\end{document} 