\documentclass[11pt]{article}

% --------------------------------------------------
% Packages
% --------------------------------------------------
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{geometry}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{csquotes}
\usepackage{hyperref}

\geometry{margin=1in}
\setstretch{1.15}

% --------------------------------------------------
% Macros
% --------------------------------------------------
\newcommand{\Hist}{\mathcal{H}}
\newcommand{\PhiField}{\Phi}
\newcommand{\VecField}{\vec{v}}
\newcommand{\Entropy}{S}

% --------------------------------------------------
% Title
% --------------------------------------------------
\title{Toward a Top-Level Ontology of Entropic Histories:\\
A Scalar--Vector--Entropy Foundation for Ontology Engineering}
\author{Flyxion}
\date{\today}

\begin{document}
\maketitle

% ==================================================
\begin{abstract}
% ==================================================

Top-level ontologies such as the Basic Formal Ontology have demonstrated the
importance of realist discipline in ontology engineering. However, their core
primitives remain fundamentally static, treating entities as primary and
processes as derivative. This limits their ability to represent irreversibility,
historical dependence, semantic drift, and field-like phenomena now central to
physics, computation, and cognition.

This paper proposes a new top-level ontology grounded in the
Relativistic Scalar--Vector--Entropy Plenum (RSVP). Rather than taking entities
as primitive, the proposed framework treats entropic histories, constrained
flows, and stabilization regimes as fundamental. Objects, relations, agents,
and information arise as low-entropy invariants within a dynamically constrained
plenum.

The resulting ontology preserves realism while extending it beyond
object-centric metaphysics, offering a unified foundation for physical,
biological, cognitive, and computational domains.
\end{abstract}

% ==================================================
\section{Introduction}
% ==================================================

Ontology engineering has matured to the point where disagreements no longer
concern only modeling conventions, but the adequacy of the ontological
primitives themselves. Persistent difficulties surrounding irreversibility,
interoperability, information, and agency suggest that current top-level
ontologies are operating at the limits of their conceptual frameworks.

This paper advances the thesis that these difficulties are not engineering
failures but ontological ones. The underlying assumption that entities are
ontologically prior to histories, processes, and constraints must be
reconsidered. We propose an alternative top-level ontology in which history,
entropy, and constrained flow are primitive, and entities are derivative.

% ==================================================
\section{Motivation for a New Top-Level Ontology}
% ==================================================

Three persistent problems motivate the present proposal.

\subsection{Irreversibility}

Most top-level ontologies encode time as an ordering relation on entities or
processes. This representation fails to capture the constitutive role of
irreversibility in physical law, biological development, learning, and
computation. History is treated as descriptive rather than generative.

\subsection{Mapping Fragility}

Ontology interoperability relies heavily on mappings that degrade under version
drift. This fragility is not accidental: mappings attempt to synchronize static
structures that evolve under incompatible dynamics. Without a shared dynamical
substrate, mappings become brittle artifacts rather than stabilizing mechanisms.

\subsection{The Ontology of Information}

Information remains an unstable category, oscillating between abstracta, mental
entities, and artifacts. Existing ontologies lack principled criteria for
distinguishing informational structure from its physical or symbolic carriers.

% ==================================================
\section{Design Commitments}
% ==================================================

The proposed ontology adopts the following commitments:

\begin{enumerate}
\item \textbf{Realism without substance fixation.} Reality exists independently
of minds, but not all real structures are object-like.
\item \textbf{Irreversibility as primitive.} Histories are not derived from
entities; entities emerge from stabilized histories.
\item \textbf{Ontological entropy.} Entropy measures degeneracy of admissible
futures, not epistemic uncertainty.
\item \textbf{Anti-conceptualism.} Ontologies are about the world, not about
concepts in minds.
\end{enumerate}

These commitments align the framework with realist ontology while rejecting
static metaphysical atomism.

% ==================================================
\section{Core Ontological Primitives}
% ==================================================

The ontology introduces three irreducible primitives.

\subsection{Scalar Fields (\(\PhiField\))}

Scalar fields represent \emph{ontic density}: the degree to which structure
persists under perturbation. Regions of high \(\PhiField\) correspond to stable
structures such as particles, organisms, or institutions. Low \(\PhiField\)
regions correspond to transient or diffuse regimes.

\subsection{Vector Fields (\(\VecField\))}

Vector fields encode directed constraint propagation, including causal,
inferential, and functional flows. Relations are not primitive; they emerge as
stable couplings of vector flows.

\subsection{Entropy Fields (\(\Entropy\))}

Entropy fields measure the degeneracy of admissible futures. Low entropy
corresponds to invariant structure; high entropy corresponds to branching,
instability, or interpretive ambiguity.

% ==================================================
\section{Derived Ontological Categories}
% ==================================================

Classical ontological categories arise as stabilized configurations:

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Classical Category} & \textbf{RSVP Interpretation} \\
\midrule
Object & Persistent low-entropy \(\PhiField\)-invariant \\
Process & Directed flow in \(\VecField\) \\
Relation & Stable coupling of vector flows \\
Information & Entropy-constrained projectable pattern \\
Agent & Self-maintaining entropy-minimizing subsystem \\
\bottomrule
\end{tabular}
\end{center}

No derived category is primitive; all are historically stabilized.

% ==================================================
\section{Event-Historical Semantics}
% ==================================================

Let \(\Hist\) denote the space of admissible histories. A history is an
irreversible trajectory through the coupled fields
\((\PhiField, \VecField, \Entropy)\). Ontological constraints restrict which
histories are admissible.

Under this semantics:
\begin{itemize}
\item Definitions specify admissibility conditions.
\item Classification partitions history space by entropy regimes.
\item Consistency is replayability under constraint.
\end{itemize}

Ontology engineering becomes the practice of constraining histories to prevent
semantic collapse.

% ==================================================
\section{Ontology Mappings Reinterpreted}
% ==================================================

Mappings are not correspondences between entities but temporary synchronization
operators between field regimes. A successful mapping reduces entropy across
systems and enables convergence. Once convergence is achieved, the mapping
becomes redundant and disappears.

This explains the observed decay of mappings without condemning them in
principle.

% ==================================================
\section{Formal Comparison with Existing Top-Level Ontologies}
% ==================================================

\subsection{Comparison with Basic Formal Ontology}

BFO enforces realist discipline through distinctions such as continuant versus
occurrent. However, these distinctions presuppose entity primacy.

The RSVP ontology differs in that:
\begin{itemize}
\item Entities are emergent, not primitive.
\item Irreversibility is constitutive, not auxiliary.
\item Information is ontologically grounded in entropy, not representation.
\end{itemize}

BFO can be interpreted as a stabilized subtheory within an RSVP regime.

\subsection{Comparison with DOLCE and Process Ontologies}

Process ontologies foreground events but typically lack a principled treatment
of entropy and stabilization. RSVP integrates process with persistence by
treating both as regimes of historical constraint.

\subsection{Relation to Information-Centric Ontologies}

Information-centric ontologies often reintroduce conceptualism implicitly.
RSVP avoids this by grounding information in physically and historically
constrained patterns.

% ==================================================
\section{Implications for AI and Cognition}
% ==================================================

Intelligence is not a substance but a field configuration. Human cognition
corresponds to a uniquely evolved \((\PhiField, \VecField, \Entropy)\) regime,
but not the only possible one. This allows rejection of simplistic AGI hype
without asserting metaphysical impossibility.

% ==================================================
\section{Governance and Ontological Stability}
% ==================================================

Top-level ontologies differ by entropy tolerance. Highly conservative ontologies
favor stability; agile ontologies favor adaptability. RSVP provides a framework
for understanding these tradeoffs as regime differences rather than ideological
conflicts.

Ontology engineering must move beyond static taxonomies toward a framework that
takes irreversibility, history, and entropy seriously. A top-level ontology of
entropic histories preserves realism while extending it to domains where objects
are not fundamental.

Ontology should ask not merely what exists, but which histories remain possible.

% ==================================================
\section{Application to Contemporary Debates in Ontology Engineering}
% ==================================================

The RSVP framework provides a unified lens for interpreting disputes in ontology
engineering by treating them as disagreements over historical constraint
placement and entropy regulation. Rather than framing such debates as clashes
between realism and pragmatism, or philosophy and engineering, RSVP reveals them
as differences in how stability, flow, and admissibility are managed across
scales.

We apply this framework to the Beverley--Smith debate series, which addresses
four recurring fault lines in applied ontology: the role of philosophy, the
status of ontology mappings, the feasibility of artificial general intelligence,
and the governance of top-level ontologies.

\subsection{Event-Historical Analysis: Ontology as Constraints on Admissible Histories}

From an event-historical perspective, ontology engineering governs the space of
irreversible representational trajectories. Let \(\Hist\) denote the space of
admissible histories. Ontological commitments function as authorization and
exclusion operators that determine which histories may be composed, replayed,
and extended.

Smith’s position emphasizes \emph{prior exclusion}. Philosophical realism
supplies ex ante constraints—truth as non-negotiable, universals as independent
of concepts, and strict use--mention discipline—that filter admissible histories
before engineering practice begins. These constraints prevent semantic collapse
by eliminating incoherent trajectories early.

Beverley’s position emphasizes \emph{iterative stabilization}. Constraints are
refined through engineering practice, disambiguation, and interoperability
pressures. Once admissibility conditions stabilize, ontology engineering may
``spin off'' from philosophy as an independent discipline.

Event-historically, both positions agree on composability as the criterion of
success. Their disagreement concerns the \emph{origin} of constraints: upstream
in metaphysics or downstream in practice. This tension resolves naturally when
constraints are understood as historically placeable operators rather than
static axioms.

Ontology mappings illustrate this distinction clearly:

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Position} & \textbf{Interpretation as Historical Operator} \\
\midrule
Smith & Entropy amplification via non-stationary rewrites and version drift \\
Beverley & Temporary synchronization enabling convergence and self-elimination \\
\bottomrule
\end{tabular}
\end{center}

Disputes over AGI reduce to admissibility criteria on histories. Smith requires
biological-historical equivalence for intelligence; Beverley accepts functional
equivalence under shared constraints. Governance questions, such as the pace of
BFO evolution, reflect bottleneck management in the selection of admissible
historical trajectories.

\subsection{RSVP Field-Theoretic Interpretation: Ontology as Entropy-Regulated Dynamics}

The same debate admits a complementary RSVP interpretation. Ontological
commitments correspond to regimes in the coupled fields
\((\PhiField, \VecField, \Entropy)\).

Smith defends high-inertia scalar regimes characterized by dense
\(\PhiField\)-attractors anchored by philosophical boundary conditions. These
caps limit entropy growth and preserve realist invariants. Beverley emphasizes
vector-field laminarization, where engineering practice reduces turbulence until
flows stabilize and explicit philosophical intervention recedes.

Ontology mappings manifest as distinct field operations:

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Position} & \textbf{RSVP Interpretation} \\
\midrule
Smith & Torsion and shear increasing entropy production \\
Beverley & Dissipative couplers enabling entropy reduction and equilibration \\
\bottomrule
\end{tabular}
\end{center}

AGI debates become questions of field topology. Smith views human intelligence as
a uniquely embodied, high-density \(\PhiField\) configuration. Beverley allows
that macroscopic invariants may be realized across distinct substrates, even if
microphysical implementations diverge.

Institutional contrasts follow naturally. BFO exemplifies a low-entropy,
high-inertia regime optimized for stability. Corporate ontologies operate in
higher-entropy regimes that prioritize rapid flow and adaptation. RSVP does not
privilege either regime absolutely; it explains their divergence through
scale-dependent entropy tolerance.

\subsection{Equivalence and Synthesis}

Event-historical and RSVP interpretations are equivalent under appropriate
adjunctions. Admissible histories in \(\Hist\) correspond to trajectories through
the coupled RSVP fields. Authorization and exclusion operators map to entropy
constraints, while disputes over philosophical priority versus engineering
agility reduce to differences in scale at which constraints are enforced.

This case study demonstrates that many persistent disagreements in ontology
engineering arise not from incompatible metaphysical commitments, but from
unacknowledged regime differences in entropy management and historical control.

Ontology engineering must move beyond static taxonomies toward frameworks that
treat irreversibility, history, and entropy as primitive. The RSVP ontology,
grounded in entropic histories and constrained flow, preserves realist discipline
while extending ontology to domains where object primacy is inadequate.

Applied to contemporary debates such as the Beverley--Smith series, RSVP
demonstrates explanatory and unifying power. It reframes disputes over philosophy
and engineering, mapping and unification, intelligence and governance, as
multiscale questions of constraint placement rather than ideological conflict.

Ontology, on this view, asks not only what exists, but which histories—and which
field regimes—remain admissible.

% ==================================================
\section{Formal Adjunction Between Histories and Fields}
% ==================================================

This section makes precise the relationship between the event-historical
semantics introduced earlier and the RSVP field-theoretic formulation. We show
that these two perspectives are not merely compatible but formally adjoint: the
space of admissible histories corresponds to trajectories through the coupled
scalar--vector--entropy fields, and vice versa.

\subsection{Histories as Primitive Objects}

Let \(\Hist\) denote the space of admissible histories. A history
\(h \in \Hist\) is defined as an irreversible, temporally ordered sequence of
events subject to ontological constraints. These constraints restrict which
continuations of a partial history remain admissible.

Formally, \(\Hist\) is not merely a set but a structured space equipped with:
\begin{itemize}
\item A precedence relation encoding irreversibility,
\item A family of admissibility predicates restricting extension,
\item A degeneracy measure quantifying the number of admissible futures.
\end{itemize}

Ontology engineering, under this semantics, specifies constraints on
\(\Hist\) such that histories remain composable, replayable, and non-degenerate.

\subsection{Fields as Dynamical Representations}

In the RSVP formulation, ontological structure is represented by a triplet of
fields \((\Phi, \vec v, S)\), where:
\begin{itemize}
\item \(\Phi\) is a scalar field representing ontic density or stability,
\item \(\vec v\) is a vector field representing directed constraint propagation,
\item \(S\) is an entropy field measuring degeneracy of admissible futures.
\end{itemize}

A field configuration defines, at each spacetime point, local constraints on
continuation and stabilization. Crucially, these fields are not static labels
but dynamical quantities whose evolution governs which structures persist.

\subsection{From Histories to Fields}

Given a history \(h \in \Hist\), we define a mapping
\[
F : \Hist \to (\Phi, \vec v, S)
\]
that assigns to each history a corresponding field configuration. Intuitively:
\begin{itemize}
\item Persistent regularities in \(h\) induce regions of high \(\Phi\),
\item Directed causal or inferential asymmetries induce nonzero \(\vec v\),
\item Branching possibilities in future extensions induce local entropy \(S\).
\end{itemize}

This mapping is coarse-grained: distinct histories may induce the same field
configuration when their differences are irrelevant at the chosen scale.

\subsection{From Fields to Histories}

Conversely, given a field configuration \((\Phi, \vec v, S)\), we define a
mapping
\[
G : (\Phi, \vec v, S) \to \Hist
\]
that yields the class of histories admissible under the local field constraints.
A history \(h\) is admissible under \((\Phi, \vec v, S)\) if and only if its
trajectory respects:
\begin{itemize}
\item Stability thresholds encoded in \(\Phi\),
\item Directional constraints encoded in \(\vec v\),
\item Entropy bounds encoded in \(S\).
\end{itemize}

Thus, fields determine not a single history but a constrained family of possible
histories.

\subsection{Adjunction Structure}

The mappings \(F\) and \(G\) form an adjunction:
\[
F \dashv G
\]
in the sense that field configurations are the most general representations that
preserve admissibility relations among histories, while histories are the most
refined realizations of a given field regime.

Under this adjunction:
\begin{itemize}
\item Event-historical admissibility corresponds to entropy-bounded field flow,
\item Historical irreversibility corresponds to non-conservative dynamics,
\item Ontological stabilization corresponds to attractor formation in \(\Phi\).
\end{itemize}

This adjunction formalizes the equivalence between history-first and field-first
ontological descriptions. They differ not in substance but in representational
emphasis.

\subsection{Ontological Consequences}

The adjunction has three immediate consequences for ontology engineering:
\begin{enumerate}
\item Histories, not entities, are ontologically primary.
\item Field descriptions provide scalable abstractions over historical detail.
\item Ontological disagreement can be analyzed as disagreement over constraint
placement rather than over existence claims.
\end{enumerate}

This completes the formal bridge between event-historical semantics and the RSVP
field ontology.

% ==================================================
\section{Embedding a Fragment of BFO as a Low-Entropy RSVP Subtheory}
% ==================================================

This section demonstrates how a representative fragment of a classical
top-level ontology may be embedded within the RSVP framework as a stabilized,
low-entropy subtheory. The purpose is not to replace existing ontologies, but to
show how their core commitments arise naturally as special regimes within a
history-first, field-theoretic ontology.

\subsection{Target Fragment and Scope}

We consider a minimal but representative fragment consisting of:
\begin{itemize}
\item Independent continuants,
\item Occurrents (processes),
\item The continuant--occurrent distinction,
\item Participation relations between continuants and occurrents.
\end{itemize}

This fragment captures the core realist commitments of entity-centric ontology
without invoking domain-specific extensions.

\subsection{RSVP Interpretation of Continuants}

In RSVP terms, an independent continuant corresponds to a region of
persistently high scalar density \(\Phi\) coupled with low entropy \(\Entropy\)
across admissible histories. Such a region exhibits:
\begin{itemize}
\item Stability under perturbation,
\item Identity preservation across historical extension,
\item Resistance to branching degeneration.
\end{itemize}

Formally, a continuant is identified with a connected region \(R\) of spacetime
such that:
\[
\Phi|_R \ge \Phi_{\min}, \qquad \Entropy|_R \le \Entropy_{\max}
\]
for thresholds \(\Phi_{\min}\) and \(\Entropy_{\max}\) determined by the
ontological regime.

Continuants are thus not primitive substances, but stabilized invariants of the
field dynamics.

\subsection{RSVP Interpretation of Occurrents}

Occurrents correspond to directed flows in the vector field \(\vec v\). A
process is not an entity but a temporally extended trajectory along which
constraints propagate. Occurrents may intersect regions associated with
continuants, but they are ontologically distinct in virtue of their directionality.

The classical distinction between continuants and occurrents arises as a
difference between:
\begin{itemize}
\item Scalar-dominated stability regimes (continuants),
\item Vector-dominated flow regimes (occurrents).
\end{itemize}

This distinction is emergent, not axiomatic.

\subsection{Participation as Field Coupling}

Participation relations are interpreted as stable couplings between scalar
regions and vector flows. A continuant participates in an occurrent when a
high-\(\Phi\) region constrains, or is constrained by, a directed flow in
\(\vec v\) without loss of identity.

Such couplings remain well-defined only in low-entropy regimes. As entropy
increases, participation relations become unstable, explaining empirically
observed failures of rigid participation modeling in highly dynamic domains.

\subsection{Low-Entropy Regime Characterization}

The fragment under consideration embeds cleanly within RSVP provided the
following regime conditions hold:
\begin{enumerate}
\item Entropy remains sufficiently low to prevent excessive branching.
\item Scalar density remains sufficiently high to preserve identity.
\item Vector flows remain laminar rather than turbulent.
\end{enumerate}

These conditions describe precisely the domains in which classical realist
ontologies perform well: relatively stable biological, anatomical, and
institutional structures.

\subsection{Interpretive Consequences}

Under this embedding:
\begin{itemize}
\item The continuant--occurrent distinction is preserved, but explained.
\item Participation is no longer primitive, but derived from coupling stability.
\item Ontological rigidity corresponds to low entropy tolerance.
\end{itemize}

Failures of classical ontology in highly dynamic or information-rich domains are
thus not errors but regime violations.

\subsection{Compatibility Without Reduction}

Importantly, this embedding does not reduce RSVP to an entity ontology, nor does
it dissolve existing ontologies into process metaphysics. Instead, it situates
classical top-level ontologies as stabilized subtheories valid under specific
historical and entropic constraints.

This demonstrates that RSVP extends, rather than opposes, existing realist
ontology frameworks.

% ==================================================
\section{Anticipated Objections and Clarifications}
% ==================================================

Any proposal for a new top-level ontology invites skepticism, particularly when
it departs from established entity-centric frameworks. This section anticipates
and addresses three likely objections: that the approach exhibits ``physics
envy,'' that it is excessively abstract, and that it lacks compatibility with
formal ontology languages such as OWL.

\subsection{``This Is Physics Envy''}

A common objection to field-based ontological frameworks is that they import
mathematical or physical formalisms where they are unnecessary or inappropriate.
The RSVP ontology does not derive its authority from physics, nor does it attempt
to reduce ontology to physical theory.

The use of scalar, vector, and entropy fields is structural rather than
physicalist. These constructs serve as formal devices for representing stability,
directionality, and degeneracy across domains. They are equally applicable to
physical systems, biological processes, institutional structures, and
computational workflows.

Crucially, RSVP does not assume that ontology must mirror fundamental physics.
Instead, it provides a unifying language for describing irreversibility and
constraint across scales. The appeal to field structure reflects the
ontological importance of continuity and flow, not a desire to import physical
authority.

\subsection{``The Framework Is Too Abstract to Be Useful''}

Abstraction is often conflated with impracticality. In ontology engineering,
however, abstraction is unavoidable at the top level. The question is not
whether an ontology is abstract, but whether its abstractions support coherent
specialization.

The RSVP ontology is abstract by design, but its abstractions are operationally
anchored. Scalar density corresponds to stability, vector flow to constraint
propagation, and entropy to branching capacity. These notions map directly onto
engineering concerns such as versioning, interoperability, and maintenance
cost.

Moreover, RSVP explains why existing ontologies succeed in some domains and fail
in others: they presuppose low-entropy regimes. Far from being detached from
practice, the framework clarifies the conditions under which abstraction remains
effective.

\subsection{``This Is Not Compatible with OWL or First-Order Logic''}

Another concern is that RSVP cannot be represented in standard ontology
languages. This objection conflates representability with reducibility.

RSVP is a top-level ontological framework, not a domain ontology or a syntax. It
does not claim that all of its primitives can be captured directly in OWL or
first-order logic. Rather, it provides semantic grounding for fragments that are
representable.

As shown in the previous section, classical ontological categories embed cleanly
as low-entropy RSVP subtheories. These subtheories can be represented in OWL
without difficulty. RSVP itself operates at a meta-level, explaining why such
representations are stable and where they break down.

In this respect, RSVP plays a role analogous to that of foundational semantics
for programming languages: it explains the behavior and limits of concrete
formalisms without being identical to them.

\subsection{Scope and Ambition}

The RSVP ontology does not seek to replace existing top-level ontologies
wholesale, nor does it propose a single modeling language. Its ambition is
diagnostic and unificatory: to provide a principled account of stability,
change, and constraint that can inform ontology engineering across domains.

Disagreement with this ambition is reasonable. However, such disagreement should
be framed in terms of alternative accounts of irreversibility and constraint,
rather than as objections to abstraction per se.

% ==================================================
\section{A Minimal OWL and First-Order Fragment}
% ==================================================

This section demonstrates that the RSVP ontology admits faithful representation
of stabilized subtheories in standard ontology languages, including OWL and
first-order logic, without reducing the framework itself to those formalisms.
The intent is not to encode RSVP in its entirety, but to show how practical
ontology engineering proceeds within RSVP-constrained regimes.

\subsection{Representational Strategy}

RSVP distinguishes between:
\begin{itemize}
\item \emph{Meta-ontological structure}, which governs admissibility,
irreversibility, and entropy, and
\item \emph{Stabilized ontological categories}, which admit symbolic
representation.
\end{itemize}

Only the latter are candidates for direct representation in OWL or first-order
logic. RSVP therefore functions as a semantic foundation that licenses, rather
than replaces, symbolic ontologies.

\subsection{Minimal Vocabulary}

We consider a minimal vocabulary sufficient to represent a low-entropy RSVP
regime:

\begin{itemize}
\item \texttt{Entity}
\item \texttt{Process}
\item \texttt{participatesIn}
\item \texttt{hasStabilityLevel}
\end{itemize}

Here, \texttt{Entity} and \texttt{Process} correspond to stabilized continuant-
and occurrent-like categories derived from RSVP field conditions.

\subsection{OWL Fragment}

The following OWL-style axioms illustrate a representable fragment:

\begin{verbatim}
Class: Entity
Class: Process

ObjectProperty: participatesIn
  Domain: Entity
  Range: Process

DataProperty: hasStabilityLevel
  Domain: Entity
  Range: xsd:float
\end{verbatim}

In this fragment, \texttt{hasStabilityLevel} represents a coarse-grained proxy
for scalar density \(\Phi\). Thresholds on this value distinguish entities that
behave as continuants within a given regime.

Crucially, no claim is made that this exhausts the ontological meaning of
stability. Rather, it captures those aspects relevant to a particular modeling
context.

\subsection{First-Order Interpretation}

The same fragment admits a first-order reading. For example:

\[
\forall x \, (Entity(x) \rightarrow \exists s \, hasStabilityLevel(x,s))
\]

\[
\forall x,y \, (participatesIn(x,y) \rightarrow Entity(x) \land Process(y))
\]

Additional constraints may enforce regime-specific stability conditions, such
as:

\[
Entity(x) \land hasStabilityLevel(x,s) \land s \ge s_{\min} \rightarrow Stable(x)
\]

These axioms characterize a low-entropy subtheory in which identity and
participation relations are well-defined.

\subsection{What Is Not Represented}

Several RSVP primitives are intentionally not represented directly:
\begin{itemize}
\item Global entropy fields \(\Entropy\),
\item Vector-field dynamics \(\vec v\),
\item Admissibility conditions over full history space \(\Hist\).
\end{itemize}

These elements govern when the above symbolic representations remain valid.
Their absence from OWL is not a deficiency but a recognition of the expressive
limits of first-order formalisms.

\subsection{Interpretive Payoff}

This separation clarifies a longstanding confusion in ontology engineering.
Symbolic ontologies fail not because they are incorrect, but because they are
applied outside the regimes for which they are stable.

RSVP provides explicit criteria for such regimes. Within them, OWL and
first-order logic remain appropriate, efficient, and robust. Outside them,
higher-level historical or field-theoretic analysis is required.

\subsection{Practical Implications}

In practice, RSVP suggests a layered workflow:
\begin{enumerate}
\item Use RSVP to characterize domain stability and entropy tolerance.
\item Identify stabilized categories appropriate for symbolic modeling.
\item Encode those categories in OWL or first-order logic.
\item Monitor regime drift as entropy increases or constraints change.
\end{enumerate}

This approach preserves the utility of existing tools while grounding their use
in a principled ontological framework.

% ==================================================
\section{Formal Results}
% ==================================================

This section states and proves a small collection of formal results that
articulate the core claims of the RSVP ontology. The purpose of these theorems
is not to introduce new primitives, but to make explicit the structural
consequences of treating entropic histories and constrained flow as ontologically
fundamental.

Throughout, let \(\Hist\) denote the space of admissible histories and let
\((\Phi, \vec v, \Entropy)\) denote the associated RSVP field configuration.

\subsection{Preliminaries}

\begin{definition}[Admissible History]
A history \(h \in \Hist\) is admissible if and only if it satisfies all imposed
ontological constraints and admits at least one non-degenerate continuation.
\end{definition}

\begin{definition}[Stabilized Structure]
A structure is stabilized if it corresponds to a region of low entropy
\(\Entropy\) and persistent scalar density \(\Phi\) across all admissible
continuations.
\end{definition}

\subsection{History Primacy}

\begin{theorem}[Ontological Primacy of Histories]
\label{thm:history-primacy}
Entities are not ontologically primitive; rather, they arise as invariants over
admissible histories.
\end{theorem}

\begin{proof}
Suppose entities were primitive. Then admissibility of histories would be
defined in terms of entity persistence. However, persistence itself presupposes
criteria of historical continuation and identity across time.

Conversely, if admissibility conditions are defined over histories, then
persistent patterns across admissible continuations induce equivalence classes
of histories. These equivalence classes satisfy the criteria traditionally
ascribed to entities.

Thus, entities depend on histories for their identity conditions, while histories
do not depend on entities. Hence histories are ontologically prior.
\end{proof}

\subsection{Entropy and Ontological Stability}

\begin{theorem}[Low Entropy as a Necessary Condition for Identity]
\label{thm:entropy-identity}
A structure can support identity across time only if the entropy of its
admissible future histories is bounded.
\end{theorem}

\begin{proof}
Identity across time requires that future continuations preserve relevant
structure. If entropy is unbounded, arbitrarily many divergent continuations are
admissible, and no invariant criterion of persistence can be maintained.

Therefore, bounded entropy is a necessary condition for identity preservation.
\end{proof}

\begin{corollary}
Classical continuants correspond to low-entropy regimes of the RSVP fields.
\end{corollary}

\subsection{Continuant--Occurrent Distinction}

\begin{theorem}[Emergence of the Continuant--Occurrent Distinction]
\label{thm:continuant-occurrent}
The continuant--occurrent distinction arises as an emergent property of scalar-
and vector-dominated regimes, rather than as a primitive ontological dichotomy.
\end{theorem}

\begin{proof}
In scalar-dominated regimes, high \(\Phi\) and low \(\Entropy\) ensure persistence
under continuation, yielding continuant-like behavior. In vector-dominated
regimes, directed flow in \(\vec v\) governs change without persistence, yielding
occurrent-like behavior.

Since these regimes are defined by field dynamics rather than axioms, the
distinction is emergent rather than fundamental.
\end{proof}

\subsection{Ontology Mappings}

\begin{theorem}[Instability of Persistent Ontology Mappings]
\label{thm:mapping-instability}
Persistent mappings between independently evolving ontologies are unstable
except in low-entropy regimes.
\end{theorem}

\begin{proof}
Mappings must preserve admissibility relations across histories. If either
ontology evolves such that entropy increases, admissible continuations diverge,
and mapping preservation fails.

Only in regimes where entropy remains low do admissible histories remain
sufficiently constrained for mappings to remain stable. Hence persistent
mappings are unstable in general.
\end{proof}

\subsection{Representability}

\begin{theorem}[Conditional Symbolic Representability]
\label{thm:symbolic}
Symbolic ontologies expressed in OWL or first-order logic are sound only within
entropy-bounded RSVP subtheories.
\end{theorem}

\begin{proof}
Symbolic formalisms presuppose stable identity conditions and fixed relations.
These presuppositions are satisfied precisely when entropy is bounded and scalar
density remains high.

Outside such regimes, admissible histories branch too widely for symbolic
constraints to remain invariant. Therefore symbolic ontologies are conditionally
sound.
\end{proof}

\subsection{Adjunction Consistency}

\begin{theorem}[Adjunction Consistency]
\label{thm:adjunction}
The adjunction between histories and fields preserves ontological admissibility.
\end{theorem}

\begin{proof}
By construction, the mapping from histories to fields preserves admissibility by
encoding stability, directionality, and entropy. The reverse mapping preserves
admissibility by generating only histories consistent with field constraints.

Thus admissibility is invariant under the adjunction.
\end{proof}

\subsection{Interpretive Summary}

Taken together, these results establish that:
\begin{itemize}
\item Ontological structure is grounded in admissible histories,
\item Classical categories arise as stabilized regimes,
\item Entropy bounds determine the validity of symbolic representation,
\item Ontology engineering is fundamentally constraint management over time.
\end{itemize}

These theorems formalize the core claims of the RSVP ontology without appealing
to domain-specific physics or cognitive assumptions.

% ==================================================
\section{A Categorical Restatement}
% ==================================================

This section reformulates the RSVP ontology and its formal results in categorical
terms. The aim is not to replace the event-historical or field-theoretic
presentations, but to show that the framework admits a natural and disciplined
categorical semantics. This restatement clarifies compositional structure,
functorial relationships, and invariance properties underlying ontology
engineering.

\subsection{Categories of Histories and Fields}

We define two categories.

\begin{definition}[Category of Histories]
Let \(\mathbf{Hist}\) be a category whose objects are admissible histories and
whose morphisms are irreversible, constraint-preserving extensions of histories.
Composition corresponds to temporal concatenation, and identity morphisms
correspond to trivial extensions.
\end{definition}

Irreversibility is encoded categorically by the absence of nontrivial
isomorphisms: morphisms in \(\mathbf{Hist}\) are generally non-invertible.

\begin{definition}[Category of Field Regimes]
Let \(\mathbf{Field}\) be a category whose objects are RSVP field configurations
\((\Phi, \vec v, \Entropy)\), and whose morphisms are coarse-graining or
refinement maps that preserve admissibility constraints.
\end{definition}

Morphisms in \(\mathbf{Field}\) correspond to changes of scale, abstraction
level, or modeling resolution.

\subsection{Functorial Correspondence}

We define two functors:
\[
F : \mathbf{Hist} \to \mathbf{Field}, \qquad
G : \mathbf{Field} \to \mathbf{Hist}.
\]

\begin{itemize}
\item The functor \(F\) maps a concrete history to the minimal field regime that
preserves its admissibility properties.
\item The functor \(G\) maps a field regime to the category of histories
admissible under its constraints.
\end{itemize}

These functors capture, respectively, abstraction from historical detail and
realization of abstract constraints.

\subsection{Adjunction}

\begin{theorem}[History--Field Adjunction]
\label{thm:categorical-adjunction}
The functors \(F\) and \(G\) form an adjoint pair:
\[
F \dashv G.
\]
\end{theorem}

\begin{proof}
For any history \(h \in \mathbf{Hist}\) and field regime
\(r \in \mathbf{Field}\), there is a natural correspondence between
constraint-preserving maps from \(F(h)\) to \(r\) and history extensions from
\(h\) to \(G(r)\).

This correspondence is natural in both arguments and satisfies the unit and
counit conditions of an adjunction. Intuitively, \(F\) forgets historical detail
while preserving admissibility, and \(G\) freely generates histories consistent
with a given regime.
\end{proof}

The adjunction formalizes the equivalence between history-first and
field-first descriptions established earlier.

\subsection{Stabilized Subcategories}

Classical ontological categories correspond to reflective subcategories.

\begin{definition}[Low-Entropy Subcategory]
Let \(\mathbf{Field}_{\mathrm{low}}\subset\mathbf{Field}\) be the full
subcategory consisting of field regimes with bounded entropy and persistent
scalar density.
\end{definition}

\begin{proposition}
The category \(\mathbf{Field}_{\mathrm{low}}\) admits a reflective embedding into
\(\mathbf{Field}\).
\end{proposition}

\begin{proof}
Given any field regime, coarse-graining operations that suppress high-entropy
variation yield a universal arrow into \(\mathbf{Field}_{\mathrm{low}}\).
\end{proof}

Entity-centric ontologies correspond to working entirely within
\(\mathbf{Field}_{\mathrm{low}}\). Their apparent rigidity reflects categorical
reflection rather than metaphysical necessity.

\subsection{Ontology Mappings}

Ontology mappings may be understood as spans in \(\mathbf{Field}\):
\[
r_1 \leftarrow r_m \rightarrow r_2,
\]
where \(r_m\) is a mediating field regime.

\begin{proposition}
Persistent ontology mappings exist only when the mediating regime lies in
\(\mathbf{Field}_{\mathrm{low}}\).
\end{proposition}

This restates the instability of mappings categorically: outside low-entropy
subcategories, no universal mediating object exists.

\subsection{Symbolic Ontologies as Presentations}

OWL and first-order ontologies correspond to presentations of objects in
\(\mathbf{Field}_{\mathrm{low}}\).

\begin{proposition}
Symbolic ontologies are sound presentations of objects in
\(\mathbf{Field}_{\mathrm{low}}\), but not of arbitrary objects in
\(\mathbf{Field}\).
\end{proposition}

This explains categorically why symbolic ontologies succeed in stable domains
and fail in highly dynamic ones.

\subsection{Interpretive Significance}

The categorical restatement highlights three structural features of the RSVP
ontology:
\begin{itemize}
\item Irreversibility corresponds to non-invertibility of morphisms.
\item Ontological abstraction corresponds to left adjoints.
\item Stability corresponds to reflectivity into low-entropy subcategories.
\end{itemize}

These features clarify that RSVP is not an alternative modeling language, but a
structural semantics for ontology engineering itself.

% ==================================================
\section{Formal Adjunction Between Histories and Fields}
% ==================================================

This section formalizes the relationship between the event-historical semantics
introduced earlier and the RSVP field-theoretic formulation. The central claim is
that these two perspectives are not merely compatible but stand in a precise
adjoint relationship. Event histories and field regimes provide dual
descriptions of the same ontological structure, differing only in the direction
of abstraction and realization.

We begin with the space of admissible histories, denoted \(\Hist\). A history in
this sense is not a mere sequence of events, but an irreversible trajectory
governed by ontological constraints that restrict which continuations remain
possible. These constraints are constitutive rather than descriptive: they do
not merely record what has occurred, but determine what can occur without
violating coherence conditions such as identity preservation, causal
consistency, or semantic stability.

In parallel, the RSVP framework represents ontological structure through a
triplet of fields \((\Phi, \vec v, \Entropy)\). Scalar density \(\Phi\) captures
the persistence and stability of structure, vector flow \(\vec v\) captures
directed constraint propagation, and entropy \(\Entropy\) measures the
degeneracy of admissible futures. A field configuration therefore specifies,
locally and globally, which historical trajectories are permitted and which are
excluded.

The mapping from histories to fields proceeds by abstraction. Given a concrete
history, one may associate to it the minimal field regime that preserves its
admissibility properties. Persistent regularities in the history induce regions
of elevated scalar density, directional asymmetries induce vector flow, and the
breadth of admissible continuations induces local entropy values. Distinct
histories may map to the same field configuration when their differences are
irrelevant at the chosen scale of description.

Conversely, a field configuration determines not a single history but a family
of admissible histories. The mapping from fields to histories therefore proceeds
by realization. A history is admissible under a given field regime if and only
if it respects the stability thresholds encoded in \(\Phi\), the directional
constraints encoded in \(\vec v\), and the entropy bounds encoded in
\(\Entropy\). In this way, fields function as generators of constrained
historical possibility rather than as static descriptors.

These two mappings form an adjoint pair. Abstraction from histories to fields is
left adjoint to realization from fields to histories. Intuitively, the
field-theoretic description is the most general abstraction that preserves
historical admissibility, while the historical description is the most refined
realization consistent with a given field regime. The unit and counit of this
adjunction express, respectively, that no admissible history is lost under
abstraction and that no inadmissible history is introduced under realization.

The ontological significance of this adjunction is substantial. It implies that
disputes between history-first and structure-first ontologies are not disputes
over what exists, but over representational emphasis. Both descriptions encode
the same admissibility structure, differing only in whether constraint is
expressed dynamically through history or statically through fields.

From the perspective of ontology engineering, this result justifies the use of
field-theoretic abstractions without abandoning historical grounding, and
conversely legitimizes event-historical semantics without collapsing into
unstructured narrative. The adjunction provides a formal bridge that preserves
realism while enabling scalable abstraction.

% ==================================================
\section{Embedding a Fragment of BFO as a Low-Entropy RSVP Subtheory}
% ==================================================

This section demonstrates how a representative fragment of a classical
entity-centric top-level ontology may be embedded within the RSVP framework as a
stabilized, low-entropy subtheory. The goal is not to reduce existing ontologies
to RSVP, but to show that their central distinctions arise naturally under
specific historical and entropic conditions. In this sense, RSVP functions as a
general ontological substrate within which traditional realist ontologies appear
as special cases.

We focus on a minimal fragment sufficient to capture the core commitments of
entity-based realism, namely the distinction between continuants and occurrents
and the relation of participation between them. These distinctions have proven
robust and useful in relatively stable domains such as anatomy, material
artifacts, and institutional structures, and thus provide an appropriate test
case for embedding.

Within the RSVP ontology, a continuant corresponds to a region of historically
persistent structure characterized by elevated scalar density \(\Phi\) and
bounded entropy \(\Entropy\). Such a region maintains its identity across
admissible continuations of history, resisting fragmentation under perturbation.
Identity, on this account, is not imposed axiomatically but emerges from the
stability of admissible futures. When entropy remains sufficiently low, the set
of admissible continuations preserves a coherent criterion of sameness, yielding
continuant-like behavior.

Occurrents, by contrast, correspond to regimes dominated by directed vector flow
\(\vec v\). Rather than persisting as stable invariants, they manifest as
temporally extended trajectories along which constraints propagate. Processes
may intersect regions associated with continuants, but their ontological
character is determined by directionality and transformation rather than by
persistence. The classical continuant–occurrent distinction thus emerges as a
contrast between scalar-dominated and vector-dominated regimes within the RSVP
fields.

Participation relations arise when these regimes couple in a stable manner. A
continuant participates in an occurrent when a region of high scalar density
constrains, or is constrained by, a directed flow without loss of identity. Such
couplings remain well-defined only so long as entropy remains bounded. As
entropy increases, admissible histories branch more widely, and participation
relations become unstable. This provides a principled explanation for the
difficulty of modeling participation rigidly in highly dynamic or information-
rich domains.

The fragment under consideration embeds cleanly within RSVP precisely because it
presupposes a low-entropy regime. Scalar density remains high, vector flows are
largely laminar, and admissible histories do not branch excessively. These are
exactly the conditions under which classical top-level ontologies perform well.
Their apparent rigidity is therefore not a metaphysical necessity, but a
reflection of the stability of the domains to which they are applied.

This embedding clarifies both the strength and the limits of entity-centric
ontologies. Their success in stable domains is explained by the existence of
low-entropy RSVP regimes that support persistent identity and well-defined
relations. Their failure in rapidly evolving, computational, or socio-technical
domains is explained by regime violation rather than conceptual error. RSVP does
not invalidate such ontologies; it situates them within a broader ontological
landscape.

Importantly, this account preserves compatibility without reduction. RSVP does
not collapse continuants into processes, nor does it dissolve entity ontology
into flux. Instead, it explains why and when entity-based distinctions arise,
and under what conditions they cease to be adequate. In doing so, it extends
realist ontology beyond static domains while retaining its core commitments.

% ==================================================
\section{Anticipated Objections and Clarifications}
% ==================================================

Any proposal for a new top-level ontology must contend not only with technical
questions but also with institutional and philosophical skepticism. The RSVP
framework is no exception. This section anticipates several likely objections
and clarifies the scope, intent, and methodological commitments of the proposed
ontology. The aim is not to preempt disagreement, but to ensure that such
disagreement is directed at substantive claims rather than misunderstandings.

A first objection is that the framework exhibits a form of ``physics envy,'' in
which ontological questions are subordinated to mathematical or physical
formalisms. This concern is understandable given the use of field-theoretic
language. However, RSVP does not derive its authority from physics, nor does it
attempt to reduce ontology to physical theory. The scalar, vector, and entropy
fields employed here are structural abstractions, introduced to capture
persistence, directed constraint propagation, and degeneracy of admissible
futures across domains. They are not posited as physical substances, but as
formal devices for expressing ontological regularities that recur in biological,
institutional, computational, and social systems as well as in physical ones.

A second objection is that the framework is excessively abstract and therefore
unlikely to be useful in practice. This objection conflates abstraction with
detachment from application. Top-level ontologies are necessarily abstract, as
their role is to articulate the conditions under which more specific modeling
choices become coherent. RSVP embraces abstraction precisely in order to explain
why certain ontological categories stabilize and why others fail under domain
pressure. Concepts such as scalar density and entropy are not introduced for
their own sake, but to make explicit the conditions under which identity,
participation, and persistence remain well-defined.

From this perspective, RSVP is not an alternative to applied ontology
engineering, but a diagnostic framework for it. It explains why classical
ontologies succeed in relatively stable domains and why they encounter
difficulties in rapidly evolving or information-rich contexts. Far from being
removed from practice, the framework provides criteria for determining when
particular modeling strategies are likely to be effective.

A third objection concerns compatibility with standard ontology languages such
as OWL and with first-order logic more generally. RSVP does not claim that all of
its primitives can be directly represented within these formalisms. Such a claim
would be incoherent, as OWL and first-order logic presuppose stable identity and
fixed relations. Instead, RSVP distinguishes between meta-ontological structure
and stabilized ontological categories. Only the latter are candidates for direct
symbolic representation. RSVP therefore operates at a level that explains the
soundness and limits of symbolic ontologies without being reducible to them.

This distinction parallels familiar relationships in other domains. Foundational
semantics in programming languages, for example, are not themselves executable
programs, but they explain why certain programs behave as they do and under what
conditions abstractions remain valid. RSVP plays an analogous role for ontology
engineering.

Finally, it is important to clarify the ambition of the framework. RSVP does not
seek to replace existing top-level ontologies wholesale, nor does it propose a
single universal modeling language. Its goal is instead to provide a principled
account of irreversibility, constraint, and historical admissibility that can
inform ontology engineering across domains. Disagreement with this ambition is
entirely reasonable. However, such disagreement should be framed in terms of
alternative accounts of history, stability, and constraint, rather than as a
rejection of abstraction or formalism as such.

% ==================================================
\section{A Minimal OWL and First-Order Fragment}
% ==================================================

This section demonstrates that the RSVP ontology supports practical
representation within standard ontology languages such as OWL and first-order
logic, provided that representation is restricted to stabilized regimes. The
purpose is not to encode RSVP in its entirety, but to clarify how symbolic
ontologies operate as regime-bound projections of a richer ontological
structure.

RSVP draws a principled distinction between meta-ontological structure and
object-level categories. Meta-ontological structure governs admissibility,
irreversibility, and entropy across histories. Object-level categories arise
only when this structure stabilizes sufficiently to support persistent identity
and fixed relations. Only the latter are candidates for direct symbolic
representation.

Within a low-entropy regime, continuant-like structures exhibit stable identity
conditions across admissible histories. These structures may be represented as
entities in OWL or as unary predicates in first-order logic. Similarly,
occurrent-like structures correspond to stabilized patterns of directed change
and may be represented as processes or events. Participation relations become
representable precisely when coupling between these structures remains stable
under continuation.

A minimal symbolic vocabulary therefore suffices to capture such regimes. One
may introduce predicates or classes corresponding to entities and processes,
together with relations expressing participation. Additional attributes may be
used to encode coarse-grained proxies for stability, such as numerical or
ordinal measures reflecting scalar density. These proxies do not exhaust the
ontological meaning of stability, but they are sufficient for reasoning within a
restricted regime.

From a first-order perspective, such representations presuppose that identity
conditions are invariant across admissible continuations. This presupposition is
not guaranteed by logic alone; it is licensed by the underlying RSVP regime. When
entropy remains bounded and admissible futures do not branch excessively,
identity-preserving predicates remain sound. When these conditions fail,
symbolic representations degrade, not because of logical inconsistency, but
because the ontological assumptions required for their interpretation are no
longer satisfied.

It is therefore a category mistake to criticize RSVP for not being fully
expressible in OWL. OWL is a representational tool designed for stable domains.
RSVP explains why such domains exist and how their stability may erode. The
relationship is thus one of semantic grounding rather than syntactic inclusion.

This perspective clarifies a recurrent pattern in ontology engineering. Symbolic
ontologies tend to perform well in domains characterized by slow change and
strong institutional or physical constraints, such as anatomy or material
artifacts. They encounter difficulty in domains characterized by rapid evolution,
versioning, or socio-technical feedback. RSVP accounts for this pattern by
explicitly modeling entropy and historical branching, thereby providing criteria
for when symbolic modeling is appropriate.

In practical terms, RSVP supports a layered modeling strategy. One first
characterizes the entropic and historical properties of a domain. One then
identifies stabilized structures within that domain and represents only those
structures symbolically. As domain conditions change, the symbolic layer may be
revised or abandoned without treating its failure as an ontological error.

In this way, RSVP preserves the utility of OWL and first-order logic while
placing their use on principled ontological footing.



% ==================================================
\begin{thebibliography}{99}
% ==================================================

\bibitem{SmithBFO}
B.~Smith, W.~Ceusters, B.~Klagges, J.~Köhler, A.~Kumar, J.~Lomax, C.~Mungall,
F.~Neuhaus, A.~Rector, and C.~Rosse.
\newblock Relations in biomedical ontologies.
\newblock \emph{Genome Biology}, 6(5):R46, 2005.

\bibitem{ArpSmithSpears}
R.~Arp, B.~Smith, and A.~D. Spear.
\newblock \emph{Building Ontologies with Basic Formal Ontology}.
\newblock MIT Press, Cambridge, MA, 2015.

\bibitem{GuarinoFormalOntology}
N.~Guarino.
\newblock Formal ontology and information systems.
\newblock In \emph{Proceedings of FOIS 1998}, pages 3--15. IOS Press, 1998.

\bibitem{MasoloDOLCE}
C.~Masolo, S.~Borgo, A.~Gangemi, N.~Guarino, and A.~Oltramari.
\newblock Ontology library.
\newblock \emph{WonderWeb Deliverable D18}, 2003.
(DOLCE foundational ontology.)

\bibitem{GrenonSmithSNAPSPAN}
P.~Grenon and B.~Smith.
\newblock SNAP and SPAN: Towards dynamic spatial ontology.
\newblock \emph{Spatial Cognition and Computation}, 4(1):69--104, 2004.

\bibitem{BeverleySmithDebate}
J.~Beverley and B.~Smith.
\newblock Debates on artificial intelligence and ontology engineering, Part~2.
\newblock Public debate transcript, 2025.

\bibitem{BarandesIndivisible}
J.~Barandes.
\newblock Indivisible stochastic processes and quantum mechanics.
\newblock \emph{Foundations of Physics}, 52:1--31, 2022.

\bibitem{Jaynes}
E.~T. Jaynes.
\newblock \emph{Probability Theory: The Logic of Science}.
\newblock Cambridge University Press, 2003.

\bibitem{FristonFEP}
K.~Friston.
\newblock The free-energy principle: a unified brain theory?
\newblock \emph{Nature Reviews Neuroscience}, 11(2):127--138, 2010.

\bibitem{Landauer}
R.~Landauer.
\newblock Irreversibility and heat generation in the computing process.
\newblock \emph{IBM Journal of Research and Development}, 5(3):183--191, 1961.

\bibitem{Shannon}
C.~E. Shannon.
\newblock A mathematical theory of communication.
\newblock \emph{Bell System Technical Journal}, 27:379--423, 623--656, 1948.

\bibitem{FloridiInformation}
L.~Floridi.
\newblock \emph{The Philosophy of Information}.
\newblock Oxford University Press, 2011.

\bibitem{Hutchins}
E.~Hutchins.
\newblock \emph{Cognition in the Wild}.
\newblock MIT Press, Cambridge, MA, 1995.

\bibitem{RosenLifeItself}
R.~Rosen.
\newblock \emph{Life Itself: A Comprehensive Inquiry into the Nature, Origin,
and Fabrication of Life}.
\newblock Columbia University Press, 1991.

\bibitem{Whitehead}
A.~N. Whitehead.
\newblock \emph{Process and Reality}.
\newblock Free Press, New York, 1929.

\bibitem{SimonSciencesArtificial}
H.~A. Simon.
\newblock \emph{The Sciences of the Artificial}.
\newblock MIT Press, 3rd edition, 1996.

\bibitem{ProppExplodingDots}
J.~Propp.
\newblock Exploding dots and their applications.
\newblock \emph{Mathematical Intelligencer}, 2014.

% ==================================================
\end{thebibliography}


\end{document}
