% orientation-monograph.tex — Orientation: Three Essays
% pdfLaTeX with mathpazo
% Author: Flyxion — Independent Researcher

\documentclass[12pt, a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{mathpazo}
\usepackage{microtype}
\usepackage{geometry}
\geometry{a4paper, top=3cm, bottom=3cm, left=3.2cm, right=3.2cm}
\usepackage{setspace}
\onehalfspacing
\usepackage{titlesec}
\titleformat{\part}[display]{\centering\Large\bfseries}{}{0em}{}[\vspace{0.5em}\rule{0.3\textwidth}{0.4pt}]
\titleformat{\section}{\large\bfseries}{\thesection.}{0.8em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection.}{0.6em}{}
\usepackage[parfill]{parskip}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.7em}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}
\setlist[itemize]{noitemsep, topsep=0pt}
\setlist[enumerate]{noitemsep, topsep=0pt}

% Theorem environments
\newtheoremstyle{essaystyle}
  {0.8em}{0.8em}{\itshape}{0pt}{\bfseries}{.}{0.5em}{}
\theoremstyle{essaystyle}
\newtheorem{principle}{Principle}[part]
\newtheorem{observation}{Observation}[part]
\newtheorem{defn}{Definition}[part]
\newtheorem{proposition}{Proposition}[part]

% Reset section numbering per part
\makeatletter
\@addtoreset{section}{part}
\makeatother

% Part page style — no running number prefix on sections
\renewcommand{\thepart}{\Roman{part}}
\renewcommand{\thesection}{\arabic{section}}

% -----------------------------------------------------------
\begin{document}

% ============================================================
% TITLE PAGE
% ============================================================
\begin{titlepage}
  \vspace*{3.5cm}
  \begin{center}
    {\LARGE\bfseries Orientation}\\[1em]
    {\large\itshape Three Essays on the Scarcity of Navigation}\\[0.6em]
    {\normalsize in Minds, Institutions, and Knowledge Traditions}\\[3em]
    {\normalsize Flyxion}\\[0.4em]
    {\small Independent Researcher}\\[3em]
    {\small June 2026}
  \end{center}
\end{titlepage}

% ============================================================
% TABLE OF CONTENTS
% ============================================================
\newpage
\tableofcontents

% ============================================================
% PREFACE
% ============================================================
\newpage
\phantomsection
\addcontentsline{toc}{part}{Preface}
\begin{center}
  {\Large\bfseries Preface}\\[0.5em]
  \rule{0.3\textwidth}{0.4pt}
\end{center}
\vspace{1em}

\input{body_preface}

% ============================================================
% PART I — ORIENTATION
% ============================================================
\newpage
\part{Orientation}
\begin{center}
  \large\itshape On the Scarcity of Navigation in an Age of Abundant Search
\end{center}

\bigskip

\begin{abstract}
\noindent
For most of the history of knowledge work, exploration and generation were tightly
coupled: to investigate a possibility one had to implement it, and implementation
was costly. Recent developments in automated generation have broken this coupling.
The cost of producing candidate solutions has collapsed dramatically, expanding
the range of possibilities that a single agent can examine in a given interval
of time. This essay argues that the primary consequence of this change is not
an increase in productive capacity but a shift in which resource is limiting.
When generation was expensive, the bottleneck was search. When generation becomes
abundant, the bottleneck shifts to orientation --- the capacity to maintain a
coherent objective through time, to distinguish progress from drift, and to
determine when exploration should give way to commitment. Orientation does not
scale with search. The mechanisms that preserve goal continuity, support
iterative repair, and sustain the distinction between exploration and formalization
become more valuable rather than less as search capacity increases. The central
problem of an age of abundant search is no longer how to generate possibilities
but how to remain oriented among them.
\end{abstract}

\input{body_orientation}

% ============================================================
% PART II — ORIENTATION AND INSTITUTIONS
% ============================================================
\newpage
\part{Orientation and Institutions}
\begin{center}
  \large\itshape How Organizations Preserve Goals Across Time
\end{center}

\bigskip

\begin{abstract}
\noindent
Institutions routinely outlive the reasons for their own rules. The familiar
explanations for this phenomenon locate the failure in agents: in capture,
incompetence, bad incentives, or the self-interest of those who administer
the rules. This essay argues that those explanations are incomplete because
they treat institutional drift as a contingent failure rather than a structural
tendency. The structural explanation is that constraints persist more readily
than the reasons that justified them. When reasons decay while constraints
remain, the institution retains its formal shape while losing its orientation.
This essay calls that process institutional projection collapse. The central
claim is a Reason Retention Principle: an institution remains oriented only
to the extent that it can reconstruct the arguments that originally justified
its constraints. Reconstruction, not mere archival preservation, is the
operative requirement. A second structural claim follows: projection collapse
becomes increasingly difficult to detect as the institution loses access to
the reasons that generated its present constraints. Sufficiently advanced drift
destroys the mechanisms required to recognize drift. These two claims together
reframe a large class of institutional failures as orientation failures ---
problems of continuity rather than problems of incentive.
\end{abstract}

\input{body_institutions}

% ============================================================
% PART III — ORIENTATION AND KNOWLEDGE
% ============================================================
\newpage
\part{Orientation and Knowledge}
\begin{center}
  \large\itshape On the Fragility of Understanding at Civilizational Scale
\end{center}

\bigskip

\begin{abstract}
\noindent
The previous two essays established that production outpaces orientation
at the scale of individual minds and at the scale of institutions. This essay
extends the argument to the scale of knowledge traditions. Knowledge accumulates
monotonically: results are added to the literature and rarely removed.
Understanding does not accumulate in the same way. Understanding is the
capacity to construct coherent structure across knowledge --- to recognize
which results bear on which questions, how findings from different domains
constrain each other, and what a new result means in the context of what is
already known. That capacity cannot be stored or transmitted by copying; it
must be reconstructed by each new understander, at increasing cost as the
body of knowledge grows. When knowledge production exceeds the rate at which
coherent structure is constructed, understanding debt accumulates. Understanding
debt is not merely a backlog of unperformed synthesis. It compounds: subsequent
knowledge is produced in the absence of integrations that would have shaped
it, deepening the fragmentation it must eventually overcome. The essay
introduces the Trajectory Preservation Principle, which unifies the three
debts of the series --- navigability debt, reason decay, and understanding
debt --- as distinct manifestations of a single process: the accumulation of
states whose generating trajectories can no longer be reconstructed. The
literature stores survivors. Understanding stores the selection history.
When the selection history disappears, the survivors lose the context that
made them significant.
\end{abstract}

\input{body_knowledge}

% ============================================================
% EPILOGUE
% ============================================================
\newpage
\phantomsection
\addcontentsline{toc}{part}{Epilogue: Orientation as Reconstruction}
\begin{center}
  {\Large\bfseries Epilogue: Orientation as Reconstruction}\\[0.5em]
  \rule{0.3\textwidth}{0.4pt}
\end{center}
\vspace{1em}

\input{body_epilogue}

% ============================================================
% UNIFIED BIBLIOGRAPHY
% ============================================================
\newpage
\phantomsection
\addcontentsline{toc}{section}{References}
\begin{thebibliography}{99}

% --- Primary scientific anchor ---
\bibitem{fireants2026}
R.~Jena, P.~Chaudhari, and J.~C. Gee.
\newblock Adaptive {R}iemannian optimization for multi-scale diffeomorphic matching.
\newblock \textit{Nature Communications}, 17:4774, 2026.
\newblock \url{https://doi.org/10.1038/s41467-026-72508-3}

% --- Flyxion framework monographs ---
\bibitem{rsvp2024}
Flyxion.
\newblock Axioms for a falling universe: {F}ield equations of the {RSVP} framework.
\newblock \textit{Independent Monograph}, 2024.

\bibitem{clio2025}
Flyxion.
\newblock {CLIO}: Constraint, locality, and irreversible ontology.
\newblock \textit{Independent Monograph}, 2025.

\bibitem{hydra2025}
Flyxion.
\newblock {HYDRA}: Hierarchical dynamics and reachability architecture.
\newblock \textit{Independent Monograph}, 2025.

\bibitem{geometry2025}
Flyxion.
\newblock The geometry of closure: {S}ix-stage ontological enlargement.
\newblock \textit{Independent Monograph}, 2025.

% --- MEM|8 (Chenoweth et al.) ---
\bibitem{mem8_chenoweth2025}
C.~Chenoweth and A.~Chenoweth.
\newblock {MEM}$|$8: A wave-based cognitive architecture for multimodal memory
  integration and consciousness simulation.
\newblock \textit{Zenodo}, 2025.
\newblock \url{https://doi.org/10.5281/zenodo.16436298}

% --- Part I: bounded rationality, exploration/exploitation, path dependence ---
\bibitem{simon1955}
H.~A. Simon.
\newblock A behavioral model of rational choice.
\newblock \textit{The Quarterly Journal of Economics}, 69(1):99--118, 1955.

\bibitem{simon1969}
H.~A. Simon.
\newblock \textit{The Sciences of the Artificial}.
\newblock MIT Press, Cambridge, MA, 1969.

\bibitem{march1991}
J.~G. March.
\newblock Exploration and exploitation in organizational learning.
\newblock \textit{Organization Science}, 2(1):71--87, 1991.

\bibitem{kahneman2011}
D.~Kahneman.
\newblock \textit{Thinking, Fast and Slow}.
\newblock Farrar, Straus and Giroux, New York, 2011.

\bibitem{gigerenzer1999}
G.~Gigerenzer, P.~M. Todd, and the {ABC} Research Group.
\newblock \textit{Simple Heuristics That Make Us Smart}.
\newblock Oxford University Press, New York, 1999.

\bibitem{arthur1994}
W.~B. Arthur.
\newblock \textit{Increasing Returns and Path Dependence in the Economy}.
\newblock University of Michigan Press, Ann Arbor, 1994.

% --- Part II: institutions, tacit knowledge, organizational memory ---
\bibitem{polanyi1966}
M.~Polanyi.
\newblock \textit{The Tacit Dimension}.
\newblock Doubleday, Garden City, NY, 1966.

\bibitem{north1990}
D.~C. North.
\newblock \textit{Institutions, Institutional Change and Economic Performance}.
\newblock Cambridge University Press, Cambridge, 1990.

\bibitem{ostrom1990}
E.~Ostrom.
\newblock \textit{Governing the Commons: The Evolution of Institutions for
  Collective Action}.
\newblock Cambridge University Press, Cambridge, 1990.

\bibitem{march1988}
B.~Levitt and J.~G. March.
\newblock Organizational learning.
\newblock \textit{Annual Review of Sociology}, 14:319--340, 1988.

\bibitem{march1989}
J.~G. March and J.~P. Olsen.
\newblock \textit{Rediscovering Institutions: The Organizational Basis of Politics}.
\newblock Free Press, New York, 1989.

% --- Part III: knowledge growth, paradigms, interdisciplinary fragmentation ---
\bibitem{kuhn1962}
T.~S. Kuhn.
\newblock \textit{The Structure of Scientific Revolutions}.
\newblock University of Chicago Press, Chicago, 1962.

\bibitem{price1963}
D.~J. de~Solla Price.
\newblock \textit{Little Science, Big Science}.
\newblock Columbia University Press, New York, 1963.

\bibitem{klein1990}
J.~T. Klein.
\newblock \textit{Interdisciplinarity: History, Theory, and Practice}.
\newblock Wayne State University Press, Detroit, 1990.

\bibitem{ryle1949}
G.~Ryle.
\newblock \textit{The Concept of Mind}.
\newblock Hutchinson, London, 1949.

% --- Memory as reconstruction; distributed cognition ---
\bibitem{bartlett1932}
F.~C. Bartlett.
\newblock \textit{Remembering: A Study in Experimental and Social Psychology}.
\newblock Cambridge University Press, Cambridge, 1932.

\bibitem{hutchins1995}
E.~Hutchins.
\newblock \textit{Cognition in the Wild}.
\newblock MIT Press, Cambridge, MA, 1995.

% --- Resilience, repair, and ecological persistence ---
\bibitem{holling1973}
C.~S. Holling.
\newblock Resilience and stability of ecological systems.
\newblock \textit{Annual Review of Ecology and Systematics}, 4:1--23, 1973.

% --- Network and graph structure ---
\bibitem{newman2010}
M.~E.~J. Newman.
\newblock \textit{Networks: An Introduction}.
\newblock Oxford University Press, Oxford, 2010.

\bibitem{barabasi2016}
A.-L. Barab\'asi.
\newblock \textit{Network Science}.
\newblock Cambridge University Press, Cambridge, 2016.

\end{thebibliography}

\end{document}
