\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{hebrew}
\usepackage{arabtex}
\usepackage{geometry}
\geometry{a4paper, margin=1in}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{tikz-cd}

\title{Chokepoint Capitalism in Knowledge Infrastructures: An RSVP-Theoretic Analysis}
\author{Flyxion}
\date{September 22, 2025}

\begin{document}

\maketitle
\begin{abstract}
This essay examines chokepoint capitalism as a mechanism for restricting knowledge diversity across digital, physical, and cultural infrastructures, using an RSVP-theoretic framework enriched with category and sheaf theory. Through cases spanning mobile operating systems, festival economics, visa policies, AI research platforms, and the historical evolution of alphabetic systems, I argue that chokepoints misprice epistemic diversity by enforcing premature evaluation, reducing negentropic potential. A functional paradigm of deferred automation, modeled as a monadic lazy-evaluation regime, is proposed as a counter-strategy, with the Arabic script’s morphological generators serving as a computational exemplar. Mathematical formalisms are offered not only to populate abstract registers but also to critique the cultural fetish for mathematical authority, with natural-language translations provided for accessibility.
\end{abstract}

\section{Introduction}
Chokepoint capitalism, as articulated by \citet{giblin2022chokepoint}, describes economic structures where gatekeepers control access to markets, extracting rents while suppressing alternatives. This essay extends the concept to knowledge infrastructures, arguing that chokepoints—whether in digital platforms, physical events, immigration policies, or historical script evolution—systematically undervalue viewpoint diversity, a critical driver of epistemic resilience. Using the RSVP framework, where scalar capacity ($\Phi$), vector flows ($v$), and entropy ($S$) model system dynamics, I demonstrate how such restrictions function as right-adjoint filters that collapse diversity colimits. A functional paradigm of deferred evaluation, inspired by lazy computation, is proposed to preserve negentropic potential, with the Arabic script’s morphology as a computational case study.

\section{Chokepoint Capitalism as Knowledge Restriction}

\subsection{Digital Platforms}
Mobile operating systems (Android, iOS) impose chokepoints by restricting system-level customization (e.g., fonts, terminals). For instance, Samsung’s theme economy licenses font design as a scarce privilege, akin to a lottery, while Android’s “Always use this app” prompt enforces premature defaults, creating hysteresis that discourages exploration of alternatives \citep{doctorow2020lockin}.

\subsection{Physical Analogues}
At festivals, organizers charge vendors exorbitant fees for exclusivity, policing independent sellers to maintain artificial scarcity. This mirrors digital chokepoints: access is monetized, and diversity (of vendors or ideas) is suppressed \citep{giblin2022chokepoint}.

\subsection{State-Level Chokepoints}
The U.S. H-1B visa fee of \$100,000, implemented on September 20, 2025, restricts labor mobility, mispricing the epistemic value of diverse perspectives from high-population countries like India and China \citep{cbc2025trumpvisa}. The policy, signed by President Donald Trump, applies only to new applicants and has sparked confusion, with the White House clarifying it does not affect current holders or renewals. Over 70 percent of H-1B holders are from India, and the policy has been denounced by India's Ministry of External Affairs for potential humanitarian disruptions to families. Critics argue it undercuts American workers by favoring lower-wage foreign hires, ignoring their role as negentropic amplifiers \citep{cbc2025trumpvisa}.

\subsection{AI Research Platforms}
AI companies invert traditional research by charging users to stress-test models, capturing trillions in negentropic value (e.g., \$15 trillion estimated for one firm) while externalizing costs. This is a chokepoint where access to knowledge production is tolled.

\section{Alphabetic Evolution as Historical Chokepoints}
The evolution of Phoenician into Hebrew, Arabic, Greek, and Latin scripts illustrates chokepoint dynamics in cultural infrastructures. Phoenician’s consonantal script was a high-entropy substrate, allowing lazy evaluation across cultures. Greek and Latin enforced eager vowelization, reducing ambiguity but creating tears in gluable diversity, while Hebrew and Arabic deferred vocalization, preserving combinatorial richness.

\section{Arabic Script as Computational Assembler}
\label{sec:arabicassembler}
The Arabic script exemplifies a computational generator system, analogous to an assembler in low-level programming. Consonants function as base instructions (generators), with vowels (fatḥa, kasra, ḍamma) and sukūn as arguments that instantiate syllables, and morphological measures (Forms I–X) as higher-order transformations that derive semantic fields from roots. This structure mirrors lazy evaluation: roots remain symbolic until context or diacritics force realization, maximizing interpretive potential. Unlike Greek’s early vowel commitment, Arabic’s deferred system supports a high-entropy lexical colimit, formalized as a monadic regime in Appendix \ref{app:arabic}.

\section{Counter-Strategy: Deferred Evaluation as Futarchy}
Deferring automation—whether in knowledge pipelines, app defaults, or script evolution—preserves epistemic diversity. Modeled as a monad in the RSVP category, this strategy delays forcing until colimits of local sections (drafts, app choices, cultural adaptations) are explored, maximizing negentropic value. My own pipeline, which generates directory trees (not just essays), embodies this principle, as does the Arabic script’s morphology.

\section{Chokepoint Field Theory for Vocabulary Choice}
In languages with modern standards like Arabic (MSA vs. dialects) or Spanish (RAE standards), chokepoints manifest as hierarchical filtering through businesses, media, and idiolects. Let $X$ be the context manifold and $\mathcal{L}_g \to X$ the concept bundle. The vocabulary field $\sigma_g: X \to \mathcal{L}_g$ minimizes the action
\[
\mathcal{S}[\sigma_g] = \int_X \big[ \alpha C_{\text{comp}}(x,\sigma_g(x)) + \beta C_{\text{prest}}(x,\sigma_g(x)) + \gamma \|\nabla_A \sigma_g(x)\|^2 + \lambda V_{\text{choke}}(x,\sigma_g(x)) \big] d\mu(x).
\]
Local choice follows softmax on energy $E(x,w)$, with coupling yielding phase boundaries. See Appendix \ref{app:vocabulary} for details.

\section{Conclusion}
Throughout this essay, we have argued that chokepoint capitalism and its linguistic analogues can be understood as field-theoretic phenomena. By treating vocabulary choice, gesture systems, and even ecological stigmergy as sections of a manifold constrained by gatekeeping potentials $V_{\text{choke}}$, we recover a unified picture: deferred diversity, entropy management, institutional curvature, and exogenous neural networks. The comparative functorial mappings—from deer tail-flagging to human waves to forest branchings—demonstrate that all communicative systems can be situated on a higher meta-language manifold. Functors translate between local codes wherever chokepoint curvature is compatible; where it fails, entropy tears and phase boundaries emerge.

Chokepoint field theory reframes capitalism, linguistics, gesture, and ecology under one principle: communication is infrastructure subject to entropy and curvature. The apparent diversity of media—words, bodies, paths—is a set of charts on a single manifold. The task is not to abolish chokepoints (impossible), but to understand, defer, and redistribute them in ways that preserve generativity and minimize destructive entropy.

Future work may formalize this framework within RSVP theory’s entropic action principles, embedding language choice and infrastructural design into the same thermodynamic ethics. In this way, questions of economy, speech, gesture, and ecology can be treated as instances of futarchy run by physical substrates: each making bets on which communicative pathways will carry meaning into the future.

\section*{Epilogue: Living With Chokepoints}
\addcontentsline{toc}{section}{Epilogue: Living With Chokepoints}

At first glance, chokepoints look like obstacles. They raise prices, limit access, and enforce defaults. But in language, gesture, and ecology, we have seen that chokepoints also shape the very forms of communication we rely on. They are not accidental; they are part of how systems generate meaning and stability.

A neighbor spotting you behind a gate, a deer flagging its tail, a path pressed into the forest floor—all of these are moments where visibility, scarcity, or constraint crystallize into a message. Each is a reminder that communication is not free-floating: it depends on substrates that carry it forward, and those substrates have their own bottlenecks and rules.

The task, then, is not to dream of a world without chokepoints, but to cultivate ways of living with them. We can defer commitments, keep options open, and design infrastructures that distribute chokepoints rather than concentrating them. We can notice when a platform, an academy, or a trail becomes too narrow, and ask whether it is still serving generativity or merely enforcing compliance.

Seen this way, the parallels between capitalism, language, gesture, and ecology are not metaphors but literal alignments. All are systems of meaning that pass through gates. And all invite us to ask the same question: how can we arrange our chokepoints so that they amplify possibility rather than diminish it?

\paragraph{Closing Thought.} 
Chokepoints do not only restrict; they also mark. 
Under certain conditions, what we call a chokepoint is better understood as a \emph{landmark}---a feature in the landscape of meaning that guides passage, anchors memory, and allows orientation across time and space.

\section*{Disclaimer on Formalism}
The mathematical translations in the appendices parody the cultural fetish for formal rigor while populating an abstract semantic register. Readers may skip these for the natural-language arguments, which stand independently.

\appendix
\section{Chokepoint Mispricing Theorem}
\label{app:mispricing}
Let RSVP be a symmetric monoidal category of states $x = (\Phi, v, S)$ and morphisms respecting conservation/relaxation. Fix a space $X$ with topology $J$. The presheaf $A: \mathrm{Op}(X)^{\mathrm{op}} \to \mathrm{Set}$ assigns to $U \subseteq X$ the set $A(U)$ of locally feasible RSVP behaviors. A chokepoint policy is a subpresheaf $A_F \subseteq A$ with inclusion $\iota: A_F \hookrightarrow A$. The diversity object is $\mathrm{Div}(A) = \mathrm{colim} \mathrm{Sec}(A)$. Let $V: \mathrm{Set} \to \mathbb{R}_{\ge 0}$ be a valuation valuation monotone in injections and colimit-superadditive, RSVP-sensitive.

\begin{lemma}
If there is a tear (local sections in $A_F$ glue in $A$ but not $A_F$), then $\kappa: \mathrm{Div}(A_F) \hookrightarrow \mathrm{Div}(A)$ is a proper mono.
\end{lemma}

\begin{proof}
A tear gives a cone in $\mathrm{Sec}(A_F)$ that maps to a gluable cone in $\mathrm{Sec}(A)$. The induced comparison is injective but not surjective.
\end{proof}

\begin{proposition}
If $\kappa$ is proper, then $V(\mathrm{Div}(A_F)) < V(\mathrm{Div}(A))$.
\end{proposition}

\begin{proof}
Proper mono into a set with non-null complement strictly increases valuation on codomain.
\end{proof}

\begin{theorem}[Chokepoint Mispricing]
Under tears and (V1)--(V3), $V(\mathrm{Div}(A_F)) < V(\mathrm{Div}(A))$.
\end{theorem}

\begin{proof}
Lemma 1 makes $\kappa$ proper; Proposition 2 yields strict loss.
\end{proof}

\section{Entropic Futarchy in Knowledge Pipelines}
\label{app:futarchy}
Let RSVP be the base category. An institution is a graph $R = (R, E)$ enriched by RSVP states $x_r$. The presheaf $H: \mathrm{Op}(X)^{\mathrm{op}} \to \mathrm{Set}$ assigns feasible hires. Policy $\pi$ yields subpresheaf $H_\pi \subseteq H$. Valuation $J(S) = \sum_r \alpha_r \mathbb{E}[\Delta N_r(S)] + \sum_e \beta_e \mathbb{E}[N(f_e|S)] - \lambda \mathrm{Risk}(S)$. Entropic Futarchy maximizes $V_\pi^{\mathrm{div}} = V(\mathrm{Div}(H_\pi))$.

\begin{theorem}[Colimit Advantage]
Under tears, $V_\pi^{\mathrm{div}} < V^{\mathrm{div}}$.
\end{theorem}

\begin{proof}
From Chokepoint Mispricing Theorem.
\end{proof}

Model. We model the null wave front as passive advection by a prescribed RSVP vector field with:

$\partial_t \psi_{\varnothing} + v\,\partial_x \psi_{\varnothing} = 0,\qquad
\psi_{\varnothing}(x,0) = \mathbf{1}_{[x_0,\,L)}(x)$,

Discretization. Use an upwind scheme (stable for CFL):

$\psi_i^{n+1} =
\begin{cases}
\psi_i^n - c\,\big(\psi_i^n - \psi_{i-1}^n\big), & v>0,\\[4pt]
\psi_i^n - c\,\big(\psi_{i+1}^n - \psi_i^n\big), & v<0,
\end{cases}
\quad
c \equiv \dfrac{v\,\Delta t}{\Delta x}$.

Parameter choice (example). $N=64$ cells, $L=1.0$, $v=0.8$, CFL$=0.8$. $\Delta x = L / N$, $\Delta t = \mathrm{CFL} \cdot \Delta x / v$, $T_\mathrm{final}=0.6$. $x = \mathrm{np.linspace}(0, L - \Delta x, N)$, $x_0=0.25$, $\psi = (x \geq x_0).\mathrm{astype(float)}$.

The simulation demonstrates the advancing null front, preserving monotonicity and aligning with the interpretation of $\psi_{\varnothing}$ as a constraint indicator.

\section{Deferred Automation as a Monad}
\label{app:deferral}
Let $T$ be a deferral monad on RSVP with $\eta, \mu$. Forcing $a: T \Rightarrow \mathrm{Id}$. For diagram $D$, lazy $X_T = a \circ T(\mathrm{colim} D)$, eager $X_E = \mathrm{colim}(a \circ T D)$.

\begin{theorem}[Lazy Dominance]
If $T$ preserves colimits and tears exist, $J(X_T) > J(X_E)$.
\end{theorem}

\begin{proof}
Compose inside $T$, glue via colimits, then force once. Earlier forcing collapses options prematurely.
\end{proof}

\begin{corollary}[Default-App Chokepoint]
Let $F$ be files, $E$ evaluators. Lazy $X_T = a \circ T(\mathrm{colim} \sum_e \mathrm{run}(e,f))$, eager $X_E = \mathrm{colim} \mathrm{run}(e^\star,f)$. If no evaluator dominates, $J(X_T) > J(X_E)$.
\end{corollary}

\begin{proof}
Non-dominance implies inequivalent amalgams lost in eager pipeline.
\end{proof}

\begin{corollary}[Alphabetic Chokepoint]
For Phoenician adaptations, if no script dominates, deferred (Phoenician open) yields $J(\mathrm{Div}(A)) > J(\mathrm{Div}(A_F))$.
\end{corollary}

\section{Arabic Morphology as a Deferred-Evaluation Monad}
\label{app:arabic}
Let Root be roots, Word forms. Monad $T$ with vowels $V = \{\fatha, \kasra, \damma\}$, measures $M_k$. Lex(R) = colim $\{M_k(\theta_v(R))\}$.

\begin{corollary}
Premature forcing yields $J(\mathrm{Lex}_{\mathrm{forced}}(R)) < J(\mathrm{Lex}(R))$.
\end{corollary}

\begin{proof}
Deferred preserves inequivalent amalgams.
\end{proof}

\section{Hebrew Abjad as a Lazy-Evaluation Regime}
\label{app:hebrew}
Let Root be roots, Form word-forms. Monad $T$ with niqqud as $\theta_v$. Binyanim $B: \mathrm{Root} \to \Form$. Lex(R) = colim $\{B(\theta_v(R))\}$.

\begin{corollary}
Omitting niqqud preserves $J(\mathrm{Lex}(R)) > J(\mathrm{Lex}_{\mathrm{forced}}(R))$.
\end{corollary}

\begin{proof}
Deferred vocalization maximizes colimit.
\end{proof}

\section{Comparative Table of Alphabetic Evaluation Regimes}
\label{app:table}
\renewcommand{\arraystretch}{1.4}
\begin{table}[h!]
\centering
\begin{tabular}{l p{2.5cm} p{2.5cm} p{2.5cm} p{2.5cm} p{2.5cm}}
\toprule
Dimension & Phoenician & Hebrew & Arabic & Greek & Latin \\
\midrule
RSVP--$\Phi$ (capacity) & 22 consonants as symbolic skeletons; high entropy potential & Consonantal skeletons; $\Phi$ remains unevaluated in writing & Consonantal skeletons + explicit vowel markers & Repurposed consonants as vowels; lower entropy & Latinized Etruscan script; fixed consonant+vowel inventory \\
RSVP--$v$ (flows) & Morphisms = adoption by different cultures; open mappings & Niqqud/maters added selectively; oral context as flow & Vowel diacritics + templatic measures direct flows & Vowel assignment forces evaluation flows early & Administrative flows standardized across empire \\
RSVP--$S$ (entropy) & Maximal diversity: many inequivalent mappings possible & High entropy preserved: roots deferred until contextual forcing & Structured entropy: root $\to$ pattern $\to$ lexical field & Entropy reduced: ambiguity suppressed in favor of clarity & Entropy minimized: enforced defaults create hysteresis \\
Lazy vs. Eager Evaluation & Fully lazy: consonants as thunks, no fixed vowels & Systemic laziness: vowels deferred, binyanim optional forcing & Semi-lazy: roots deferred, measures as controlled functors & Eager: vowels forced as defaults, lowering ambiguity & Fully eager: defaults enforced at systemic/imperial level \\
Monadic Operators & Root monad: $\eta$ lifts symbols, no $a$ (forcing) chosen & Niqqud/maters = optional $\theta_v$ evaluators; binyanim = higher-order functors & Diacritics = $\theta_v$ arguments; measures = functors $M_k$ & Vowel assignment = global forcing algebra $a$ & Default Latinization = right-adjoint filter $a$, destroying colimits \\
Chokepoint Effects & No chokepoint: diversity of alphabets emerges & Mild chokepoints only when niqqud becomes standardized & Regional chokepoints under Islamic expansion; script uniformity & Greek literacy stabilized by vowel forcing; limits diversity & Latin imposed as ``always this app'' across empire; alternatives marginalized \\
\bottomrule
\end{tabular}
\caption{Comparative evaluation regimes of Phoenician-derived alphabets under RSVP and lazy/eager functional paradigms.}
\end{table}

\section{Arabic Verb Forms (I–X) with Voice and Aspect}
\label{app:forms}
\renewcommand{\arraystretch}{1.3}
\begin{table}[h!]
\centering
\begin{tabular}{|c|c|c|c|c|}
\hline
Form & Template & Perfective Active & Imperfective Active & Perfective Passive \\
\hline
I   & فَعَلَ & faʿala & yafʿalu & fuʿila \\
II  & فَعَّلَ & faʿʿala & yufaʿʿilu & fuʿʿila \\
III & فَاعَلَ & fāʿala & yufāʿilu & fūʿila \\
IV  & أَفْعَلَ & ʾafʿala & yufʿilu & ʾufʿila \\
V   & تَفَعَّلَ & tafaʿʿala & yatafaʿʿalu & tufuʿʿila \\
VI  & تَفَاعَلَ & tafāʿala & yatafāʿalu & tufūʿila \\
VII & اِنْفَعَلَ & infaʿala & yanfaʿilu & unfuʿila \\
VIII& اِفْتَعَلَ & iftaʿala & yaftaʿilu & uftaʿila \\
IX  & اِفْعَلَّ & ifʿalla & yafʿallu & ufʿilla \\
X   & اِسْتَفْعَلَ & istafʿala & yastafʿilu & ustufʿila \\
\hline
\end{tabular}
\caption{Arabic verb forms (I–X) showing canonical perfective, imperfective, and passive patterns. These function as higher-order combinators in the Arabic Assembler calculus.}
\end{table}

\section{Arabic Nominal Derivations}
\label{app:nominals}
\renewcommand{\arraystretch}{1.3}
\begin{table}[h!]
\centering
\begin{tabular}{|c|c|c|p{5.5cm}|}
\hline
Category & Template & Example from Root ف–ع–ل & Semantics / Function \\
\hline
Verbal Noun (Maṣdar) & فَعْل / فِعَال / فُعُول & faʿl, fiʿāl, fuʿūl & Abstract action/event (“doing, deed, action”). Multiple canonical templates per form. \\
\hline
Active Participle & فَاعِل & fāʿil & “Doer” (agent noun); maps verb $\Verb$ to function $\iota \to o$. \\
\hline
Passive Participle & مَفْعُول & mafʿūl & “Object” (patient noun); maps verb $\Verb$ to result/state. \\
\hline
Instrumental Noun & مِفْعَال / مِفْعَلَة & mifʿāl, mifʿala & Tool/device noun (“instrument of doing”). \\
\hline
Place/Time Noun & مَفْعِل / مَفْعَل & mafʿil, mafʿal & Noun of locus or occasion (“place/time of doing”). \\
\hline
Abstract Noun & فُعُولَة / فَعَالَة & fuʿūla, faʿāla & Nominalization of qualities (“quality/state of doing”). \\
\hline
\end{tabular}
\caption{Nominal derivations from triconsonantal roots, exemplified with ف–ع–ل. Each acts as a templatic combinator producing $\Noun$ or $\Partic$ types in the Arabic Assembler calculus.}
\end{table}

\section{Typing Hierarchy for Arabic Assembler}
\label{app:hierarchy}
\begin{center}
\begin{tikzcd}[row sep=large, column sep=huge]
 & \Verb \arrow[dl, swap, "P_{\text{masdar}} \; (\Phi\mapsto S)"] 
       \arrow[d, "P_{\text{actPart}} \; (\Phi\mapsto v)"] 
       \arrow[dr, "P_{\text{passPart}} \; (\Phi\mapsto v^{-1})"] 
       \arrow[drr, bend left=12, "P_{\text{instr}} \; (v\mapsto \Phi)"] 
       \arrow[drrr, bend left=18, "P_{\text{loc}} \; (S\mapsto \Phi)"] & & & & \\[0.5em]
\Noun_{\text{event}} \; (S) 
& \Partic_{\text{act}} \; (v) \arrow[d, dashed, "agent \mapsto \Noun \; (\Phi\otimes v)"] 
& \Partic_{\text{pass}} \; (v^{-1}) \arrow[d, dashed, "patient \mapsto \Noun \; (S\otimes v^{-1})"] 
& \Noun_{\text{instr}} \; (\Phi) 
& \Noun_{\text{locus}} \; (\Phi\otimes S) \\[0.5em]
 & \Noun_{\text{agent}} \; (\Phi\otimes v) 
 & \Noun_{\text{patient}} \; (S\otimes v^{-1}) 
 & & 
\end{tikzcd}
\end{center}

\paragraph{Legend.}
\begin{itemize}
  \item $\Phi$ = latent postural potential.
  \item $v$ = directed flow of agency.
  \item $S$ = entropic resolution.
  \item Verb = live transformation.
  \item Maṣdar = $\Phi \mapsto S$.
  \item Active participle = $\Phi \mapsto v$.
  \item Passive participle = $\Phi \mapsto v^{-1}$.
  \item Instrument noun = $v \mapsto \Phi$.
  \item Place/time noun = $S \mapsto \Phi$.
\end{itemize}

\section{Arabic Derivation as RSVP Sheaf Theory}
\label{app:sheaf}
Let Root, Templ, Word be categories of roots, templates, words. Measure functors $M_k: \mathrm{Root} \to \mathrm{Templ}$, vowel functors $V_\delta: \mathrm{Templ} \to \mathrm{Word}$. Composite $F_{k,\vec{\delta}} = V_{\delta_n} \circ \cdots \circ M_k$. Presheaf $\mathcal{D}: \mathrm{Op}(X)^{\mathrm{op}} \to \mathrm{Cat}$ assigns local derivations. Natural transformations $\alpha_{k\to k'}: M_k \Rightarrow M_{k'}$, $\beta_{\delta\to\delta'}: V_\delta \Rightarrow V_{\delta'}$.

\begin{tikzcd}[column sep=huge,row sep=large]
\mathbf{Root}
  \arrow[r, bend left=12, "M_k"]
  \arrow[r, bend right=12, swap, "M_{k'}"]
& \mathbf{Templ}
  \arrow[r, bend left=12, "V_\delta"]
  \arrow[r, bend right=12, swap, "V_{\delta'}"]
  \arrow[loop, distance=2em, out=80, in=100, "\ T\ "]
& \mathbf{Word} \arrow[r, "\llbracket-\rrbracket"] & \boldsymbol\Lambda \\
%
& \arrow[uu,Leftarrow,shorten <=0.6em,shorten >=0.6em,"\alpha_{k\to k'}" description, yshift=1ex] {}
& \arrow[uu,Leftarrow,shorten <=0.6em,shorten >=0.6em,"\beta_{\delta\to\delta'}" description, yshift=1ex] {}
&
\end{tikzcd}

\begin{theorem*}[Arabic as Sheafed RSVP]
Arabic derivation is a presheaf of functorial pipelines whose sheaf gluing realizes lexical diversity. Enforcement reduces RSVP valuation unless dominant.
\end{theorem*}

\section{Chokepoint Field Theory for Arabic Vocabulary Choice}
\label{app:vocabulary}
Let $X$ be context manifold, $\mathcal{L}_g \to X$ concept bundle. Field $\sigma_g: X \to \mathcal{L}_g$ minimizes
\[
\mathcal{S}[\sigma_g] = \int_X \big[ \alpha C_{\text{comp}}(x,\sigma_g(x)) + \beta C_{\text{prest}}(x,\sigma_g(x)) + \gamma \|\nabla_A \sigma_g(x)\|^2 + \lambda V_{\text{choke}}(x,\sigma_g(x)) \big] d\mu(x).
\]
Local choice softmax on $E(x,w)$. Coupling yields fronts.

\subsection{Methods for Estimation}
Data: Contexts $X$, concepts $G$, variants per $g$. Decompose $E(x,w)$ with proxies. Fit $\alpha,\beta,\gamma,\lambda,T$ via MLE/MAP. Ablations quantify diversity loss.

\subsection{Hierarchical Filtering}
$V_{\text{choke}} = \theta_{\text{biz}} \phi_{\text{biz}} + \theta_{\text{media}} \phi_{\text{media}} + \theta_{\text{edu}} \phi_{\text{edu}}$. Fit $\theta$ by domain.

\subsection{Meta-Language Manifold}
Systems as charts on a manifold. Functors translate; curvature obstructs.

\begin{theorem*}[Universality]
Pairwise functors exist where chokepoints compatible, yielding a sheaf of manifolds.
\end{theorem*}

\subsection{Worked Example: Functorial Mapping}
$F(\text{tail-flag}) = \text{wave}$, $G(\text{wave}) = \text{branch-split}$. Composition shows cross-modality coherence.

\section{Chokepoint Field Theory Beyond Spoken Language}
\label{app:gestural}
\subsection{Neighbor Spotting as Greeting}
Gestures: \{duck, sit, look-aside, hide\}. Semantics: visibility acknowledgment. Chokepoint: line-of-sight constraints.

\subsection{Generative Gestural Languages in Animals}
Deer/primate: tail-flags, head-bobs. $\Phi$: repertoire, $v$: signals, $S$: ambiguity. Generativity: combinations.

\subsection{Stigmergic Path Clearing}
Paths as axons, animals as neurons. $\Phi$: routes, $v$: traffic, $S$: attractors.

\subsection{Comparative Table}
\renewcommand{\arraystretch}{1.3}
\begin{table}[h!]
\centering
\begin{tabular}{|l|p{4cm}|p{4cm}|p{4cm}|}
\hline
Dimension & Human Gestures & Animal Gestures & Stigmergic Forest \\
\hline
RSVP--$\Phi$ & Body repertoire & Gestural inventory & Potential routes \\
RSVP--$v$ & Recognition flows & Directional signals & Movement flows \\
RSVP--$S$ & Ambiguity of intent & Gesture meanings & Path convergence \\
Generativity & Posture sequencing & Combinatorial rules & Trace accumulation \\
Chokepoints & Visibility constraints & Hierarchy/perception & Terrain bottlenecks \\
Exogenous Substrate & Built environment & Social hierarchy & Forest floor \\
\hline
\end{tabular}
\caption{Gestural and exogenous systems as RSVP fields.}
\end{table}

\subsection{Theorem: Universality of Chokepoints}
Nonzero $V_{\text{choke}}$ + $\gamma > 0$ yields lock-in and fronts. Defer reduces action unless dominance.

\newpage
\bibliographystyle{plainnat}
\bibliography{references}
\end{document}
